### 2022-12-31 MSIS_model_explain/MSIS_model_explain.Rmd
preliminary tutorial material

---
title: "Age-structured transmission dynamic models of RSV"
subtitle: "Tutorial 1 for transition"
author: 
- Gigi (Zhe Zheng)^[zhe.zheng@yale.edu; zhe.zheng@aya.yale.edu; gigi.zhe.zheng@gmail.com]
- Dan (Daniel Weinberger) 
- Ginny (Virginia Pitzer)
date: "2022-12-22"
fontsize: 10pt
output: 
  beamer_presentation:
  theme: CambridgeUS
  colortheme: "beaver"
  fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

## References

This age structured transmission dynamic model of RSV was first developed by Ginny Pitzer in her publication: https://doi.org/10.1371/journal.ppat.1004591. With the help of Dan Weinberger, we translated Ginny's Matlab codes to R codes. We further developed/modified this model, resulting in several mechanistic models corresponding to a variety of scenarios. Please refer to:

- https://doi.org/10.1001/jamanetworkopen.2021.41779
- https://doi.org/10.1038/s41541-022-00550-5
- https://doi.org/10.1101/2022.11.10.22282132

The complete code example for this tutorial please refer to
- https://github.com/weinbergerlab/RSV_metapop/tree/master/SimpleModel

## The Compartmental Model Structure (states of transmission dynamics)

::: columns

:::: {.column width="25%"}

![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)

::::

:::: {.column width="75%"}

- Each white compartment correspond to an underlying status in transmission dynamics. These statuses are **unobservable**. 
    - **M** All infants born are born with protective maternal immunity, which wanes exponentially. The waning rate is $\omega$.
    - $\boldsymbol{S_0}$ After protective maternal immunity wanes, infants are susceptible to RSV infection. Being exposed to infection, susceptible infants progress to infected status  at rate $\lambda$.
    - $\boldsymbol{I_0}$ We ignored the exposed status (short and do not affect transmission dynamics) and assumed infected individuals are infectious. The first-time infected individuals recovered at rate $\gamma_1$
    - $\boldsymbol{S_n}$ Here $n$ is the number of previous infections. Previous infection reduced the risk of re-infection. $\sigma_1 \sim \sigma_3$
    - $\boldsymbol{I_n}$ Here $n$ is the number of infections. Infectious periods are shorter in subsequent infections. $\gamma_2 \sim \gamma_3$
::::

:::

## The Compartmental Model Structure (rates)

::: columns

:::: {.column width="25%"}

![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)
\

\small
\textcolor{red}{Parameters in red are estimated for each state} 
\
\
\textcolor{blue}{Parameters in blue are from published literature (either as fixed value inputs or ranges as priors)}

::::

:::: {.column width="75%"}

- Rates in this compartmental model.  
    - $\textcolor{red}\omega$ is the waning rate of protective maternal immunity
    $$\textcolor{red}\omega=\frac{1}{\textcolor{blue}{\text{duration of maternal immunity}}}$$
    -  $\lambda$ is the force of infection
    $$\lambda= \textcolor{red}{\beta_0}(1+\textcolor{red}{A} \text{cos}(2\pi\nu t-\textcolor{red}{\phi})) I$$ 
    $\textcolor{red}{\phi}$ has a minus sign for moving towards a later time period.
    $$\nu= \frac{1}{\textcolor{blue}{\text{period}}}$$
     $$I=I_1+ \textcolor{blue}{\rho_1}I_2+\textcolor{blue}{\rho_2}(I_3+I_4)$$
    here $\textcolor{blue}{\rho_1 \sim \rho_2}$ are the relative infectiousness for 2nd and subsequent infections (probabilities)
    - $\textcolor{blue}{\gamma_1 \sim \gamma_3}$ are the rates of recovery
    $$\textcolor{blue}{\gamma_1 \sim \gamma_3} = \frac{1}{\textcolor{blue}{\text{duration of infectiousness}}}$$
::::

:::

## The Compartmental Model Structure (probabilities)

::: columns

:::: {.column width="25%"}

![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)

::::

:::: {.column width="75%"}

- Probabilities in this compartmental model.  
    - $\textcolor{blue}{\sigma_1 \sim \sigma_3}$ are the  relative risks of infection following 1st, 2nd, 3rd+ infections
    \
    - $\textcolor{blue}{\rho_1 \sim \rho_2}$ are the relative infectiousness for 2nd and subsequent infections
    \
    - $\textcolor{blue}{d_1(a) \sim d_2(a)}$ are the age- and infection-specific risks of developing lower respiratory tract illness given infections (LRTIs). In Ginny's model, only the first two infections will result in LRTIs. Later, we modified the model to assume subsequent infections will also lead to LRTIs $\textcolor{blue}{d_3(a) \sim d_4(a)}$. This is very important for model calibration in older adult populations (For example, this will be needed in the cost-effectiveness of older adult RSV vaccine project).
    \
    - $\textcolor{blue}{h_1(a) \sim h_4(a)}$ are the probability of requiring inpatient care given LRTIs. These probabilities are also age- and infection-specific.
\
- \textcolor{blue}{From published literature (Either as fixed value inputs or ranges as priors)}

::::

:::

## The Compartmental Model Structure (observed states for calibration)

::: columns

:::: {.column width="25%"}

![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)
::::

:::: {.column width="75%"}

- Each white compartment correspond to a disease status. These statuses are **observable**. These outputs of transmission dynamic models will be used for the model calibration.
\
\
    - **D** Individuals in each age group who develop lower respiratory tract illness. We can find this information in state emergency department visits and state inpatient visits.
    \
    \
    - **H** Individuals in each age group who are admitted to hospitals because of RSV infection. We can find this information in state inpatient visits. For my previous work, I calibrated the transmission model outputs to time-series generated from State Inpatient Databases. 
    \
    \
    -  Click to see information on [\textcolor{blue}{State Emergency Department Databases}](https://www.hcup-us.ahrq.gov/db/state/sedddbdocumentation.jsp) and [\textcolor{blue}{State Inpatient Databases}](https://www.hcup-us.ahrq.gov/db/state/siddbdocumentation.jsp).
::::

:::

## The Compartmental Model Structure (observed statuses for calibration)

::: columns

:::: {.column width="25%"}

![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)

::::

:::: {.column width="75%"}

- Each white compartment correspond to a disease status. These statuses are **observable**. These outputs of transmission dynamic models will be used for the model calibration. 
    \
    \
    - Note that although the model outputs specify the number of previous infection, we are unlikely to observe this information in real world.
    \
    \
    - Note that [\textcolor{blue}{State Emergency Department Databases}](https://www.hcup-us.ahrq.gov/db/state/sedddbdocumentation.jsp) and [\textcolor{blue}{State Inpatient Databases}](https://www.hcup-us.ahrq.gov/db/state/siddbdocumentation.jsp) only captured the RSV visits that are tested and recorded. This may not be a problem in pediatric populations but it needs to be corrected for the testing and recording ratios in adult populations.

::::

:::

## R Code in Details (Basic Age-structure MSIS model)

### Setting up the packages

```{r, echo=T}
## load required packages
library(deSolve) 
# for solving the ordinary differential equations
library(RColorBrewer) 
# color palettes for making plots
library(reshape2) 
# for reshape data frames
```

## Setting up inputs

```{r, echo=T}
# T is time points. T contains both the burn-in period 
# and evaluation period.
# N_ages is number of age groups.

Pop1 <- readRDS('./SimpleModel/data_and_parms/pop1.rds') 
#initial population, by age group.

B <-  readRDS('./SimpleModel/data_and_parms/Birth_rate.rds')
# birth rate in each age group
# a matrix with T rows and  N_ages columns
# The first column is the population birth rate 
# They are born into the 0 month age group
# The rest columns are all zeros (1 months to 80+ age group)  
# For U.S., you will find this information in CDC wonder 
```

Click to see information on [\textcolor{blue}{CDC wonder}](https://wonder.cdc.gov/natality-expanded-current.html)

## Setting up inputs

```{r, echo=T}
c2 <- readRDS( './SimpleModel/data_and_parms/c2.rds')
# The contact patterns in each age group.
# The patterns affect the age-specific likelihood 
# of a susceptible individual come into contact with 
# an infectious individual.
# The exact values does not matter because we will
# estimate beta_0.
# This information can be found in published literature
# Then you may consider rearrange the matrix to
# reflect the age groups in your model
# I wrote an r script to rearrange the contact matrix
# see contactmatrix.r under code source folder
```

## Age structure in this example

In Ginny's paper, the age groups are as follows: The <12 month olds, were divided into monthly age classes. The remaining population was divided into 6 classes: 1–4 years old, 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old. In this example, the 1-4 year old are divided into 12 month age classes as well. 
\
\
How to divide age depends on the goal of your project. It relates closely to the parameters for model calibration and the difficulty of model calibration.
\
\
We set names of age groups 
```{r, echo=T}
N_ages <- length(Pop1) 
agenames <- paste0('Agegrp', 1:N_ages) 
#Could replace this with vector of actual age names
```

## Initialize the compartments (states) 
Please refer to page 3: [The Compartmental Model Structure (states of transmission dynamics)]
```{r, echo=TRUE}
StateNames <- c('M','S0','I1','S1','I2','S2','I3','S3','I4')

# N age groups x K states
yinit.matrix <- array(NA, dim=c(N_ages, length(StateNames)))

# assign row names and column names
dimnames(yinit.matrix)[[1]] <- agenames
dimnames(yinit.matrix)[[2]] <- StateNames

# Initializes population with infants under 3 months
# are protected by maternal immunity and
# with 1 infected person per age group in other age groups
yinit.matrix[,c('S1','I2','S2','I3','S3','I4')]  = 0
yinit.matrix[,'M'] = c(Pop1[1:3], rep(0,N_ages-3))
yinit.matrix[,'S0'] = c(rep(0,3),Pop1[4:N_ages]-rep(N_ages-3)) 
yinit.matrix[,'I1'] = c(rep(0,3), rep(1,N_ages-3))  
```

## Vectorize the states for ODE input

```{r, echo=TRUE}
#Vectorize the ynit matrix
yinit.vector <- as.vector(yinit.matrix)

# Create array that has the labels by age, state 
# and use this to name the yinit.vector
name.array <- array(NA, dim=dim(yinit.matrix))
for(i in 1:dim(name.array)[1]){
 for(j in 1:dim(name.array)[2]){
   name.array[i,j] <- paste(dimnames(yinit.matrix)[[1]][i],
                            dimnames(yinit.matrix)[[2]][j])
    }
  }

name.vector <- as.vector(name.array)
names(yinit.vector) <- name.vector
```

## Contact matrix in our case

```{r, echo=T, fig.dim=c(2.8, 2.8)}
n.cols=100
nice.cols <-  colorRampPalette(brewer.pal(9,
                                  "YlOrRd"))(n.cols)
heatmap(c2/sum(diag(c2)), 
        Rowv=NA, Colv=NA, scale='none', col=nice.cols)
```

## Section outlines
\
- In the first section, we will learn how to run transmission model with all parameters known. 
\
\
- In the second section, we will learn to use maximum likelihood estimation (MLE) to estimate the unknown parameters (one value for each unknown parameter). 
\
\
- In the third section, we will learn to use STAN, a probabilistic programming language for statistical inference^[https://en.wikipedia.org/wiki/Stan_(software)], to get the 95% credible intervals from the posterior samples of these unknown parameters.

## Section 1: run transmission model with all parameters known
\
We first assign values to all required parameters, including rates and probabilities. Most of these are described in Table 2 of Pitzer et al, PLOS Pathogens^[https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1004591]
\
\
```{r, echo=T}
###########################################
# omega = 1/duration of maternal immunity 
DurationMatImmunityDays= 112 #days
###########################################

##############################################
# gamma1~gamma3 = 1/duration of infectiousness 
# Duration in days
dur.days1 <- 10 #days
dur.days2 <- 7 #days
dur.days3 <- 5 #days
############################################## 
```

## Parameters related to the force of transmission
$$\lambda= \textcolor{red}{\beta_0}(1+\textcolor{red}{A} \text{cos}(2\pi\nu t-\textcolor{red}{\phi})) I$$

$\beta_0$ is the effective contact rate, meaning the number of infections per unit time per susceptible per infected.\
\
$\beta_0$ = per capita transmission probability * total contact rate
\
```{r, echo=T}
#These parameters will likely be estimated in practice
##################################################### 
#Seasonal components--
Amp = 0.2 #Seasonal amplitude
phi = 3.327749 #Seasonal phase shift 
# (depends on the starting month;here 0=peak @ July 1)

##################################################### 
# per capita transmission probability
baseline.txn.rate <- 9.2 
# this is now for the entire infectious period
# it needs to divide the length of infectious period
##################################################### 
```

## Parameters related to the force of transmission (continued)
$\beta_0$ = total contact rate * per capita transmission probability

Note, In Density Dependent transmission, the contact rate (c) depends on the population density.  In Frequent Dependent transmission, the contact rate (c’) does not depend on the population density. (In this example, we assumed frequency-dependent)
\
\
```{r, echo=T}
############################################### 
c=c2 # contact rate

q=1 # use this to switch between density (q=0) vs 
# frequency-dependent (q=1) transmission
###############################################
```
\
To better understand the different in Frequent Dependent transmission and Density Dependent transmission, please check out: https://parasiteecology.wordpress.com/2013/10/17/density-dependent-vs-frequency-dependent-disease-transmission/

## Probabilities related to the number of infections
Please refer to page 5: [The Compartmental Model Structure (probabilities)]
```{r, echo=T}
##########################################################
#Relative infectiousness for 2nd and subsequent infections
rho1 = 0.75
rho2 = 0.51

##########################################################
#Relative risk of infection following 1st, 2nd, 3rd+ infections
sigma1=0.76
sigma2=0.6
sigma3=0.4
##########################################################
```

## Age-specific probabilities related to the number of infections

In this example, the LRI and hospitalization probabilities are from several cohort studies. These studies reported the probabilities in aggregated age groups (every 3 month). In later section, if you would like to fit the model outputs to self-defined age groups, you will need to come up with the age-specific probabilities first. This can be done by creating a polynomial regression fit to the reported probabilities of the cohort studies.

```{r, echo=T}
###########################################
# LRI probability
###########################################

#proportion of first infections that are LRI (by age)
delta1=c(rep(.40,3),rep(.39,3),rep(.21,3),
         rep(.20,3),0.16,rep(.14,3),rep(0.05,N_ages-16))
#proportion of second infections that are LRI 
delta2=.5*delta1
#proportion of third infections that are LRI
delta3=.7*delta2
```

## Age-specific probabilities related to the number of infections (continued)

```{r, echo=T}
###########################################
# Hospitalization probability
###########################################

#proportion of first infection that are hospitalized
hosp1=c(.18*rep(.40,3),0.08*rep(.39,3),
        0.07*rep(.21,3),0.06*rep(.20,3),0.06*0.16,
        0.05*rep(.14,3),0.02*rep(0.05,N_ages-16))
# = hosp prob given LRI * LRI prob given infection

#proportion of second infection that are hospitalized
hosp2=.4*hosp1

#proportion of subsequent infection that are hospitalized 
#(The last two probabilities come from the previous 
#fitting of the transmission dynamic model)
hosp3=c(rep(0,N_ages-2),0.00001,0.00004)
```

## Parameters governing population dynamics

```{r, echo=T}
############################################### 
# birth rate  
# Matrix: T rows, N_ages columns; columns 2:N_ages all 0s
PerCapitaBirthsYear=B 
############################################### 
# net rate of crude deaths (+) and immigration (-) 
# You should calibrate this parameter 
# so we can reproduce the population growth
um= -0.0002227 #(from all age groups)

############################################### 
# Aging rate = 1/width age class (months) 
# Vector of long N_age
WidthAgeClassMonth = c(rep(1,times=12), 
      rep(12,times=4),  60, 120, 240, 240, 240)  
###############################################
```

## Save parameters in a list

```{r saveparms, echo=T}
parms<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
            DurationMatImmunityDays=DurationMatImmunityDays,
            WidthAgeClassMonth=WidthAgeClassMonth,
            um=um, # net growth rate
            Amp=Amp, # seasonal amplitude
            phi=phi, # seasonal peak timing
            rho1=rho1, # Relative infectiousness (2nd)
            rho2=rho2, # Relative infectiousness (3rd+)
            dur.days1=dur.days1,# Duration of infectiousness
            dur.days2=dur.days2,# Duration of infectiousness
            dur.days3=dur.days3,# Duration of infectiousness
            yinit.matrix=yinit.matrix, # initial states
            baseline.txn.rate = baseline.txn.rate,
            q=q, # Frequency or Density dependent
            contact=c2, # contact matrix
            sigma1=sigma1, # Relative risk of infection (2nd)
            sigma2=sigma2, # Relative risk of infection (3rd)
            sigma3=sigma3, # Relative risk of infection (4th+)
            time.step='month')
```

## Read in the model

```{r, echo=T}
#Read in the model
source('./SimpleModel/data_and_parms/simple_model.R') 
```

In the following slides, we will show the source function of ordinary differential equations. You can modify this function to reflect different assumptions.
\
\

Note, please modify the r script **simple_model.R** in the folder **data_and_parms** under **SimpleModel**. In this PDF, the function of transmission dynamic model is separated into several chunks for the tutorial purpose.

## Source function of the transmission dynamic model
```{r,echo=T,eval=FALSE}
simple_model <- function(t,y,parms,time.step='month'){

  # read in initial states and their names
  States<-array(y, dim=dim(parms$yinit.matrix))
  dimnames(States) <- dimnames(parms$yinit.matrix)

  # unify the time unit of parameter inputs
  if(parms$time.step=='month'){
    period=12
    length.step=30.44 #days
  }else if(parms$time.step=='week'){
    period=52.1775
    length.step=7 #days
  }
```

## Source function of the transmission dynamic model (continued)

Note: Here we need to convert all the rates from 1/days to 1/length.step
\
```{r,echo=T,eval=FALSE,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
  # waning rate of maternal immunity (by time step)
  omega = 1/(parms$DurationMatImmunityDays/length.step)

  # aging rate (by time step)
  mu= 1/parms$WidthAgeClassMonth
  if(parms$time.step=='week'){
    mu= 1/(WidthAgeClassMonth*4.345)}

  # rate of recovery of first infection
  gamma1= 1/(parms$dur.days1/length.step) 
  # rate of recovery of second infection
  gamma2= 1/(parms$dur.days2/length.step)
  # rate of recovery of third infection
  gamma3= 1/(parms$dur.days3/length.step)  
  gamma4= gamma3  
  #gamma3 stands for rate of recovery from subsequent infection
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
  # Relative risk of infection (2nd)
  sigma1=parms$sigma1 
  # Relative risk of infection (3rd)
  sigma2=parms$sigma2
  # Relative risk of infection (4th+)
  sigma3=parms$sigma3

  # Relative infectiousness (2nd)
  rho1=parms$rho1 
  # Relative infectiousness (3rd+)
  rho2=parms$rho2
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE,tidy=TRUE, tidy.opts=list(width.cutoff=60)}
  #Pull out the states for the model as vectors
  M <-  States[,'M'] # protected by maternal immunity
  S0 <-  States[,'S0'] # purely susceptible population
  I1 <-  States[,'I1'] # first time infection (infectious)

  S1 <-  States[,'S1'] 
  # susceptible population with build-up immunity
  I2 <-  States[,'I2'] # second time infection

  S2 <-  States[,'S2']
  # susceptible population with lower risk of re-infection
  I3 <-  States[,'I3'] # third time infection

  S3 <-  States[,'S3']
  # susceptible population with lowest risk of re-infection
  I4 <-  States[,'I4'] # subsequent time infection

  N_ages <- length(M) # the number of age groups
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE, tidy.opts=list(width.cutoff=60)}
## parameter related to force of infection ################
# per capita transmission probability
baseline.txn.rate=parms$baseline.txn.rate
# transmission probability per unit time
b <- baseline.txn.rate/ (parms$dur.days1/length.step) 
q=parms$q # q depends on transmission type 
# (whether depends on population density or not)
contact=parms$contact # c2 is the contact matrix
# transmission probability per unit time in each age group
beta <- (b/100)/(sum(yinit.matrix)^(1-q))*contact 
# 100 is a scaling factor for the contact matrix we choose
# (see Ginny's paper and Matlab code for details)
# this does not matter because most likely
# you will need to estimate baseline.txn.rate
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE}
## parameter related to force of infection ################
Amp=parms$Amp # seasonal amplitude
phi=parms$phi # seasonal phase shift
#seasonality
seasonal.txn <- (1+Amp*cos(2*pi*(t-phi*period)/period))

# seasonal transmission probability
beta_a_i <- seasonal.txn * beta 
infectiousN <- (I1+rho1*I2+rho2*I3+rho2*I4)/sum(States)
# for frequency dependent transmission

lambda <- infectiousN %*% beta_a_i # force of transmission
lambda <- as.vector(lambda) # vectorize force of transmission
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE}
# create a matrix to record the changing variables
dy <- matrix(NA, nrow=N_ages, ncol=ncol(States))
colnames(dy) <- colnames(States)

period.birth.rate <- 
  log(parms$PerCapitaBirthsYear[t,]+1)/period
# get period birth rate from annual birth rate
# see the following page for birth rate calculation

#um is death rate
um=parms$um
#mu represents aging to the next class
Aging.Prop <- c(0,mu[1:(N.ages-1)])
```

## Period birth rate calculation

For birth rate^[https://journals.plos.org/plospathogens/article/file?id=10.1371/journal.ppat.1004591.s016&type=supplementary]:
\
\
The weekly per capita birth rate $B_w$ is equal to log(1+B)/52.18 because the data on the birth rate is annual and the differential equation model inherently assumes that growth is occurring exponentially. So if the annual birth rate is equal to 12 per 1000 per year (B=0.012), for example, then we are assuming that:
$$N_1=N_0e^{(B_w*52.10)}$$ where $N_1$ is the population after 1 year and $N_0$ is the baseline population, and $N_1=N_0*(1+B)$.
$$1.012=1*e^{(B_w*52.18)}$$
$$B_w=log(1.012)/52.18$$.
The same goes for getting monthly birth rate.

## Source function of the transmission dynamic model (continued)

For the equations, please refer to [the supplementary document of Ginny's paper](https://journals.plos.org/plospathogens/article/file?id=10.1371/journal.ppat.1004591.s016&type=supplementary)
```{r,echo=T,eval=FALSE}
# ordinary differential equations
  dy[,'M'] <- period.birth.rate*sum(States) - 
    (omega+(mu+um))*M +
    Aging.Prop*c(0,M[1:(N_ages-1)])

  dy[,'S0'] <- omega*M -
    lambda*S0 -
    (mu + um)*S0 + 
    Aging.Prop*c(0,S0[1:(N_ages-1)])

  dy[,'I1'] <-   lambda*S0 - 
    (gamma1 + mu + um)*I1 + 
    Aging.Prop*c(0,I1[1:(N_ages-1)]) 
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE}
  dy[,'S1'] <- gamma1*I1 - 
    sigma1*lambda*S1 - 
    (mu+um)*S1 + 
    Aging.Prop*c(0,S1[1:(N_ages-1)])

  dy[,'I2'] <- sigma1*lambda*S1 - 
    gamma2*I2-(mu + um)*I2 + 
    Aging.Prop*c(0,I2[1:(N_ages-1)])

  dy[,'S2'] <- gamma2*I2 - 
    sigma2*lambda*S2 -
    (mu+um)*S2 + 
    Aging.Prop*c(0,S2[1:(N_ages-1)])

  dy[,'I3'] <- sigma2*lambda*S2 -
    (gamma3 + mu+um)*I3 +  
    Aging.Prop*c(0,I3[1:(N_ages-1)]) 
```

## Source function of the transmission dynamic model (continued)
```{r,echo=T,eval=FALSE}
  dy[,'S3'] <- gamma3*I3 +  
    gamma4*I4 -
    sigma3*lambda*S3 -
    (mu + um)*S3 + 
    Aging.Prop*c(0,S3[1:(N_ages-1)])

  dy[,'I4'] <- sigma3*lambda*S3 - 
    gamma4*I4 - 
    (mu + um)*I4 + 
    Aging.Prop*c(0,I4[1:(N_ages-1)])

  derivs <- as.vector(dy)

  res <- list(derivs)

  return(res)
}
```

## Run the model

**NOTE: time step here is in months--you need to adjust seasonality accordingly for the time step you choose**
```{r,echo=T}
 start_time = 1 # start date (years)
 tmax = nrow(B)
# end_time = 25 # end date (years)
my_times <- seq(start_time, tmax, by = 1) 
# gives a sequence from start to end in increments of 1
```
\
Using the ODE function, we get the results of the transmission dynamic model^[NOTE: here we are just simulating with set parameters, not fitting to data].
\
```{r,echo=T}
results <- ode(y=yinit.vector, t=my_times,  
               func=simple_model, 
            parms=parms)
# y is the initial population in each state and age
# t is the time steps of evaluation
# func specify the ordinary differential equations
# parms are all the parameter inputs
```

## Extract population growth in each age group and the entire population

```{r,echo=T}
#Ginny Pitzer used a 40-50 YEAR burn in period
burnN <- 25*12 # you will need to evaluate 
# whether after burn-in the model reach quasi-equilibrium
results.burned <- results[-c(1:burnN),]

# first get the total population in each time point
pop.all <- rowSums(results.burned[,-1])

# Then check the population growth in each age group
# from wide format to long format
all.m <- melt(results.burned[,
              grep('Agegrp',colnames(results.burned))])
# then we get the name of age group of each state
all.m$agegrp <- sub(" .*", "",all.m$Var2) 
# sum within age groups to get the age-specific
# population in each time point
all.c <- dcast(all.m, Var1~agegrp, fun.aggregate = sum)
```

## Exact the prevalence of infection from the results
```{r,echo=T}
##Any infected person
infected.cols <- results.burned[,
                c(grep('I1', colnames(results.burned)),
                 grep('I2', colnames(results.burned)),
                 grep('I3', colnames(results.burned)),
                 grep('I4', colnames(results.burned)))]

## Sum up the infected individuals at each time point
infected.all <- apply(infected.cols,1,sum)# = rowsum

## this chunk of code get the prevalence of infection
## in each age group at each time point.
infected.cols.m <- melt(infected.cols)
infected.cols.m$agegrp <- sub(" .*", "",infected.cols.m$Var2 )
infected.cols.c <- dcast(infected.cols.m,
                         Var1~agegrp,
                         fun.aggregate = sum)
```

## Visualize population growth and prevalence of infection

```{r,echo=T}
par(mfrow = c(2, 1))
plot(pop.all,type="l",xlab="time (month)",ylab="population")
plot(infected.all, type='l',xlab="time (month)",
     ylab="prevalence of infection")
```

## Calculate the number of hospitalizations

### Take out all parameters needed to calculate hospitalizations^[This is not needed in our case since we already save them in the environment.]
```{r,echo=T}
q=1 # frequency dependent
# transmission probability per unit time
b= parms$baseline.txn.rate/(parms$dur.days1/30.44)
contact=parms$contact # c2 is the contact matrix
#transmission probability per unit time in each age group
beta <- (b/100)/(sum(yinit.matrix)^(1-q))*contact
Amp=parms$Amp # seasonal amplitude
phi=parms$phi # seasonal phase shift

rho1=parms$rho1 # Relative infectiousness (2nd)
rho2=parms$rho2 # Relative infectiousness (3rd+)
sigma1=parms$sigma1 # Relative risk of infection (2nd)
sigma2=parms$sigma2 # Relative risk of infection (3rd)
sigma3=parms$sigma3 # Relative risk of infection (4th+)
t0=nrow(results.burned) # length of time for evaluation
```

## Calculate the number of hospitalizations (continued)

### Take out all states needed to calculate hospitalizations (S and I)
```{r,echo=T}
I1 <- results.burned[,grep('I1', colnames(results.burned))]
I2 <- results.burned[,grep('I2', colnames(results.burned))]
I3 <- results.burned[,grep('I3', colnames(results.burned))]
I4 <- results.burned[,grep('I4', colnames(results.burned))]
S0 <- results.burned[,grep('S0', colnames(results.burned))]
S1 <- results.burned[,grep('S1', colnames(results.burned))]
S2 <- results.burned[,grep('S2', colnames(results.burned))]
S3 <- results.burned[,grep('S3', colnames(results.burned))]
```

### calculate the force of infection
```{r,echo=T}
#################Force of infection################
lambda1=matrix(0,nrow=t0,ncol=N_ages)
 for (t in 1:t0)
 {lambda1[t,]<-as.vector((1+Amp*cos(2*pi*(t-phi*12)/12))*
              ((I1[t,]+rho1*I2[t,]+rho2*I3[t,]+rho2*I4[t,])
               %*%beta)/sum(results.burned[t,]))}
## (see pages 4, 29, and 30 for detailed explanation)
```

## Calculate the number of hospitalizations (continued)

$$\textcolor{blue}{\lambda S_0 = \text{number of new first infection}}$$
$$\textcolor{blue}{\sigma_1\lambda S_1 = \text{number of new second infection}}$$
$$\textcolor{blue}{\sigma_2\lambda S_2 = \text{number of new third infection}}$$
$$\textcolor{blue}{\sigma_3\lambda S_3 = \text{number of new subsequent infection}}$$

number of hospitalizations = the probability of hospitalization given infection * the number of new infection
\
\
```{r,echo=T}
####Number of hospitalizations by age############## 
   H1=matrix(0,nrow=t0,ncol=N_ages)
   for (i in 1:N_ages){
     H1[,i]=hosp1[i]*S0[,i]*lambda1[,i]+
       hosp2[i]*sigma1*S1[,i]*lambda1[,i]+
       hosp3[i]*sigma2*S2[,i]*lambda1[,i]+
       hosp3[i]*sigma3*S3[,i]*lambda1[,i]}
```

## Plot the number of hospitalizations over time 
```{r,echo=T, fig.dim=c(5, 3)}
## time-series of number of hospitalizations ##
plot(1:108,rowSums(H1),type="l",xlab="months",
     ylab="simulated RSV hospitalizations")
## the decrease at the end is likely caused by 
## the drop of birth rate 0.0115 vs 0.013 (in 1980)
```

## Plot the age distribution of hospitalizations

```{r,echo=T, fig.dim=c(5, 3)}
## age distribution of hospitalizations ##
barplot(c(sum(colSums(H1)[1:3]),
          sum(colSums(H1)[4:6]),
          sum(colSums(H1)[7:9]),
          sum(colSums(H1)[10:12]),
          colSums(H1)[13:21])/sum(H1),
        xlab="age group",ylab="proportion")
```

### 2022-12-31 MSIS_model_explain/SimpleModel/simple-age-structure model.Rmd
preliminary tutorial material

---
title: "age structure"
author: "ZHE ZHENG, Dan Weinberger, modified from Matlab code by Virginia Pitzer"
date: "12/26/2020"
output: html_document
---
```{r}
library(deSolve)
library(RColorBrewer)
library(reshape2)
```
In Pitzer et al: The <12 month olds, were divided into monthy age classes. The remaining population was divided into 6 classes: 1–4 years old, 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old. In this example, the 1-4 year olds are divided into 12 month age classes as well.

```{r}
#T is N time points; N_ages is number of age groups
parmset <- readRDS('./data_and_parms/parmset.rds')
Pop1 <- readRDS('./data_and_parms/pop1.rds') #initial population
B <-  readRDS('./data_and_parms/Birth_rate.rds')
c2 <- readRDS( './data_and_parms/c2.rds') # contact matrix

```

#View contact matrix
```{r}
n.cols=100
nice.cols <-  colorRampPalette(brewer.pal(9, "YlOrRd"))(n.cols)
heatmap(c2/sum(diag(c2)), Rowv=NA, Colv=NA, scale='none', col=nice.cols)
```
Set names of Age groups 
```{r}
N_ages <- length(Pop1) 
agenames <- paste0('Agegrp', 1:N_ages) #Could replace this with vector of actual age names
```

```{r}
###########################################
# 1/duration of maternal immunity (DAYS)
DurationMatImmunityDays= 112
###########################################

###########################################
# update LRI and hospitalization probability
###########################################
al=21
delta1=c(rep(.40,3),rep(.39,3), rep(.21,3),rep(.20,3),0.16,rep(.14,3),rep(0.05,al-16))#proportion of first infections that are LRI (by age)
delta2=.5*delta1#proportion of second infectious that are LRI 
delta3=.7*delta2#proportion of third infections that are LRI
hosp1=c(.18*rep(.40,3),0.08*rep(.39,3),0.07*rep(.21,3),0.06*rep(.20,3),0.06*0.16,0.05*rep(.14,3),0.02*rep(0.05,al-16))#proportion of first LRIs that are hospitalized
hosp2=.4*hosp1#proportion of second LRIs that are hospitalized
hosp3=c(rep(0,al-2),0.00001,0.00004)#proportion of third LRIs that are hospitalized

```

## Population size by time
```{r}
p <- sum(Pop1)  # Total population at each time, a vector of length T
```

## Initialize the compartments (States) 
```{r}
StateNames <- c('M','S0','I1','S1','I2','S2','I3','S3','I4')
States <- array(NA, dim=c(N_ages, length(StateNames) )) #  N age groups xN parameters 
dimnames(States)[[1]] <- agenames
dimnames(States)[[2]] <- StateNames

yinit.matrix <- array(NA, dim=c(N_ages, length(StateNames) ))

dimnames(yinit.matrix)[[1]] <- agenames
dimnames(yinit.matrix)[[2]] <- StateNames

yinit.matrix[,c('S1','I2','S2','I3','S3','I4')]  = 0
yinit.matrix[,'M'] = c(Pop1[1:6], rep(0,N_ages-6))
yinit.matrix[,'S0'] = c(rep(0,6),Pop1[7:N_ages]-rep(N_ages-6)) 
yinit.matrix[,'I1'] = c(rep(0,6), rep(1,N_ages-6))  #initializes with 1 infected person per age group

yinit.vector <- as.vector(yinit.matrix) #Vectorize the ynit matrix

# Create array that has the labels by age, State and use this to name the yinit.vector
name.array <- array(NA, dim=dim(yinit.matrix))
for(i in 1:dim(name.array)[1]){
  for(j in 1:dim(name.array)[2]){
      name.array[i,j] <- paste(dimnames(yinit.matrix)[[1]][i],dimnames(yinit.matrix)[[2]][j]  )
    }
  }

name.vector <- as.vector(name.array)
names(yinit.vector) <- name.vector

```

## Read in model and parameters

```{r}
source('./data_and_parms/simple_model.R') #Read in the model
```

**NOTE: time step here is in months--need to adjust seasonality accordingly**
```{r}
 start_time = 1 # start date (years)
 tmax = nrow(B)
# end_time = 25 # end date (years)
my_times <- seq(start_time, tmax, by = 1) # gives a sequence from start to end
											 # in increments of 1
```

##Settings parameters
Most of these are described in Table 2 of Pitzer et al, PLOS Pathogens: https://journals.plos.org/plospathogens/article?id=10.1371/journal.ppat.1004591

For birth rate:
The weekly per capita birth rate (B_w) is equal to log(1+B)/52.18 because the data on the birth rate is annual and the differential equation model inherently assumes that growth is occurring exponentially. So if the annual birth rate is equal to 12 per 1000 per year (B=0.012), for example, then we are assuming that:

$N1=N0exp(B_w*52.10)$     where N1 is the population after 1 year and N0 is the baseline population, and $N1=N0*(1+B)$.

$1.012=1*exp(B_w*52.18)$

$B_w=log(1.012)/52.18$.

```{r setparms}
#########################################
#Matrix of dimension N_ages x N_ages
#beta = beta 
#########################################

#########################################
#Seasonal components--
#???Should prob be estimated?
b1=0.1830827 #Seasonal amplitude
phi=3.327749 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
#########################################

#########################################
#Relative infectiousness for 2nd and subsequent infections
rho1 = 0.75
rho2 = 0.51
#########################################

#########################################
# duration of infectiousness (months)
#Duration in days
dur.days1 <- 10 #days
dur.days2 <- 7 #days
dur.days3 <- 5 #days
###########################################

###########################################
# 1/duration of maternal immunity (DAYS)
DurationMatImmunityDays= 16*7
###########################################

###########################################
WidthAgeClassMonth = c(rep(1,times=12), rep(12,times=4),  60, 120, 240, 240, 240 )  #Aging rate=1/width age class (months) Vector of long N_age
############################################

############################################
#????
#What is time(-1) unit?
#um= -0.0001833333  #?? is this right? only die from last age class--this maybe helps with this "We assumed deaths occurred from the last age class and adjusted the net rate of immigration/emigration and death from other age groups in order to produce a rate of population growth and age structure similar to that of the US." 
#?????????
#From Giny's code: um=log(0.993)/52 #net rate of crude deaths (+) and immigration (-) from all age groups (per week): can adjust this to approximate population growth in state# I calibrated this parameter so we can reproduce the population growth
um= -0.0002227
#############################################

#############################################
#Birth rate (births/person/YEAR
#Matrix: T rows, N_ages columns; columns 2:N_ages all 0s
PerCapitaBirthsYear=B 
#############################################

#############################################
#Relaive risk of infection following 1st, 2nd, 3rd+ infections
sigma1=0.76
sigma2=0.6
sigma3=0.4
#############################################

##To ESTIMTE from DATA?: Baseline transmission rate
#############################################

baseline.txn.rate <- 9.0
q=1
c2=c2

```

#Save parameters in a list
```{r saveparms}
parms<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
            DurationMatImmunityDays=DurationMatImmunityDays,
            WidthAgeClassMonth=WidthAgeClassMonth,
            um=um,
            b1=b1,
            phi=phi,
            rho1=rho1,
            rho2=rho2,
            dur.days1=dur.days1,
            dur.days2=dur.days2,
            dur.days3=dur.days3,
            yinit.matrix=yinit.matrix,
            baseline.txn.rate = baseline.txn.rate,
            q=q,
            c=c2,
            sigma1=sigma1,
            sigma2=sigma2,
            sigma3=sigma3,
            time.step='month'
            )
```

## Run the model 
Note here we are just simulating with set parameters, not fitting to data
```{r}
results <- ode(y=yinit.vector, t=my_times,  
               func=simple_model, 
            parms=parms )

```

## Extract the relevant results
```{r}

#Pitzer used a 40-50 YEAR burn in period
burnN <- 20*12
results.burned <- results[-c(1:burnN),]

##Any infected person
infected.cols <- results.burned[,c(grep('I1', colnames(results.burned)),grep('I2', colnames(results.burned)),grep('I3', colnames(results.burned)),grep('I4', colnames(results.burned)  )) ]

infected.all <- apply(infected.cols,1, sum)

infected.cols.m <- melt(infected.cols)
infected.cols.m$agegrp <- sub(" .*", "",infected.cols.m$Var2 )
infected.cols.c <- dcast(infected.cols.m, Var1~agegrp, fun.aggregate = sum)

#Pop
all.m <- melt(results.burned[,grep('Agegrp',colnames(results.burned))])
all.m$agegrp <- sub(" .*", "",all.m$Var2 )
all.c <- dcast(all.m, Var1~agegrp, fun.aggregate = sum)
pop.all <- apply(all.c[,-1],1,sum)

plot(infected.all, type='l')
plot(pop.all,type="l")

```
```{r}
  St <- results[-c(1:300),-1]
   q=1
   b= parms$baseline.txn.rate/(parms$dur.days1/30.44)
   beta=(b/100)/(sum(yinit.matrix)^(1-q))*c2
   b1= parms$b1
   phi=parms$phi
   rho1=parms$rho1
   rho2=parms$rho2
   sigma1=parms$sigma1
   sigma2=parms$sigma2
   sigma3=parms$sigma3
   t0=108
   I1 <- St[,grep('I1', colnames(St))]
   I2 <- St[,grep('I2', colnames(St))]
   I3 <- St[,grep('I3', colnames(St))]
   I4 <- St[,grep('I4', colnames(St))]
   S1 <- St[,grep('S1', colnames(St))]
   S2 <- St[,grep('S2', colnames(St))]
   S3 <- St[,grep('S3', colnames(St))]
   S0 <- St[,grep('S0', colnames(St))]
   lambda1=matrix(0,nrow=t0,ncol=al)#Force of infection
    for (t in 1:t0)
    {lambda1[t,] <- as.vector((1+b1*cos(2*pi*(t-phi*12)/12))*((I1[t,]+rho1*I2[t,]+rho2*I3[t,]+rho2*I4[t,])%*%beta)/sum(St[t,]))}

   H1=matrix(0,nrow=t0,ncol=al)#Number of hospitalizations by age
   for (i in 1:al){
     H1[,i]=hosp1[i]*S0[,i]*lambda1[,i]+hosp2[i]*sigma1*S1[,i]*lambda1[,i]+hosp3[i]*sigma2*S2[,i]*lambda1[,i]+hosp3[i]*sigma3*S3[,i]*lambda1[,i]
   }

```

```{r}
plot(1:108,rowSums(H1),type="l",xlab="months",ylab="simulated RSV hospitalizations")
# need to get rid of the burn-in period

barplot(c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1),xlab="age group",ylab="Number of Hospitalizations")

# simulated age distribution Age groups: The <12 month olds, were divided into per 3-months age classes. The 1-4 year olds are  divided into 1yearly age classes, then 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old

Hosp_sim <- as.integer(rowSums(H1))
agedist_Sim <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1)
```

## fit the model to simulated data in order to estimate parameter sets
```{r}
parmset<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
            WidthAgeClassMonth=WidthAgeClassMonth,
            um=um,
            rho1=rho1,
            rho2=rho2,
            dur.days1=dur.days1,
            dur.days2=dur.days2,
            dur.days3=dur.days3,
            yinit.matrix=yinit.matrix,
            q=q,
            c=c2,
            sigma1=sigma1,
            sigma2=sigma2,
            sigma3=sigma3,
            time.step='month'
            )

fitmodel <-  function(parameters,dat) {   # takes the parameter values and dataset as inputs
  protrans <- parameters[1] # estimate parameter related to R0 (baseline transmission rate)
  amp <- parameters[2] # estimate parameter related to seasonal amplitude
  trans <- parameters[3] # estimate parameter related to seasonal peak timing
  DMD <- parameters[4] # estimate parameter related to the duration of maternal immunity
  b1 <- exp(amp) #ensure positive 
  baseline.txn.rate <- exp(protrans) #ensure positive
  phi <-  (2*pi*(exp(trans))) / (1+exp(trans)) # transform to its scale in the model
  durx <- exp(DMD) #ensure positive
  durx <<- durx #somehow we need this for the function and model to recognize these parameters
  baseline.txn.rate <<- baseline.txn.rate
  b1 <<- b1
  phi <<- phi

   # Simulate the model with initial conditions and timesteps defined above, and parameter values from function call
  results <- ode(y=yinit.vector, t=my_times,  
               func=simple_model, 
            parms=c(parmset,baseline.txn.rate=baseline.txn.rate,b1,phi,DurationMatImmunityDays=durx))
  t0 <- 108
  results<- tail(results,t0)
  St <- results[,-1]

   b=baseline.txn.rate/(parmset$dur.days1/30.44)
   beta=(b/100)/(sum(yinit.matrix)^(1-1))*c2

   lambda1=matrix(0,nrow=t0,ncol=al)#Force of infection
    for (t in 1:t0)
    {lambda1[t,] <- as.vector((1+b1*cos(2*pi*(t-phi*12)/12))*((St[t,(2*al+1):(3*al)]+rho1*St[t,(4*al+1):(5*al)]+rho2*St[t,(6*al+1):(7*al)]+rho2*St[t,(8*al+1):(9*al)])%*%beta)/sum(St[t,]))}

  H1=matrix(0,nrow=t0,ncol=al)#Number of hospitalizations by age
  for (i in 1:al){
    H1[,i]=hosp1[i]*St[,(al+i)]*lambda1[,i]+hosp2[i]*sigma1*St[,(3*al+i)]*lambda1[,i]+hosp3[i]*sigma2*St[,(5*al+i)]*lambda1[,i]+hosp3[i]*sigma3*St[,(7*al+i)]*lambda1[,i]
    }
  H <- rowSums(H1)

  agedist <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
            sum(colSums(H1)[7:9]),
            sum(colSums(H1)[10:12]),
            colSums(H1)[13:21])/sum(H1)

   LLall <- sum(dpois(x = dat, lambda =H, log = TRUE)) # fit to timeseries
   LLmulti <- dmultinom(x= agedist,prob = agedist_Sim,log = T) # fit to age distribution

   #prior
   durprior <- dgamma(x=durx,22,5,log=T)

   #total LL
   LL <- LLall+LLmulti+durprior

   return(LL)
}

fitLL <- optim(par = c(2.198522,  -1.638387 , -1.301047 ,  4.619743 ),           # starting values for beta and gamma - you should get the same result no matter which values you choose here
      fn = fitmodel,        # the distance function to optimise
      dat = Hosp_sim,         # the dataset to fit to ("dat" argument is passed to the function specified in fn)
      control = list(fnscale=-1)) # the log likelihood is negative; here we minimize the negative log likelihood
```

### 2022-12-31 MSIS_model_explain/SimpleModel/data_and_parms/simple_model.R
preliminary tutorial material

@@ -1,0 +1,153 @@
+simple_model <- function(t,y,parms,time.step='month'){
+  
+  # read in initial states and their names
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  # unify the time unit of parameter inputs
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  # waning rate of maternal immunity (by time step)
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+  
+  # aging rate (by time step)
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(WidthAgeClassMonth*4.345)}
+  
+  # rate of recovery of first infection
+  gamma1= 1/(parms$dur.days1/length.step) 
+  # rate of recovery of second infection
+  gamma2= 1/(parms$dur.days2/length.step)
+  # rate of recovery of third infection
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  
+  #gamma3 stands for rate of recovery from subsequent infection
+  
+  # Relative risk of infection (2nd)
+  sigma1=parms$sigma1 
+  # Relative risk of infection (3rd)
+  sigma2=parms$sigma2
+  # Relative risk of infection (4th+)
+  sigma3=parms$sigma3
+  
+  # Relative infectiousness (2nd)
+  rho1=parms$rho1 
+  # Relative infectiousness (3rd+)
+  rho2=parms$rho2
+  
+  #Pull out the states for the model as vectors
+  M <-  States[,'M'] # protected by maternal immunity
+  S0 <-  States[,'S0'] # purely susceptible population
+  I1 <-  States[,'I1'] # first time infection (infectious)
+  
+  S1 <-  States[,'S1'] 
+  # susceptible population with build-up immunity
+  I2 <-  States[,'I2'] # second time infection
+  
+  S2 <-  States[,'S2']
+  # susceptible population with lower risk of re-infection
+  I3 <-  States[,'I3'] # third time infection
+  
[…]

### 2022-12-31 MSIS_model_explain/SimpleModel/data_and_parms/waning_immunity.R

@@ -1,0 +1,168 @@
+waning_immunity_model <- function(t,y,parms, time.step='month'){
+  
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  # if(t>=322 & t<= 335){
+  #   I_ex_t= sum(States)*parms$ex_proportion*parms$ex_increase_new[t-321] # ex_increase_new is a vector indicating how 
+  #   # travel restritions and virus activities in other regions affects virus importation
+  # }else{
+  #   I_ex_t <- sum(States)*parms$ex_proportion 
+  #   ## external infections
+  #   ##  I_ex_t=1/100000*St ; I_ex_t=5/100000*St ; I_ex_t=10/100000*St and etc
+  # }
+  # 
+  
+  I_ex_t=parms$I_ex*parms$airtravel[t]
+  
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  if(parms$AllowWaning=='Yes'){
+  theta1= 1/(parms$dur.immunity1/length.step) # waning immunity from S3 to W3;from S2 to W2
+  theta2= 1/(parms$dur.immunity2/length.step) }
+  else{theta1= 0 # waning immunity from S3 to W3;from S2 to W2
+  theta2= 0}# waning immunity from W3 to S2;from W2 to S1
+  
[…]

### 2023-01-02 MSIS_model_explain/MSIS_model_explain.Rmd
add MLE section

@@
 ```
+
+## Section 2: use maximum likelihood estimation (MLE) to estimate the unknown parameters
+
+In this example, we will use maximum likelihood estimation (MLE) to estimate the (1) per capita transmission probability: baseline.txn.rate, (2) seasonal amplitude: Amp, (3) seasonal peak timing: phi, (4) duration of maternal immunity: DurationMatImmunityDays.
+\
+\
+We will treat the time-series and age distribution of RSV hospitalizations that we just simulated as observed status and calibrate our transmission dynamic models using MLE to estimate these "unknown" parameters.

### 2023-01-02 MSIS_model_explain/MSIS_model_explain.Rmd
add files to section 2 and 3 tutorials

--- 2022-12-31 MSIS_model_explain/MSIS_model_explain.Rmd
@@
 In this example, the 1-4 year old are divided into 12 month age classes as well.
+In under 1 year old, we calibrate the model to age category of every 3 month.
 \
@@
 \
-- In the second section, we will learn to use maximum likelihood estimation (MLE) to estimate the unknown parameters (one value for each unknown parameter). 
+- In the second section, we will learn to use maximum a posteriori / maximum likelihood estimation (MLE) to estimate the unknown parameters (one value for each unknown parameter). 
 \
@@
 # omega = 1/duration of maternal immunity 
-DurationMatImmunityDays= 112 #days
+DurationMatImmunityDays= 30 #days
 ###########################################
@@
 # per capita transmission probability
-baseline.txn.rate <- 9.2 
+baseline.txn.rate <- 9 
 # this is now for the entire infectious period
@@
 ## age distribution of hospitalizations ##
-barplot(c(sum(colSums(H1)[1:3]),
+barplot(c(sum(colSums(H1)[1:3]),# 0-2 months
+          sum(colSums(H1)[4:6]),# 3-5 months
+          sum(colSums(H1)[7:9]),# 6-8 months
+          sum(colSums(H1)[10:12]),# 9-11 months
+          colSums(H1)[13:21])/sum(H1), 
+        # 1,2,3,4 years old, 5–9 years,
+        # 10–19 years, 20–39 years, 
+        # 40–59-years, and 60+ years old.
+        xlab="age group",ylab="proportion")
+```
+
+## Section 2: use maximum a posteriori (variation of MLE) to estimate the unknown parameters
+
+In this example, we will use maximum a posteriori to estimate the (1) per capita transmission probability: baseline.txn.rate, (2) seasonal amplitude: Amp, (3) seasonal peak timing: phi, (4) duration of maternal immunity: DurationMatImmunityDays.
+\
+\
+We will treat the time-series and age distribution of RSV hospitalizations that we just simulated as observed status and calibrate our transmission dynamic models using MLE to estimate these "unknown" parameters.
+
+```{r, echo=T}
+###### save simulated time-series and age distribution
+Hosp_sim <- as.integer(rowSums(H1)) # time-series
+
+agedist_Sim <- c(sum(colSums(H1)[1:3]),
                  sum(colSums(H1)[4:6]),
@@
             sum(colSums(H1)[10:12]),
-          colSums(H1)[13:21])/sum(H1),
-        xlab="age group",ylab="proportion")
+            colSums(H1)[13:21])/sum(H1)
+            # age distribution
+
+sim <- list(Hosp_sim=Hosp_sim,agedist_Sim=agedist_Sim)
 ```
+
+## Create a list of known parameters
+
+```{r createparms, echo=T}
+parmset<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
+            WidthAgeClassMonth=WidthAgeClassMonth,
+            um=um, # net growth rate
+            rho1=rho1, # Relative infectiousness (2nd)
+            rho2=rho2, # Relative infectiousness (3rd+)
+            dur.days1=dur.days1,# Duration of infectiousness
+            dur.days2=dur.days2,# Duration of infectiousness
+            dur.days3=dur.days3,# Duration of infectiousness
+            yinit.matrix=yinit.matrix, # initial states
+            q=q, # Frequency or Density dependent
+            contact=c2, # contact matrix
+            sigma1=sigma1, # Relative risk of infection (2nd)
+            sigma2=sigma2, # Relative risk of infection (3rd)
+            sigma3=sigma3, # Relative risk of infection (4th+)
+            time.step='month',
+            t0=t0, # length of time for evaluation
+            N_ages=N_ages) # number of age group
+```
+
+## Create a function to calculate the log likelihood of given values
+
+### Read in the function
+
+```{r, echo=T}
+source('./code source/Loglikelihood_MSIS.R') 
+```
+
+### View log likelihood calculation function
+```{r,echo=T,eval=FALSE}
+fitmodel <-  function(parameters,dat) { 
+# takes the parameter values and dataset as inputs
+
+# parameter related to R0 (baseline transmission rate)
+  protrans <- parameters[1] 
+# parameter related to seasonal amplitude
+  b1 <- parameters[2] 
+# parameter related to seasonal peak timing
+  trans <- parameters[3]
+# parameter related to the duration of maternal immunity
+  DMD <- parameters[4] 
+```
+
+## Log likelihood calculation function (continued)
+
+```{r,echo=T,eval=FALSE}
+  Amp <- exp(b1) #ensure positive 
+  baseline.txn.rate <- exp(protrans) #ensure positive
+# transform to its scale in the model  
+  phi <-  (2*pi*(exp(trans))) / (1+exp(trans))
+  durx <- exp(DMD) #ensure positive
+
+# we need this for the function and model 
+# to recognize these parameters
+  durx <<- durx 
+  baseline.txn.rate <<- baseline.txn.rate
+  Amp <<- Amp
+  phi <<- phi
+
+# Run transmission model with initial conditions 
+# and time steps defined above, 
+# and parameter values from function call
+  results <- ode(y=yinit.vector, t=my_times,  
+                 func=simple_model, 
+                 parms=c(parmset,
+                 baseline.txn.rate=baseline.txn.rate,
+                 Amp=Amp,
+                 phi=phi,
+                 DurationMatImmunityDays=durx))
+```
+
+## Log likelihood calculation function (continued)
+
+```{r,echo=T,eval=FALSE}
+   t0 <- parmset$t0 
+  # make sure the evaluation period are the same
+  results<- tail(results,t0) # get rid of burn-in period
+  St <- results[,-1]
+
+  I1 <- St[,grep('I1', colnames(St))]
+  I2 <- St[,grep('I2', colnames(St))]
+  I3 <- St[,grep('I3', colnames(St))]
+  I4 <- St[,grep('I4', colnames(St))]
+  S0 <- St[,grep('S0', colnames(St))]
+  S1 <- St[,grep('S1', colnames(St))]
+  S2 <- St[,grep('S2', colnames(St))]
+  S3 <- St[,grep('S3', colnames(St))]
+
+b=baseline.txn.rate/(parmset$dur.days1/30.44)
+beta=(b/100)/(sum(yinit.matrix)^(1-parmset$q))*parmset$contact
+```
+
+## Outcomes of interests given parameter inputs (function continued)
+
+```{r,echo=T,eval=FALSE}
+  #Force of infection
+  lambda1=matrix(0,nrow=t0,ncol=parmset$N_ages)
+  for (t in 1:t0)
+  {lambda1[t,]<-as.vector((1+Amp*cos(2*pi*(t-phi*12)/12))*
+                 ((I1[t,]+rho1*I2[t,]+rho2*I3[t,]+rho2*I4[t,])
+                  %*%beta)/sum(St[t,]))}
+
+ H1=matrix(0,nrow=t0,ncol=parmset$N_ages)
+  for (i in 1:parmset$N_ages){
+    H1[,i]=hosp1[i]*S0[,i]*lambda1[,i]+
+      hosp2[i]*sigma1*S1[,i]*lambda1[,i]+
+      hosp3[i]*sigma2*S2[,i]*lambda1[,i]+
+      hosp3[i]*sigma3*S3[,i]*lambda1[,i]}
+  H <- rowSums(H1)
+
+  agedist <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
+               sum(colSums(H1)[7:9]),
+               sum(colSums(H1)[10:12]),
+               colSums(H1)[13:21])/sum(H1)
+```
+
+## Calculate the log likelihood of model outputs with given parameters (function continued)
+```{r,echo=T,eval=FALSE}
+LLall<-sum(dpois(x=sim$Hosp_sim,lambda=H,log=T)) 
+# fit to timeseries 
+# (number of cases follows poisson distribution)
+Lmulti<-dmultinom(x=sim$agedist_Sim,
+                  prob=agedist_Sim,log = T) 
+# fit to age distribution 
+# proportion (sum to 1) follows mutinomial distribution
+
+  #prior
+  durprior <- dgamma(x=durx,22,5,log=T)
+  # give a prior for the duration of maternal immunity
+
+  #total Loglikelihood (because of log, we sum up)
+  LL <- LLall+LLmulti+durprior
+
+  return(LL)
+}
+```
+
+## Use Optim function to maximum a posteriori (calibrate outputs of the transmission model to observed status)
+
+```{r,echo=T,eval=FALSE}
+fitLL <- optim(par=c(2.2,-1.6,1.2,3.16),
+  # remember to exponential these values
+  # to get the parameter estimates
+  # starting values for beta and gamma 
+  # - you should get the similar/same result 
+  # no matter which values you choose here
+      fn = fitmodel, # the distance function to optimise
+      dat = sim,  # the dataset we fit to 
+# ("dat" argument is passed to the function specified in fn)
+    control = list(fnscale=-1))# negative log likelihood
+# here we minimize the negative log likelihood
+```
+
+*"optimizers in statistical packages usually work by minimizing the result of a function...
+since the log likelihood and likelihood function have the same increasing or decreasing trend, you can minimize the negative log likelihood in order to actually perform the maximum likelihood estimate of the function you are testing."*^[https://stats.stackexchange.com/questions/141087/why-do-we-minimize-the-negative-likelihood-if-it-is-equivalent-to-maximization-o]
+
+## Compare the estimations and real values
+
+```{r,echo=T,eval=FALSE}
+fitLL # (check if the convergence = 0)
+# if not run for several times to make sure
+# the function did not stuck at local minimum
+
+exp(fitLL$par)
+# output: 8.99  0.20  3.3 (same as+/-pi) 
+# 23
+
+# Model input:
+# 9, 0.2, 3.33 30 days
+
+# estimations are very similar to model inputs
+```
+
+## Section 3: Stan model to estimate posterior samples
+
+Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the seperated tutorials in the same file path.

### 2023-01-02 MSIS_model_explain/code source/Loglikelihood_MSIS.R
add files to section 2 and 3 tutorials

@@ -1,0 +1,83 @@
+fitmodel <-  function(parameters,dat) {   # takes the parameter values and dataset as inputs
+ 
+  protrans <- parameters[1] # estimate parameter related to R0 (baseline transmission rate)
+  b1 <- parameters[2] # estimate parameter related to seasonal amplitude
+  trans <- parameters[3] # estimate parameter related to seasonal peak timing
+  DMD <- parameters[4] # estimate parameter related to the duration of maternal immunity
+  Amp <- exp(b1) #ensure positive 
+  baseline.txn.rate <- exp(protrans) #ensure positive
+  phi <-  (2*pi*(exp(trans))) / (1+exp(trans)) # transform to its scale in the model
+  durx <- exp(DMD) #ensure positive
+  
+  #we need this for the function and model to recognize these parameters
+  durx <<- durx 
+  baseline.txn.rate <<- baseline.txn.rate
+  Amp <<- Amp
+  phi <<- phi
+  
+  
+  # Run transmission model with initial conditions 
+  # and timesteps defined above, and parameter values from function call
+  results <- ode(y=yinit.vector, t=my_times,  
+                 func=simple_model, 
+                 parms=c(parmset,
+                         baseline.txn.rate=baseline.txn.rate,
+                         Amp=Amp,
+                         phi=phi,
+                         DurationMatImmunityDays=durx))
+  
+  t0 <- parmset$t0 
+  # make sure the evaluation period are the same
+  results<- tail(results,t0) # get rid of burn-in period
+  St <- results[,-1]
+  
+  I1 <- St[,grep('I1', colnames(St))]
+  I2 <- St[,grep('I2', colnames(St))]
+  I3 <- St[,grep('I3', colnames(St))]
+  I4 <- St[,grep('I4', colnames(St))]
+  S0 <- St[,grep('S0', colnames(St))]
+  S1 <- St[,grep('S1', colnames(St))]
+  S2 <- St[,grep('S2', colnames(St))]
+  S3 <- St[,grep('S3', colnames(St))]
+  
+  
+  b=baseline.txn.rate/(parmset$dur.days1/30.44)
+  beta=(b/100)/(sum(yinit.matrix)^(1-parmset$q))*parmset$contact
+  
[…]

### 2023-01-02 MSIS_model_explain/MSIS_model_explain.Rmd
add stan codes section

@@
 ```{r,echo=T,eval=FALSE}
-fitLL <- optim(par=c(2.2,-1.6,1.2,3.16),
+fitLL <- optim(par=c(2.2,-1.6,0.1,3.5),
   # remember to exponential these values
@@
 
-exp(fitLL$par)
-# output: 8.99  0.20  3.3 (same as+/-pi) 
-# 23
+exp(fitLL$par)[c(1:2)] # trans prob and Amp est
+phi_est = (2*pi*(exp(fitLL$par[3]))) / (1+exp(fitLL$par[3]))
+dur_est=exp(fitLL$par)[4]
+# output: 8.99  0.20  
+# 3.33 (same as+/-pi) 
+# 23.4 days
 
 # Model input:
-# 9, 0.2, 3.33 30 days
+# 9, 0.2, 3.33, 30 days
 
@@
 
-Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the seperated tutorials in the same file path.
+Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the separated tutorials in the same file path. 
+\
+\
+You will find three code documents inside code_source/stan_code file.
+age-structureMSIS_deterministic.stan is the stan code contains the ODE function, input data, input parameters, priors and the distribution of posterior distributions.
+ageMSIS_deterministric_annual.R is the R code that create the interface between stan and R to sample from posterior distribution.
+msis_wa_D.sh is a file to run the model on cluster.
+We will start with the stan code.
+\
+\
+I suggest you first run the model in local for short iterations for debug purpose. If the stan model works, you can move to HPC by using the msis_wa_D.sh script.
+
+## Stan code in details (page 55 to 68)
+Function: page 55 to 60.
+```{stan,output.var="model",echo=T,eval=FALSE}
+// function is the similar as our simpleModel.R
+// except for the syntax.
+functions {
+  real[] msis (real t, real[] y, real[] theta,
+             real[] x_r, int[] x_i) {
+// t for time
+
+// state, the volumes in each compartment, y;
+
+// theta, variables used to compute function
+// which depend on the model parameters;
+
+// x_r, real variables used to evaluate f
+// which only depend on fixed data;
+
+// x_i, integer values used to evaluate f
+// which only depend on fixed data.
+
+      int agegroups = x_i[1];
+      int q=x_i[2]; // Freq or Density dependent  
+```
+
+## You need to specify the variable type and locations of the parameter for each value
+Note, remember to add **;** after each line of code.
+\
+```{stan,output.var="model",echo=T,eval=FALSE}
+real birthrate[agegroups] = x_r[1:agegroups];
+real um = x_r[agegroups+1]; // net growth
+real rho1 = x_r[agegroups+2];
+real rho2 = x_r[agegroups+3];
+real gamma1 = x_r[agegroups+4];
+real gamma2 = x_r[agegroups+5];
+real gamma3 = x_r[agegroups+6];
+real sigma1 = x_r[agegroups+7];
+real sigma2 = x_r[agegroups+8];
+real sigma3 = x_r[agegroups+9];
+real u[agegroups]= x_r[(agegroups+10):(9+2*agegroups)];
+//aging rate
+real c2[agegroups*agegroups]=
+      x_r[(10+2*agegroups):(9+2*agegroups+agegroups*agegroups)];
+      // contact vector 
+// (I have not yet figured out how to input matrix)
+```
+
+## When create a new variable, you also need to specify the length of the variable 
+```{stan,output.var="model",echo=T,eval=FALSE}
+// parameters to estimate
+real beta = theta[1];
+real b1 = theta[2];
+real phi = theta[3];
+real omega = theta[4];
+
+// outcomes of interests that need to be recorded      
+real dydt[9*agegroups]; // state change
+real lambda[agegroups]; // force of infection
+real InfectN[agegroups];// number of infectious individuals
+
+// seasonal component
+real season_txn = (1+b1*cos(2*pi()*(t-phi*12)/12));
+```
+
+## Write down ordinary differential equations
+```{stan,output.var="model",echo=T,eval=FALSE}
+for (k in 1:agegroups) {
+      InfectN[k] = (y[k+2*agegroups]+
+      rho1*y[k+4*agegroups]+
+      rho2*y[k+6*agegroups]+
+      rho2*y[k+8*agegroups]);
+      } // (I1+rho1*I2+rho2*I3+rho2*I4)
+
+for (a in 1:agegroups) {
+  // force of infection    
+  lambda[a] = season_txn*beta*gamma1/sum(y)*
+  sum(to_vector(c2[((a-1)*agegroups+1):(a*agegroups)])
+  .*to_vector(InfectN));
+
+   //correspond to M.     
+  dydt[a] =  log(birthrate[a]+1)/12*sum(y) - 
+  (omega+u[a]+um)*y[a];
+```
+
+## Ordinary differential equations (continued)
+corresponding R code: [Source function of the transmission dynamic model (continued)] except for the aging in process (include in the next page).
+```{stan,output.var="model",echo=T,eval=FALSE}
+//S0
+  dydt[a+agegroups] =  omega*y[a] - 
+  lambda[a]*y[a+agegroups] - 
+  (um+u[a])*y[a+agegroups];
+
+//I1  
+ dydt[a+2*agegroups] =  lambda[a]*y[a+agegroups] - 
+ gamma1*y[a+2*agegroups] - 
+ (um+u[a])*y[a+2*agegroups];
+
+//S1 
+dydt[a+3*agegroups] =  gamma1*y[a+2*agegroups] - 
+sigma1*lambda[a]*y[a+3*agegroups] - 
+(um+u[a])*y[a+3*agegroups];
+```
+
+We will skip explaining the rest of the stan code of ordinary differential equations in this tutorial as they are the direct translation of the corresponding R codes.
+
+## Aging in process
+In age group >= 1 month, they receive the inflows from the younger groups at rate u. 
+```{stan,output.var="model",echo=T,eval=FALSE}
+if (a >1 ){
+dydt[a] = dydt[a]+ u[a-1]*y[a-1];
+
+dydt[a+agegroups] = dydt[a+agegroups]+ 
+u[a-1]*y[a+agegroups-1];
+
+dydt[a+2*agegroups] = dydt[a+2*agegroups] + 
+u[a-1]*y[a+2*agegroups-1];
+
+//...//
+
+dydt[a+8*agegroups] = dydt[a+8*agegroups]+
+u[a-1]*y[a+8*agegroups-1];}
+
+ }
+    return dydt;
+  }
+} // This is the end of the ODE function
+```
+
+## Data block
+
+Compared to R, the data block in stan code contains both fixed parameter values and also input observed data.
+
+```{stan,output.var="model",echo=T,eval=FALSE}
+// Fixed data is declared in the data block:
+data {
+  int<lower=1> n_months;
+  int<lower=1> agegroups;
+  int q;
+  real y0[9*agegroups];
+  real t0;
+  real ts[n_months];
+  int N;
+  real birthrate[agegroups];
+  real um;
+//...omit some parameters and data to fit into slides//
+  int<lower=1> hosp_cases[n_months];
+  int hosp_age[8];
+  real hosp1[agegroups];
+  real c2[agegroups*agegroups];
+}
+```
+
+## Transformed data block
+Link data block to function inputs: real[ ] x_r, int[ ] x_i
+```{stan,output.var="model",echo=T,eval=FALSE}
+transformed data {
+  real x_r[204];
+  int x_i[2];
+  x_i[1] = agegroups;
+// the first value of x_i is the number of age groups
+  x_i[2] = q;
+// the second value of x_i is parameter q
+
+  x_r[1:agegroups]= birthrate;
+// the first value of x_r is birth rate
+  x_r[agegroups+1]=um;
+//likewise, omit some parameters and data to fit into slides//
+  x_r[agegroups+9]=sigma3;
+  x_r[(agegroups+10):(9+2*agegroups)]=u;
+  x_r[(10+2*agegroups):(9+2*agegroups+agegroups*agegroups)]=c2;
+}
+```
+
+## Parameter block
+Specify the parameters we would like to estimate and their ranges.
+```{stan,output.var="model",echo=T,eval=FALSE}
+parameters {
+  real<lower=0> beta; // transmission prob
+  real<lower=0,upper=1> b1; // Amp
+  real<lower=0,upper=2*pi()> phi; // phase
+  real<lower=0> omega;
+  // waning rate of maternal immunity
+  real<lower=0,upper=1> report_ratio;
+  // report ratio of RSV hospitalizations
+  }
+```
+
+## Transformed parameters clock
+
+This part of stan codes corresponds to using ode() function in r to get the transmission states of each age group in each time step and then calculate the number of RSV hospitalizations and age distribution.
+```{stan,output.var="model",echo=T,eval=FALSE}
+transformed parameters{
+real y[n_months, 9*agegroups]; 
+// y is the states matrix that has row length=n_months and columns=9.
+  real rel_inf[n_months,agegroups];
+// number of infectious individuals
+  real lambda[n_months, agegroups]; // force of infection
+  // outcomes
+  // Number of hospitalizations by age
+  matrix[n_months,agegroups] HOSP_AGE;// corresponds to H1
+  vector[8] output_hosp_age; 
+  // age distribution of RSV hosp in children
+  vector[n_months] total_children; 
+  // total hospitalizations in children <5 years
+  vector<lower = 0>[n_months] output_hosp; 
+  //number of RSV hospitalizations by month
+```
+
+## Transformed parameters clock (continued)
+```{stan,output.var="model",echo=T,eval=FALSE}
+  real theta[4]; // parameters that need to be estimate
+  {
+    theta[1] = beta;
+    theta[2] = b1;
+    theta[3] = phi;
+    theta[4] = omega;
+
+    y = integrate_ode_bdf(msis, y0, t0, ts, theta, x_r, x_i,
+    1.0E-10, 1.0E-10, 1.0E3);
+    // relative tolerance for the ODE solver = 1.0E-10
+    // absolute tolerance for the ODE solver = 1.0E-10
+    // maximum number of steps to take in the ODE solver
+    // = 1.0E3
+    // these prevent stan takes too long to run
+    }
+```
+
+## Calculate force of infection (continued transformed parameters)
+
+```{stan,output.var="model",echo=T,eval=FALSE}
+for(i in 1:n_months){
+  for(j in 1:9*agegroups){
+    if (y[i,j] <= 0.0) y[i,j] = 1e-12;}
+// this prevent negative values in states
+
+      for (k in 1:agegroups) {
+      rel_inf[i,k] = (y[i,k+2*agegroups]+
+      rho1*y[i,k+4*agegroups]+
+      rho2*y[i,k+6*agegroups]+
+      rho2*y[i,k+8*agegroups]);
+      }// (I1+rho1*I2+rho2*I3+rho2*I4)
+
+// calculate force of infection; same as inside function     
+      for (a in 1:agegroups) {
+       lambda[i,a] = (1+b1*cos(2*pi()*(i-phi*12)/12))*
+       beta*gamma1/sum(y[i,])*
+       sum(to_vector(c2[((a-1)*agegroups+1):(a*agegroups)])
+       .*to_vector(rel_inf[i,]));
+```
+
+## Calculate hospitalizations (continued transformed parameters)
+
+```{stan,output.var="model",echo=T,eval=FALSE}
+HOSP_AGE[i,a] = hosp1[a]*lambda[i,a]*y[i,a+agegroups]
+			+hosp2[a]*lambda[i,a]*sigma1*y[i,a+3*agegroups]	
+			+hosp3[a]*lambda[i,a]*sigma2*y[i,a+5*agegroups]
+			+hosp3[a]*lambda[i,a]*sigma3*y[i,a+7*agegroups];
+       } // same as H1 calculation
+
+  output_hosp[i] = report_ratio*sum(HOSP_AGE[i,]);
+  // total reported RSV hospitalizations each month
+
+  total_children[i]=HOSP_AGE[i,1]+HOSP_AGE[i,2]+
+  HOSP_AGE[i,3]+HOSP_AGE[i,4]+HOSP_AGE[i,5]+HOSP_AGE[i,6]+
+  HOSP_AGE[i,7]+HOSP_AGE[i,8]; 
+  // total hospitalizations in children <5 years
+    }
+for (b in 1:8) {
+ output_hosp_age[b] = sum(HOSP_AGE[,b])/sum(total_children);}} 
+ // age distribution of hospitalizations in children <5 years
+```
+
+## Model block
+
+(1) specify the priors of the parameters we would like to estimate and their ranges.
+/
+/
+(2) specify the sample distributions of the posteriors.
+```{stan,output.var="model",echo=T,eval=FALSE}
+
+model {
+  //priors
+  beta ~ normal(3,1) T[1,5]; //truncated at 0
+  b1 ~ normal(0.2,0.05) T[0.05,1]; //truncated at 0,1
+  phi ~ normal(1.4,0.5) T[0,2*pi()];
+  omega ~ lognormal(-1,0.6) T[0,5];
+  report_ratio ~ beta(2,2); //truncated at 0,1
+
+  //sampling distribution
+  for(i in 1:n_months){
+  target += poisson_lpmf(hosp_cases[i]| output_hosp[i]);}
+
+  target += multinomial_lpmf (hosp_age|output_hosp_age);
+}
+```
+
+## R function to call stan code
+
+```{r,echo=T,eval=FALSE}
+library(rstan) # function to link stan code and R code
+rstan_options (auto_write = TRUE) 
+# this code is recommended for running stan in HPC
+
+setwd("~/project/Rstan/WA") 
+# this is where I save the data and all these codes
+data_msis<- readRDS("data_msis_wa_annual.rds")
+# read in input data. 
+# This would come from the inpatient data sets
+# and known parameters
+
+print(paste("number of cores:", future::availableCores()))
+# this is to check the cores you can use in HPC/local
+
+# call the stan code
+model <- stan_model("age-structureMSIS_deterministic.stan")
+
+# set the initial values as the output of MLE function
+# This helps the running time
+init_fun <- function() list(beta=3.17,b1=0.198,
+                    phi=1.46,omega=0.61,hosp_ratio=0.76)
+```
+
+## R function to sample the posteriors from stan code
+
+```{r,echo=T,eval=FALSE}
+## this code samples the posteriors of parameters of interests
+fit_msis <- sampling(model, # the stan model
+            data = data_msis, 
+          # inpatient data and known parameters
+            iter = 2000,# total iterations
+			      warmup = 1000, #iterations discard
+            chains = 4, # total chains
+            seed = 112, # for reproducible purpose
+            sample_file = file.path(getwd(), 
+              'WA_MSIS_12_deterministic.csv'),
+          # this is where the posterior samples are saved
+            cores=4,#parallel: one core for one chain
+            open_progress=T, # to see immediate output
+			      init= init_fun, # initial values
+            control = list(stepsize=0.1)) # sample step size 
+```
+
+## Shell script (bash codes, Sample SLURM Scripts) to run R and STAN on high performance computer (HPC)
+```{r,echo=T, engine = 'bash',eval=FALSE}
+# indicate the script uses the bash shell.
+#!/bin/bash
+
+#SBATCH --time=30-  
+# This tells the cluster the running time for your script. 
+# Don't set it too high or your job will get low priority. 
+#SBATCH --mail-type=ALL 
+# This tells the cluster what results should send to your email.
+#SBATCH --mail-user=zhe.zheng@yale.edu 
+# This is your email address
+#SBATCH --cpus-per-task=4 
+# This is for running in parallel. 
+# Set the cores to the number of chains you have
+#SBATCH --mem-per-cpu=30000 
+# This is for memory for each core
+
+module load R/4.1.0-foss-2020b 
+# you first need to check the r version available on cluster
+Rscript ageMSIS_deterministric_annual.R # R script to run
+```

### 2023-01-02 MSIS_model_explain/code source/stan_code/ageMSIS_deterministric_annual.R
add stan codes section

@@ -1,0 +1,24 @@
+library(rstan)
+library(gridExtra)
+rstan_options (auto_write = TRUE)
+
+setwd("~/project/Rstan/WA")
+
+data_msis<- readRDS("data_msis_wa_annual.rds")
+
+print(paste("number of cores:", future::availableCores()))
+
+model <- stan_model("age-structureMSIS_deterministic.stan")
+
+init_fun <- function() list(beta=3.17,b1=0.198,phi=1.46,omega=0.61,hosp_ratio=0.76)
+
+fit_msis_poisson <- sampling(model,
+                             data = data_msis,
+                             iter = 2000,
+			     warmup = 1000,
+                             chains = 4, 
+                             seed = 112,sample_file = file.path(getwd(), 'WA_MSIS_12_deterministic.csv'),
+                             cores=4,
+                             open_progress=T,
+			     init= init_fun,
+                	     control = list(stepsize=0.1))

### 2023-01-02 MSIS_model_explain/code source/stan_code/msis_wa_D.sh

@@ -1,0 +1,11 @@
+#!/bin/bash
+
+#SBATCH --time=10-
+#SBATCH --mail-type=ALL
+#SBATCH --mail-user=zhe.zheng@yale.edu
+#SBATCH --cpus-per-task=4
+#SBATCH --mem-per-cpu=30000
+
+
+module load R/4.1.0-foss-2020b
+Rscript ageMSIS_deterministric_annual.R

### 2023-01-04 RSV_timing_explain/RSV_timing_tutorial.Rmd
add more materials (discussion point with Deus next week?)

---
title: "Estimate RSV onset and peak timing"
subtitle: "Tutorial 2 for transition"
author: 
- Gigi (Zhe Zheng)^[zhe.zheng@yale.edu; zhe.zheng@aya.yale.edu; gigi.zhe.zheng@gmail.com]
- Dan (Daniel Weinberger) 
- Ginny (Virginia Pitzer)
date: "2023-01-04"
fontsize: 10pt
output: 
  beamer_presentation:
  theme: CambridgeUS
  colortheme: "beaver"
  fonttheme: "structurebold"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T)
```

## Background

```{r figure1,echo = F,out.width = "100%", out.extra="page=4"}
knitr::include_graphics(path = "Epidemics_conference.pdf")
```

## Outline

- In section 1, we will first introduce how to find the peak timing of RSV epidemics using harmonic regression (given regular annual/biennial seasonality). We will then talk about identifying the onset of RSV epidemics using second derivative method, regardless of the seasonality of RSV.
\
\
- In section 2, we will learn how to use R to identify peak timing and onset of RSV epidemics.
\
\
- In section 3, we will incorporate the spatial component of RSV epidemics. We will first introduce the concept of spatial autocorrelation and then lean to use R to account for spatial autocorrelation.

## References

Relevant readings:
\
\
- [RSV onset timing at county level](https://doi.org/10.1093/cid/civ331)
\
\
- [RSV peak timing on state level](https://doi.org/10.1371/journal.ppat.1004591)
\
\
- [Comparing RSV onset timing before and during the COVID-19 pandemic](https://doi.org/10.1111/irv.12965)
\
\
- [RSV peak timing at ZIP code level and the drivers of RSV spread](https://doi.org/10.1126/sciadv.abd6421)

### 2023-01-05 RSV_timing_explain/Lecture 11/r examples.Rmd
add Dan and Ginny's class materials

---
title: "Quantitative Methods in Infectious Disease"
subtitle: "Time series examples"
author: "Dan Weinberger"
date: 'November 28, 2018'
runtime: shiny
output: ioslides_presentation
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
require(RCurl)
require(shiny)
set.seed(123)
```

## Examples

Here are some worked examples of what we are covering in class

## Generating harmonic curves
-First create an index variable for time (1,2,3...t)
-Then use sine and cosine functions to generate a wave with specified period
-12 month harmonic (sine and cosine waves)
```{r harmonics, echo=TRUE}
t<-1:120
sin12<-sin(2*pi*t/12)
cos12<-cos(2*pi*t/12)
plot(sin12, type='l', col='red')
points(cos12, type='l', lty=2, col='blue')
```

##Vary the period 
```{r period.slider, echo=FALSE}
#library(shiny)
shinyApp(

  ui = fluidPage(
    sliderInput("period", "Period:",
               min=2, max=36, value=12),
    plotOutput("periodPlot")
  ),

  server = function(input, output) {
    output$periodPlot = renderPlot({
      plot(sin12, type='l', col='gray')
      points(sin(2*pi*t/input$period),type='l', col='red', xlab='time' )
      title(paste0('sin(2*pi*t/',input$period,')'))
    },width = "auto", height = "auto")
  }
)
```

#Harmonic Regression

##Generate count data with known periodicity
-120 months total (10 years of data)
-Has both a 10 and 20 month cycle
-Amplitude=1.5
```{r gen_data , echo=TRUE}
n=120
t<-1:n
set.seed(123)
#Generate some  data that has a frequency of 1/10 and 1/20  frequency (period of 10 ,10 )
amp1=0.75
freq1=1/12 # high frequency

amp2=1.5
freq2=1/24 #low frequency

xt1a=amp1*cos(2*3.14159*t*freq1) #component 1
xt1b=amp2*cos(2*3.14159*t*freq2) #component 2
intercept<-0.0
lambda<- exp(intercept+xt1a+xt1b)
xt1=rpois(n,lambda)

```

## View the data we generated
```{r plots1, echo=FALSE}
par(mfrow=c(2,2), mar=c(2,2,1,1))
plot(xt1a, bty='l' , main='Component 1', type='l' , ylim=range(c(xt1a,xt1b)))
plot(xt1b, bty='l' , main='Component 2' , type='l', ylim=range(c(xt1a,xt1b)))
plot(xt1a+xt1b, bty='l' , main='log(Component 1 +Component 2)' , type='l', ylim=range(xt1a+xt1b))
plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",ylab="Cases of Disease", xlab="Month(n)", main="Cases")
```

##Create the needed harmonic variables with 10, 20, and 40 month periods
```{r harm1, echo=TRUE}
t<-1:120
#Create harmonic variables
sin12=sin(2*pi*t/12)
cos12=cos(2*pi*t/12)
sin24=sin(2*pi*t/24)
cos24=cos(2*pi*t/24)
sin48=sin(2*pi*t/48)
cos48=cos(2*pi*t/48)
```
plot(xt1)

## Fit a simple poisson regression with just 12 month period
```{r harm_reg1}
fit1a <- glm(xt1 ~ sin12+cos12, family='poisson' )
pred1a<- fitted(fit1a)
#summary(fit1a)
par(mfrow=c(1,1))
plot (t, xt1,bty="l", xaxt="n",yaxt="n",xlab="Time",ylab="Cases of Disease", type="p")
points(t,pred1a, type="l",lty=1,col="blue", lwd=3)
```

## Add in 24 month periodicity
```{r harm_reg2}
fit2a <- glm(xt1 ~ sin12+cos12 +sin24 +cos24, family='poisson' )
pred2a<- fitted(fit2a)
#summary(fit2a)
par(mfrow=c(1,1))
plot (t, xt1,bty="l", xaxt="n",yaxt="n",xlab="Time",ylab="Cases of Disease", type="p")
points(t,pred2a, type="l",lty=1,col="blue", lwd=3)
```

## Add in 48 month period 
```{r harm_reg3}
fit3a <- glm(xt1 ~ sin12+cos12 +sin24 +cos24 +sin48 +cos48, family='poisson' )
pred3a<- fitted(fit3a)
#summary(fit3a)
par(mfrow=c(1,1))
plot (t, xt1,bty="l", xaxt="n",yaxt="n",xlab="Time",ylab="Cases of Disease", type="p")
points(t,pred3a, type="l",lty=1,col="blue", lwd=3)
```

## Determine best model with AIC (smaller=better)
Winner is Model 2 (12 and 24 month period). Models 2 and 3 have similar AIC score (within 2 points), so we prefer the simpler model
```{r aic}
 AIC(fit1a)
 AIC(fit2a)
 AIC(fit3a)
```

#Calculate amplitude of the 10 and 20 month periods and their ratio
```{r amp}
beta_sin12<-coef(fit2a)['sin12']
beta_cos12<-coef(fit2a)['cos12']
beta_sin24<-coef(fit2a)['sin24']
beta_cos24<-coef(fit2a)['cos24']

amp12<-sqrt(beta_sin12^2 +beta_cos12^2)
amp24<-sqrt(beta_sin24^2 +beta_cos24^2)
ratio12_24<-amp24/amp12
ratio12_24
amp12 #True value=0.75
amp24 #True value=1.5
```
## log(x+1)=bo+b1sin12+b2cos12+b3sin24+b4cos24

##Fast Fourier Transform
If we don't already know what periods should be present, can we use first use FFT to detect the correct period?
Can sqrt or log stranform data prior to FFT
```{r fft1, echo=TRUE}
# Do a FFT
fft1=fft(sqrt(xt1))
fft1[1:10] 
```
Generates a bunch of imaginary numbers that are hard to interpret

##Function to convert raw power spectrum into something interpretable
```{r fft_func, echo=FALSE}
plot.frequency.spectrum <- function(X.k, xlimits=c(5,n/2)) {
  plot.data  <- cbind(0:(length(X.k)-1), Mod(X.k))

  plot.data[2:length(X.k),2] <- 2*plot.data[2:length(X.k),2] 
  plot.data[2:length(X.k),1] <- n/plot.data[2:length(X.k),1]

  plot(plot.data, t="h", lwd=2,  
       xlab="Period", ylab="Strength", 
       xlim=xlimits, ylim=c(0,max(Mod(plot.data[,2]))), bty='l', main="FFT power spectrum")
}
```

#FFT power spectrum
```{r plot_fft}
par(mfrow=c(1,1))
  plot.frequency.spectrum(fft1, xlimits=c(2,30))

```

##FFT slider
```{r fft.slider, echo=FALSE}
#library(shiny)
shinyApp(

  ui = fluidPage(
    sidebarLayout(position = "left",
    sidebarPanel(
    sliderInput("period1", "Period1:",
               min=2, max=48, value=12),
     sliderInput("amplitude1", "Amplitude1:",
               min=0.2, max=2.5, value=0.75),
    sliderInput("period2", "Period2:",
               min=2, max=48, value=24),
      sliderInput("amplitude2", "Amplitude2:",
               min=0.2, max=2.5, value=1.5),
      sliderInput("int1", "Intercept:",
               min=0, max=5, value=0)
      ),
    mainPanel(
       plotOutput("periodPlot")
    )
  )
  ),

  server = function(input, output) {
    t=1:120
    xt1a=reactive({input$amplitude1*cos(2*3.14159*t/input$period1)}) #component 1
    xt1b=reactive ({input$amplitude2*cos(2*3.14159*t/input$period2)}) #component 2
    intercept<-reactive({input$int1})
    lambda<- reactive ({exp(intercept()+xt1a()+xt1b())})
    xt.slide=reactive ({rpois(n,lambda())})
    fft.s=reactive ({fft(sqrt(xt.slide()))})

    output$periodPlot = renderPlot({
      par(mfrow=c(2,1), mar=c(3,2,1,1))
   plot(t, xt.slide(),type="l", lwd=3, bty="l",ylab="Cases of Disease", xlab="Month(n)", main="Cases")
      plot.frequency.spectrum(fft.s(), xlimits=c(2,max(input$period1, input$period2)))

    },width = "auto", height = "auto")
  })
```

#Wavelets

##Pseudo-RSV data: Simulate time series with 12 and 24 month periods
Imaine this is RSV case data from 2 states, and we want to investigate the
epidemic characteristics in these states and the lag between states
```{r waveset, echo=FALSE}
set.seed(123)
  n=120 # 10 years
  t <- seq(1,n)
 amp1=2.5
 freq1=1/12 # high frequency
 amp2=0.5
 freq2=1/24 #low frequency

 xt1a=amp1*cos(2*3.14159*t*freq1)
 xt1b=amp2*cos(2*3.14159*t*freq2)

#other series shifted by 3 months and with smaller biennial component
 xt2a=amp1*cos(2*3.14159*t*freq1+1)
 xt2b=0.5*amp2*cos(2*3.14159*t*freq2+1)

 #Simulate some poisson count data
 xt1=rpois(n,exp((1+xt1a + xt1b)))
 xt2= rpois(n,exp(2+xt2a+xt2b))
 plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",ylab="Cases of RSV", xlab="Month(n)", ylim=range(c(xt1,xt2)))
 points(t,xt2, type="l",col="red")
```

```{r wave1, echo=FALSE}
##Source the Wavelet functions from github. Originally from Johannson PLOS Medicine https://journals.plos.org/plosmedicine/article?id=10.1371/journal.pmed.1000168
urlX<-getURL("https://raw.githubusercontent.com/weinbergerlab/quant_meth_id/master/WaveletPkg.R")
eval(parse(text = urlX))
#source('WaveletPkg.R')
```

##Run wavelets on the sqrt(series) we generated
```{r wave2, echo=TRUE}
	wave1<- as.data.frame(cbind("x"=(sqrt(xt1)), "t"=t))
	cwt1 <- complete.cwt(wave1, dj=1/10)
	plot.cwt(cwt1)
		wave2 <-as.data.frame(cbind("x"=(sqrt(xt2)), "t"=t)) 
	cwt2 <- complete.cwt(wave2, dj=1/10)
		plot.cwt(cwt2)

```

##Compare phase angle of the 2 series we generated
```{r wave3, echo=FALSE}

	phase1<-phase(cwt1, s=c(10, 14))
	phase2<-phase(cwt2, s=c(10, 14))
	plot(t, phase1,type='l', col="blue", lty=1)
	points(t, phase2,type='l', col="red", lty=1)
```

##Calculate and view the phase difference
```{r wave4, echo=TRUE}
#Take modulo (%%) of phase difference to account for edge effects as described by Grenfell. This corrects for problems at start and begining of cycles
	phase.diff2.1= phase2-phase1
	phasediff2.1_correct <- ((phase.diff2.1 + 3*pi) %% (2*pi)) - (pi)
	#shows corrected phase diff in modulo for the 12 month phase
	plot(t,phasediff2.1_correct,type='l',col="blue", lty=1,ylim=c(-pi,pi))

```

##Convert phase difference to monthhs
```{r wave5, echo=FALSE}
	phase.diff.month <- phasediff2.1_correct * 12/(2*pi)
	plot(t,phase.diff.month,type='l',col="blue", lty=1,ylim=c(-12,12))
```

##Reconstruct series, filtering out noise
```{r wave.recon, echo=FALSE}
 recon1 <- reconstruct.series(cwt1)  #square to get back to original scale
 recon1a <- reconstruct.series(cwt1, sp1=10, sp2=27)^2 #keep periods of 10-27
 recon1b <- reconstruct.series(cwt1, sp1=10, sp2=14)^2 # only keep periods of 10-14
 plot(t,recon1,type='l',col="blue", lty=1)
 points(t,recon1a,type="l",col="red",lty=1)
 points(t,recon1b,type="l",col="green",lty=1)
```

#Compare biennial strength in the 2 series
#Confirms that biennial pattern is stronger in series 1 than series 2
```{r annual.biennial}

biennial.func<-function(ds, period.range.short=c(10,14),period.range.long=c(20,30)){
  annual.indices<-which(ds$periods>=period.range.short[1] & ds$periods<=period.range.short[2])
  biennial.indices<-which(ds$periods>=period.range.long[1] & ds$periods<=period.range.long[2])
  annual.power<-mean(ds$global[annual.indices])
  biennial.power<-mean(ds$global[biennial.indices])
  biennial.annual.ratio<-biennial.power/annual.power
}

series1.biennial.ratio<-biennial.func(ds=cwt1)
series2.biennial.ratio<-biennial.func(ds=cwt2)
series1.biennial.ratio
series2.biennial.ratio
```

##Coherence analysis 
Evaluate coherence between the 2 series. At which periods and which time points are the series in sync?
```{r coherence}
#Coherence analysis
coh1.2 <- complete.coh(wave1,wave2, mc.sim=100, dj=1/10)
plot.coh(coh1.2)
```

### 2023-01-05 RSV_timing_explain/RSV_timing_tutorial.Rmd
add Dan and Ginny's class materials
update timing materials

@@
 - [RSV peak timing at ZIP code level and the drivers of RSV spread](https://doi.org/10.1126/sciadv.abd6421)
+
+## Harmonic regression to estimate the peak timing of RSV epidemics
+
+**Note: Most of the following materials came from Dan and Ginny's Lecture notes abd Harmonic Regression by NCSS^[https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Harmonic_Regression.pdf].**
+
+Please check out:
+
+-  Dan's class: Public Health Surveillance 
+\
+-  Ginny's class: Quantitative Methods in Infectious Diseases
+
+$$X_t=\mu+R\cos(ft+d)+e_t$$
+Where 
+\
+$X_t$ is the time-series contains a periodic (cyclic) component.
+\
+$\mu$ is mean of the series.
+\
+$R$ is the amplitude of seasonality.
+\
+$f=\frac{1}{period}$ is the frequency of the periodic.
+\
+$d$ is the phase or horizontal offset.
+\
+$e_t$ is the random error (noise) of the series.
+\
+$t$ is the time step
+
+## Pseudo-RSV data: Simulate time series with a 12 month period
+Imagine this is RSV case data from 2 states, and we want to investigate the
+epidemic characteristics in these states and the lag between states.
+
+```{r simulate, echo=T}
+set.seed(123)
+  n=120 # 10 years
+  t <- seq(1,n)
+ amp1=2.5 # high amplitude
+ freq=1/12 # frequency = 1/period
+ amp2=0.5 # low amplitude
+
+ xt1a=amp1*cos(2*3.14159*t*freq)
+
+#other series shifted by 3 months
+ xt2a=amp1*cos(2*3.14159*t*freq+1)
+
+#Simulate some poisson count data
+ xt1=rpois(n,exp((1+xt1a)))
+ xt2= rpois(n,exp(2+xt2a))
+```
+
+## The observed pseudo-RSV cases over time in two states
+
+```{r, echo=T}
+ plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",
+      ylab="Cases of RSV", xlab="Month(n)", 
+      ylim=range(c(xt1,xt2)))
+ lines(t,xt2,col="red")
+```

### 2023-01-10 RSV_timing_explain/RSV_timing_tutorial.Rmd
minor change

@@
 
-$$X_t=\mu+R\cos(ft+d)+e_t$$
+$$X_t=\mu+R\cos(2\pi ft+d)+e_t$$
 Where 
@@
 ## Pseudo-RSV data: Simulate time series with a 12 month period
-Imagine this is RSV case data from 2 states, and we want to investigate the
-epidemic characteristics in these states and the lag between states.
+Imagine this is RSV case data from 2 states, and we want to investigate the epidemic characteristics in these states and the lag between states.
 
@@
 
-#other series shifted by 3 months
+#other series shifted by 2 months
  xt2a=amp1*cos(2*3.14159*t*freq+1)
@@
 ```
+## Investigate the epidemic characteristics of the pseudo-RSV time-serieses
+
+This is based on the prior knowledge that RSV has an annual cycle in temperate regions.
+For other viruses or RSV circulation in other climate, you should consider using wavelet analysis to identify the periodicity first.
+
+### Create the needed harmonic variables with 12 month periods
+```{r harm1, echo=TRUE}
+t<-1:120
+#Create harmonic variables
+sin12=sin(2*pi*t/12)
+cos12=cos(2*pi*t/12)
+```
+
+### Fit a simple poisson regression with 12 month period

### 2023-01-11 RSV_timing_explain/RSV_timing_tutorial.Rmd
update

--- 2023-01-05 RSV_timing_explain/RSV_timing_tutorial.Rmd
@@
 freq=1/12 # frequency = 1/period
- amp2=0.5 # low amplitude
+amp2=2 # low amplitude
 
@@
 
-#other series shifted by 3 months
- xt2a=amp1*cos(2*3.14159*t*freq+1)
+#other series shifted by 2 months
+xt2a=amp2*cos(2*3.14159*t*freq+1)
 
@@
 xt1=rpois(n,exp((1+xt1a)))
- xt2= rpois(n,exp(2+xt2a))
+xt2=rpois(n,exp(2+xt2a))
 ```
@@
 
-```{r, echo=T}
+```{r, echo=F}
 plot(t, xt1,type="l", lwd=3, xaxt="n",bty="l",
       ylab="Cases of RSV", xlab="Month(n)", 
-      ylim=range(c(xt1,xt2)))
+      ylim=range(c(xt1,xt2)),
+     xlim=c(0,max(t)+10))
 lines(t,xt2,col="red")
+axis(2, at = seq(0,120,20))
+axis(1, at = seq(0,120,12))
+text(x=max(t)+8, y=xt1[119], labels = "state1",col="black")
+text(x=max(t)+8, y=xt2[119], labels = "state2",col="red")
 ```
+
+## Investigate the epidemic characteristics of the pseudo-RSV time-serieses
+
+This is based on the prior knowledge that RSV has annual cycle in temperate regions and biannual cycle in high latitude regions.
+For other viruses or RSV circulation in other climate, you should consider using wavelet analysis to identify the periodicity first.
+
+```{r harm1, echo=TRUE}
+# Create the needed harmonic variables with 
+# 6 month, 12 month, and 24 month periodicities
+t<-1:120
+#Create harmonic variables
+sin6=sin(2*pi*t/6)
+cos6=cos(2*pi*t/6)
+sin12=sin(2*pi*t/12)
+cos12=cos(2*pi*t/12)
+sin24=sin(2*pi*t/24)
+cos24=cos(2*pi*t/24)
+```
+
+When you write the equations with sin and cos terms, you will need to include them both in an equation.
+For example, if the coefficient of cos12 is significant, you will need to include sin12 even though the coefficient is not significant in the summary.
+
+## Fit a simple poisson regression with 12 month period for state 1
+
+```{r harm_reg1}
+fit1a <- glm(xt1~sin12+cos12, family='poisson')
+pred1a<- fitted(fit1a)
+#summary(fit1a)
+```
+```{r ,echo=FALSE}
+plot (t, xt1,bty="l", xaxt="n",yaxt="n",col="blue",
+      xlab="Month(n)",ylab="Cases of RSV", type="p",
+      ylim=range(c(xt1,pred1a)),
+     xlim=c(0,max(t)+25),main="State 1")
+lines(t,pred1a, lty=1, lwd=3)
+axis(2, at = seq(0,120,20))
+axis(1, at = seq(0,120,12))
+text(x=max(t)+15, y=pred1a[120], labels = "fitted",col="black")
+text(x=max(t)+15, y=xt1[120], labels = "observed",col="blue")
+```
+
+## Add in 24 month periodicity
+
+\scriptsize
+```{r harm_reg2}
+fit2a <- glm(xt1~sin12+cos12+sin24+cos24,family='poisson')
+pred2a<- fitted(fit2a)
+summary(fit2a)
+```
+
+## Add in 6 month periodicity
+
+\tiny
+```{r harm_reg3}
+fit3a<-glm(xt1~sin12+cos12+sin24+cos24+sin6+cos6,family='poisson' )
+pred3a<-fitted(fit3a)
+summary(fit3a)
+```
+
+## Determine best model with AIC (smaller=better)
+Winner is Model 1 (12 period). Also, when models have similar AIC score (within 2 points), we prefer a simpler model.
+```{r aic}
+ AIC(fit1a)
+ AIC(fit2a)
+ AIC(fit3a)
+```
+
+## Fit a simple poisson regression with 12 month period for state 2
+
+```{r harm_reg4}
+fit1b <- glm(xt2~sin12+cos12, family='poisson')
+pred1b<- fitted(fit1b)
+#summary(fit1a)
+```
+
+```{r, echo=FALSE}
+plot (t,xt2,bty="l", xaxt="n",yaxt="n",col="blue",
+      xlab="Month(n)",ylab="Cases of RSV", type="p",
+      pch = 20,
+      ylim=range(c(xt2,pred1b)),main="State 2")
+lines(t,pred1b, lty=1, lwd=1,col="red")
+axis(2, at = seq(0,120,20))
+axis(1, at = seq(0,120,12))
+```
+
+## Calculate amplitudes of the 12 periods in state1 and state2
+```{r amp}
+beta_sin12_xt1<-coef(fit2a)['sin12']
+beta_cos12_xt1<-coef(fit2a)['cos12']
+
+amp12_xt1<-sqrt(beta_sin12_xt1^2+
+                  beta_cos12_xt1^2)
+
+amp12_xt1 #True value=2.5
+
+beta_sin12_xt2<-coef(fit1b)['sin12']
+beta_cos12_xt2<-coef(fit1b)['cos12']
+
+amp12_xt2<-sqrt(beta_sin12_xt2^2+
+                  beta_cos12_xt2^2)
+
+amp12_xt2 #True value=2
+```

### 2023-01-11 RSV_timing_explain/RSV_timing_tutorial.Rmd
add peak timing calculation

@@
 
-$$X_t=\mu+R\cos(2\pi ft+d)+e_t$$
+$$X_t=\mu+R\cos(2\pi ft-d)+e_t$$
 Where 
@@
 #other series shifted by 2 months
-xt2a=amp2*cos(2*3.14159*t*freq+1)
+xt2a=amp2*cos(2*3.14159*t*freq-1)
 
@@
 ```
+
+## Calculate phase of the 12 month period in state1 and state2
+
+```{r}
+# Phase angle
+phase12_xt1 <- -atan(beta_sin12_xt1/beta_cos12_xt1)
+phase12_xt1
+# True value = 0
+phase12_xt2 <- -atan(beta_sin12_xt2/beta_cos12_xt2)
+phase12_xt2
+# True value = 1
+```
+
+## Calculate peak timing in month in state1 and state2
+
+```{r}
+# Average peak timing
+12*(1-phase12_xt1/(2*pi)) # True value = 0 month 
+12*(1-phase12_xt2/(2*pi)) # True value = 2 month
+```
+**Note: the period is 12 month** 
+- Therefore, peak timing of 12.06 month (year 1) equal to peak timing of 0.06 month (year 0)
+\
+\
+- peak timing of 13.93 month (year 1) equals to 1.93 month (year 0)

### 2023-01-11 RSV_timing_explain/RSV_timing_tutorial.Rmd
add second derivatives

@@
 - [RSV peak timing at ZIP code level and the drivers of RSV spread](https://doi.org/10.1126/sciadv.abd6421)
+\
+\
+- [Assessment and optimization of respiratory syncytial virus prophylaxis in Connecticut, 1996–2013](https://www.nature.com/articles/s41598-021-90107-8)
+\
+\
+- [Disease outbreak outcome estimation using penalized splines](https://cran.r-project.org/web/packages/pspline.inference/vignettes/seasonal.html)
 
@@
 - peak timing of 13.93 month (year 1) equals to 1.93 month (year 0)
+
+## Identify the onset of an epidemic using the second derivative method
+
+When we look at the pseudo-RSV cases in the last 12 months, it is hard to determine when is the onset by eyes.
+
+```{r, echo=F}
+plot(1:12, tail(xt1,12),type="l", lwd=3, xaxt="n",bty="l",
+      ylab="Cases of RSV", xlab="Month(n)", 
+      ylim=range(c(xt1,xt2)),
+     xlim=c(0,15))
+lines(1:12,tail(xt2,12),col="red")
+axis(2, at = seq(0,120,20))
+axis(1, at = seq(0,12,1))
+text(x=14, y=xt1[119], labels = "state1",col="black")
+text(x=14, y=xt2[119], labels = "state2",col="red")
+```
+
+## Fit a regression to the observed RSV cases
+
+Because now we only observe part of the epidemic, we cannot use the harmonic regression to fit what we observe.
+Instead, we use p-spline regression, a generalized additive model.
+Please check out [Smooth terms in GAM](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/smooth.terms.html), [P-splines in GAMs](https://stat.ethz.ch/R-manual/R-devel/library/mgcv/html/smooth.construct.ps.smooth.spec.html) and Iris's work on [p-spline inference](https://cran.r-project.org/web/packages/pspline.inference/vignettes/seasonal.html) for introduction.
+
+```{r}
+require(mgcv)
+irregular_model <- gam(cases ~ s(x=time, k=5, bs="ps"),
+                       family=poisson, method="REML", data=data.frame(time=seq(1:12), cases=tail(xt1,12))) 
+# k is knots
+# The value for k set the upper limit on the 
+# wiggliness of the smooth function
+# You can choose the value based on AIC scores as well 
+```
+
+## Comparing the model fit and what we observed
+The dots are what we observed and the line is the model fit
+```{r,fig.dim=c(4, 4)}
+plot(1:12, tail(xt1,12),type="p", xaxt="n",bty="l",
+      ylab="Cases of RSV", xlab="Month(n)", 
+      ylim=range(c(xt1,xt2)),
+     xlim=c(0,12))
+lines(1:12,fitted(irregular_model))
+```
+
+## Function to calculate the first and second derivatives
+```{r}
+deriv <- function(x, y) diff(y) / diff(x) # function to calculate derivative
+middle_pts <- function(x) x[-1] - diff(x) / 2 # function to assist derivative calculation
+```
+
+```{r}
+#install.packages("pspline.inference")
+library(pspline.inference)
+t <- seq(0.5,12,0.01)
+dtime <- seq(0.5,12,0.01)
+
+# we use Iris's package to get the uncertainty interval
+# of disease trajectory. 
+data=pspline.sample.timeseries(irregular_model, data.frame(time=t), pspline.outbreak.cases, samples=1)
+
+deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),time=c(1:length(diff(t)))) # calculate the first derivative
+```

### 2023-01-12 RSV_timing_explain/RSV_timing_tutorial.Rmd
add second derivative for onset estimates

@@
 ## Function to calculate the first and second derivatives
+
 ```{r}
-deriv <- function(x, y) diff(y) / diff(x) # function to calculate derivative
-middle_pts <- function(x) x[-1] - diff(x) / 2 # function to assist derivative calculation
+deriv <- function(x, y) diff(y) / diff(x) 
+# function to calculate derivative
+
+middle_pts <- function(x) x[-1] - diff(x) / 2 
+# function to assist derivative calculation
 ```
 
-```{r}
+```{r, echo=T,message=FALSE}
 #install.packages("pspline.inference")
+# require did not work for this package
 library(pspline.inference)
+require(dplyr)
+
 t <- seq(0.5,12,0.01)
 dtime <- seq(0.5,12,0.01)
+```
 
-# we use Iris's package to get the uncertainty interval
-# of disease trajectory. 
-data=pspline.sample.timeseries(irregular_model, data.frame(time=t), pspline.outbreak.cases, samples=1)
+## Find uncertainty interval of disease trajectory and onset timing
 
-deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),time=c(1:length(diff(t)))) # calculate the first derivative
+\scriptsize
+```{r}
+# we use Iris's package to get the uncertainty interval of disease trajectory. 
+cases.samples=pspline.sample.timeseries(
+  irregular_model, data.frame(time=t), 
+  pspline.outbreak.cases, samples=150)
+
+onset.samples = cases.samples %>% 
+  group_by(pspline.sample) %>% # for each sample do the following
+  do((function(data){
+    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
+                            time=c(1:length(diff(t))))
+    # calculate the first derivative
+
+    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
+      time=c(1:(length(diff(t))-1))) # calculate the second derivative
+
+    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
+    # only look at second derivatives in the increasing segment (first derivative>0 )
+    second_d_test <- second_d[second_d$time%in%indicator$time,]
+
+    onset = dtime[second_d_test$time[second_d_test$second.deriv==
+        max(second_d_test$second.deriv)]] 
+# find when the second derivative of  the smooth functions reached its maximum 
+    data.frame(
+      onset = onset,
+      cases = data$cases[which(data$time==onset)]) 
+    # find the number of RSV 
+    # when the second derivative reach its maximum
+  })(.))
 ```
+
+## Plot the range of the onset on top of the epidemic curves
+
+The dots are what we observed and the grey lines are the fitted epidemic curves.
+The blue area are the 95% confidence interval of epidemic onsets and the vertical dotted line shows the mean of the epidemic onset estimate.
+```{r, message=FALSE,echo=FALSE}
+require(ggplot2)
+# observation data
+Epi_data=data.frame(cases=tail(xt1,12),
+                    time=1:12)
+
+names(cases.samples)[3]="cases"
+# change the the default column name
+
+ggplot() + 
+  geom_line(data=cases.samples,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
+  geom_point(data=Epi_data,aes(x=time,y=cases)) + # plot observed case number
+geom_vline(xintercept=median(onset.samples$onset),
+           linetype="dotted",size=0.2)+ # plot median of the epidemic onset estimate
+  annotate("rect", xmin = quantile(onset.samples$onset,0.025),
+           xmax = quantile(onset.samples$onset,0.975), 
+           ymin = 0, ymax = Inf,
+        alpha = .2,fill="blue")+ # show the 95% confidence interval of epidemic onsets
+  theme_classic()
+```
+
+## Plot the onset estimates of state 2.
+
+The estimates show bimodal distribution because we cut through the epidemic season. To correctly analyze the data, we need to limit the time to after the peak of previous outbreak.
+
+```{r, message=FALSE,echo=FALSE}
+irregular_model_2 <- gam(cases ~ s(x=time, k=5, bs="ps"),
+                       family=poisson, method="REML", data=data.frame(time=seq(1:12), cases=tail(xt2,12)))
+
+cases.samples_2=pspline.sample.timeseries(
+  irregular_model_2, data.frame(time=t), 
+  pspline.outbreak.cases, samples=150)
+
+onset.samples_2 = cases.samples_2 %>% 
+  group_by(pspline.sample) %>% # for each sample do the following
+  do((function(data){
+    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
+                            time=c(1:length(diff(t))))
+    # calculate the first derivative
+
+    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
+      time=c(1:(length(diff(t))-1))) # calculate the second derivative
+
+    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
+    # only look at second derivatives in the increasing segment (first derivative>0 )
+    second_d_test <- second_d[second_d$time%in%indicator$time,]
+
+    onset = dtime[second_d_test$time[second_d_test$second.deriv==
+        max(second_d_test$second.deriv)]] 
+# find when the second derivative of  the smooth functions reached its maximum 
+    data.frame(
+      onset = onset,
+      cases = data$cases[which(data$time==onset)]) 
+    # find the number of RSV 
+    # when the second derivative reach its maximum
+  })(.))
+
+# observation data
+Epi_data_2=data.frame(cases=tail(xt2,12),
+                    time=1:12)
+
+names(cases.samples_2)[3]="cases"
+# change the the default column name
+
+ggplot() + 
+  geom_line(data=cases.samples_2,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
+  geom_point(data=Epi_data_2,aes(x=time,y=cases)) + # plot observed case number
+geom_vline(xintercept=median(onset.samples_2$onset),
+           linetype="dotted",size=0.2)+ # plot mean of the epidemic onset estimate
+  geom_vline(xintercept=onset.samples_2$onset,
+           linetype="solid",size=0.02)+
+  annotate("rect", xmin = quantile(onset.samples_2$onset,0.1),
+           xmax = quantile(onset.samples_2$onset,0.9), 
+           ymin = 0, ymax = Inf,
+        alpha = .2,fill="blue")+ # show the 80% confidence interval of epidemic onsets
+  theme_classic()
+```
+
+## Retry the analysis for the onset estimates of state 2.
+
+This analysis will work better with weekly data and after the epidemic take off as you will have more data points.
+
+```{r, message=FALSE,echo=FALSE}
+irregular_model_2 <- gam(cases ~ s(x=time, k=5, bs="ps"),
+                       family=poisson, method="REML", data=data.frame(time=seq(1:11), cases=tail(xt2,11)))
+
+t <- seq(0.5,11,0.01)
+dtime <- seq(0.5,11,0.01)
+
+cases.samples_2=pspline.sample.timeseries(
+  irregular_model_2, data.frame(time=t), 
+  pspline.outbreak.cases, samples=150)
+
+onset.samples_2 = cases.samples_2 %>% 
+  group_by(pspline.sample) %>% # for each sample do the following
+  do((function(data){
+    deriv.pred = data.frame(deriv=diff(data$cases)/diff(data$time),
+                            time=c(1:length(diff(t))))
+    # calculate the first derivative
+
+    second_d <- data.frame(second.deriv=deriv(middle_pts(dtime),deriv(dtime,data$cases)),
+      time=c(1:(length(diff(t))-1))) # calculate the second derivative
+
+    indicator = deriv.pred[which(deriv.pred$deriv>0),] 
+    # only look at second derivatives in the increasing segment (first derivative>0 )
+    second_d_test <- second_d[second_d$time%in%indicator$time,]
+
+    onset = dtime[second_d_test$time[second_d_test$second.deriv==
+        max(second_d_test$second.deriv)]] 
+# find when the second derivative of  the smooth functions reached its maximum 
+    data.frame(
+      onset = onset,
+      cases = data$cases[which(data$time==onset)]) 
+    # find the number of RSV 
+    # when the second derivative reach its maximum
+  })(.))
+
+# observation data
+Epi_data_2=data.frame(cases=tail(xt2,11),
+                    time=1:11)
+
+names(cases.samples_2)[3]="cases"
+# change the the default column name
+
+ggplot() + 
+  geom_line(data=cases.samples_2,aes(x=time,y=cases,group = pspline.sample),alpha = .2,color="grey")+ # plot fitted epidemic curves
+  geom_point(data=Epi_data_2,aes(x=time,y=cases)) + # plot observed case number
+geom_vline(xintercept=median(onset.samples_2$onset),
+           linetype="dotted",size=0.2)+ # plot median of the epidemic onset estimate
+  annotate("rect", xmin = quantile(onset.samples_2$onset,0.025),
+           xmax = quantile(onset.samples_2$onset,0.975), 
+           ymin = 0, ymax = Inf,
+        alpha = .2,fill="blue")+ 
+  theme_classic()
+```

### 2023-01-13 RSV_timing_explain/RSV_timing_tutorial.Rmd
include theoratical background knowledge

@@
 - In section 1, we will first introduce how to find the peak timing of RSV epidemics using harmonic regression (given regular annual/biennial seasonality).
-We will then talk about identifying the onset of RSV epidemics using second derivative method, regardless of the seasonality of RSV.
+We will learn how to use R to identify peak timing of periodic RSV epidemics.
 \
 \
-- In section 2, we will learn how to use R to identify peak timing and onset of RSV epidemics.
+- In section 2, we will then talk about identifying the onset of RSV epidemics using second derivative method, regardless of the seasonality of RSV.  
 \
@@
 
-## Harmonic regression to estimate the peak timing of RSV epidemics
+## Section 1: Harmonic regression to estimate the peak timing of RSV epidemics
 
-**Note: Most of the following materials came from Dan and Ginny's Lecture notes abd Harmonic Regression by NCSS^[https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Harmonic_Regression.pdf].**
+\small
+
+**Note: Most of the following materials came from Dan and Ginny's Lecture notes and Harmonic Regression by NCSS^[https://www.ncss.com/wp-content/themes/ncss/pdf/Procedures/NCSS/Harmonic_Regression.pdf].**
 
@@
 
-## Identify the onset of an epidemic using the second derivative method
+## Section 2: Identify the onset of an epidemic using the second derivative method
 
@@
 ```
+
+## So what are the methods to identify epidemic onsets?
+
+The most common way is fixed-value thresholds based on the number of cases defined by researchers.
+For example, in Dr.
+Rachel Baker's paper^[Baker, R.E., Mahmud, A.S., Wagner, C.E.
+et al.
+Epidemic dynamics of respiratory syncytial virus in current and future climates.
+Nat Commun 10, 5512 (2019).], they defined onset as when the normalized mean RSV incidence per week exceeds 0.2.
+However, this type of threshold is based on the assumption that the threshold "...
+will be low enough to constitute onset but high enough to exceed random fluctuations in the data...".
+It generally works well but it requires extensive knowledge of the epidemic curves in different locations to define this value.
+Therefore, we introduce a new method, second derivative, to estimate the onset of an outbreak.
+
+## Theoretical background on first and second derivative
+
+::: columns
+
+:::: {.column width="25%"}
+
+![Observed curve, first derivative, and second derivative](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/RSV_timing_explain/Representative-smoothed-reflectance-first-derivative-and-second-derivative-data.png)
 

### 2023-01-13 RSV_timing_explain/RSV_timing_tutorial.Rmd
update

@@
 
-:::: {.column width="25%"}
+:::: {.column width="35%"}
 
 ![Observed curve, first derivative, and second derivative](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/RSV_timing_explain/Representative-smoothed-reflectance-first-derivative-and-second-derivative-data.png)
+
+::::
+
+:::: {.column width="5%"}
+
+::::
+
+:::: {.column width="60%"}
+
+ \small
+
+The figure shows the relationship among the observed curve, the first derivative of the observed curve, and the second derivative of the observe curve."Peak in the first derivative curve identify the location of the red edge.
+Peak in the second derivative identify the edge of the red edge."^[\tiny Randy M Hamilton et al.
+(2009).
+Pre-visible Detection of Grub Feeding in Turfgrass using Remote Sensing.
+Photogrammetric Engineering & Remote Sensing.
+75.
+179-192.
+10.14358/PERS.75.2.179.]
+\
+\
+When the first derivative is positive, the original curve is increasing.
+In epidemic, this corresponds to the early stage of an outbreak as cases are increasing.
+When the second derivative reach its maximum in the segment that the first derivative is positive, it means that the growth rate of the increasing trend reach its maximum.
+This fits for the definition of the starting point of an disease outbreak.
+
+::::
+
+:::
 
@@
 
-```{r}
+```{r,message=FALSE}
 require(mgcv)
 irregular_model <- gam(cases ~ s(x=time, k=5, bs="ps"),
-                       family=poisson, method="REML", data=data.frame(time=seq(1:12), cases=tail(xt1,12))) 
+                       family=poisson, method="REML",
+                       data=data.frame(time=seq(1:12),
+                                  cases=tail(xt1,12))) 
 # k is knots

### 2023-01-13 HCUP_data_clean/Data directionary.Rmd
instruction on HCUP data

---
title: "HCUP data"
author: "ZHE ZHENG"
date: "2023-01-13"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## File directory and its contents

The parent directory is named **hcup-sid**. Inside of the direcotry you will find:

  - \textcolor{red}{HCUPCode}. It contains the sas codes to read in the asc files from HCUP data.
  \
  \
  - \textcolor{red}{HCUPData}. It contains the original asc data files from the zip files you decompress from CD.
  \
  \
  - \textcolor{red}{ParquetData}. This is the final data files you will use for analysis. Each state has two files. One file contains the core data of all years, including diagnoses, date of birth, date of admission etc. The other file named chgs contains the charges information (useful for cost-effectiveness studies). 
  \
  \
  - \textcolor{red}{SASCode}. These are the codes that Iris and I (Gigi) wrote to convert all the asc files to csv files.
  \
  \
  - \textcolor{red}{SASData}. This contains the output csv files from the SAS program.

## Process to clean the HCUP data

(1) Read in CD. You need to ensure your computer have at least 200G space for storing the data.
  \
  \
(2) Uncompressed the zip files by inputting the passwords for each year and each state.
  \
  \
(3) Copy the asc files to directory \textcolor{red}{HCUPData}.
  \
  \
(4) Run the *sid time series.sas* in \textcolor{red}{SASCode} to run the SAS codes in \textcolor{red}{HCUPCode}. This will load all the asc data into SAS environment. And then run the *ConvertCSV.sas* to convert all data in SAS environment into csv files.
  \
  \
(5) Download R file: https://github.com/Gigi112/MedicareCMSExploration/blob/main/2.CreateParquet_states.Rmd and modify it to convert all csv files to Parquet files.
  \
  \
(6) Modify R file: https://github.com/VA-CareDisruptions/MedicareCMSExploration/blob/main/1.%20exploratory%20analyses.Rmd to analyze parquet data in R environment.

### 2023-01-16 Jags_Bayesian_model.Rmd
add jags tutorials

---
title: "Bayesian model in JAGS"
subtitle: "Tutorial 3 for transition"
author: "Zhe Zheng (Gigi)"
date: "2023-01-16"
fontsize: 10pt
output: 
  beamer_presentation:
  theme: CambridgeUS
  colortheme: "beaver"
  fonttheme: "structurebold"
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(rjags)
#"Note that the rjags package does not include a copy of  
# the JAGS you must install this separately.  
# For instructions on downloading JAGS,
# see the home page at https://mcmc-jags.sourceforge.io."
require(mcmcplots) # to plot the trace plots
require(coda) # for gelman diagnostic
```

## Outline

In this tutorial, we will learn how to use JAGS (Just Another Gibbs Sampler) in R to estimate unknown parameters with Markov Chain Monte Carlo (MCMC) methods in Bayesian framework. If you are interested in learning more about the theoretical knowledge of MCMC and Bayesian modeling, please check out courses "Probability and Statistics" (taught by Prof. Joseph Chang) and "Bayesian Statistics" (taught by Prof. Josh Warren). 
\
\
In the first section, we will introduce a simple jags model to help understand the syntax for rjags package.
\
\
In the second section, we will learn how to build up complicated models step by step (censor data and spatial autocorrelation). Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing]( https://doi.org/10.1111/irv.12965)

## Example 1: a simple jags model to estimate the unknown parameters

### First: Simulate a dataset for analysis

In this example, I simulate my own data in R in order to compare the MCMC results with our input parameters. I create a continuous outcome variable $y$ as a function of one predictor $x$ and an error term $\epsilon$ . I simulate 100 observations. The linear regression looks like
$$y=\beta_0+\beta_1x+\epsilon$$
**Our goal is to estimate the unknown parameters** $\boldsymbol{\beta_0}$ **and** $\boldsymbol{\beta_1}$ **given the observations and the linear relationship**

```{r}
n.sim=100; set.seed(123)
x=rnorm(n.sim, mean = 5, sd = 3)
epsilon=rnorm(n.sim, mean = 0, sd = 1)
beta0=5
beta1=3.2
y=beta0 + beta1 * x + epsilon
```

## Visualize the observations

Question: what is the intercept and slope for this linear regression?
```{r}
plot(x,y,type="p")
```

## rjags mamual

The following is quoted from rjags document, authored by Martyn Plummer:

"JAGS is a clone of BUGS (Bayesian analysis Using Gibbs Sampling). See Lunn et al (2009) for
a history of the BUGS project. Note that the rjags package does not include a copy of the JAGS
library: you must install this separately. For instructions on downloading JAGS, see the home page at https://mcmc-jags.sourceforge.io.
\
\
To fully understand how JAGS works, you need to read the JAGS User Manual. The manual explains the basics of modelling with JAGS and shows the functions and distributions available in the
dialect of the BUGS language used by JAGS. It also describes the command line interface. The
rjags package does not use the command line interface but provides equivalent functionality using
R functions."

## rjags mamual (continued)
\
\
"Analysis using the rjags package proceeds in steps:
1. Define the model using the BUGS language in a separate file.
\
\
2. Read in the model file using the jags.model function. This creates an object of class “jags”.
\
\
3. Update the model using the update method for “jags” objects. This constitutes a ‘burn-in’
period.
\
\
4. Extract samples from the model object using the coda.samples function. This creates an object of class “mcmc.list” which can be used to summarize the posterior distribution. The coda
package also provides convergence diagnostics to check that the output is valid for analysis
(see Plummer et al 2006)."

## Jags model structure (Step 1)

```{r}
basic_mod <-  "model{
 #model
  for(i in 1:n.sim){
   y[i] ~ dnorm(mu[i], tau)
   mu[i] = beta0 + beta1 * x[i]
  }

 #priors
 beta0 ~ dnorm(0, 0.01)
 beta1 ~ dnorm(0, 0.01)
 tau ~ dgamma(0.01,0.01)
 }
 "
```

## Initialize model with known data and initial guess of unknown parameters (Step 2)

```{r}
######## Define known data ############
datalist=list("y"=y,"x"=x,"n.sim"=n.sim)

######## Set initial values for parameter estimate ######## 
inits=function(){list("beta0"=rnorm(1), "beta1"=rnorm(1))}
```

## Read in the model file using the jags.model function (Step 3)

```{r}
### Read in the model file using the jags.model function ##
jags.basic.mod = jags.model(textConnection(basic_mod),
                            data = datalist,
                            inits = inits,
                            n.chains = 4)
```

## Discard burn-in period and sample for posterior distribution (Step 4)

```{r}
############ burn-in period ############
update(jags.basic.mod,5000) # n.burnin = 1000

##### sample for posterior distribution #######
model_sample <- coda.samples(jags.basic.mod,
                             c("beta0","beta1"),
                             100000,thin=5)
#You can also use jags.samples(). The results will be the same
```

## Summarize the posterior distribution 
```{r}
summary(model_sample)
```

## Check for model convergence (gelman diagnotics)

```{r}
gelman.diag(model_sample) ## <=1.1 means well converge 
```

## Check for model convergence (gelman plot)

```{r}
gelman.plot(model_sample)
```

## Plot the trace plot

```{r}
par(mfrow=c(1,2))
traceplot(model_sample)
```

## Compare the posterior density and real value for beta0

```{r}
denplot(model_sample, parms = c("beta0"))
abline(v=5)  # True value
```

## Compare the posterior density and real value for beta1

```{r}
denplot(model_sample, parms = c("beta1"))
abline(v=3.2) # True value
```

### 2023-01-16 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
comment

--- 2023-01-16 Jags_Bayesian_model.Rmd
@@
    y[i] ~ dnorm(mu[i], tau)
+# tau is the error term
+
    mu[i] = beta0 + beta1 * x[i]
+# beta0 and beta1 are the unknown parameters
   }
 
- #priors
+ #priors; these are weakly informative priors
  beta0 ~ dnorm(0, 0.01)
+# dnorm(mean, precision)
+# precision = 0.01 --> var = 100
  beta1 ~ dnorm(0, 0.01)
  tau ~ dgamma(0.01,0.01)
+# dgamma(shape,rate)
  }
@@
 
-```{r}
+```{r,message=TRUE}
 ############ burn-in period ############

### 2023-01-17 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
update jags

@@
 In the second section, we will learn how to build up complicated models step by step (censor data and spatial autocorrelation).
-Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing]( https://doi.org/10.1111/irv.12965)
+Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing](https://doi.org/10.1111/irv.12965)
 
@@
 ```
+
+## Section 2: estimate the parameters for the relative RSV timing
+
+After we estimated the [RSV timing](https://doi.org/10.1111/irv.12965) using second derivative method (please check: RSV_timing_explain), we would like to know if some potential drivers contribute to the relative timing difference in RSV epidemic onsets before and during COVID-19 pandemic.
+
+The model is given as: 
+$$Y_i=\alpha_{0i}+\beta_1\overline{Z_i}+\epsilon_i$$
+$$\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$$
+
+## Meaning of the variables and parameters
+
+where $Y_i$ is the relative timing of RSV epidemic onset with respect to Florida in 2021 in state $i$ (measured in weeks), and $\overline{Z_i}$ is the average relative timing of RSV seasons in state $i$ compared to Florida from 2016-2019.
+The global intercept parameter $\beta_0$ and slope parameter $\beta_1$ describe the overall similarity between a typical year’s relative onset timing across all states and the relative timings observed in 2021.
+The observation-level random effect $\epsilon_i  ~ N(0,\sigma^2)$ accounts for residual variability in the data.
+\
+\
+The state-specific intercepts, $\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$, represent the state-level similarity in relative timing for 2021 compared to previous seasons, where $x_i$ is the vector of covariates that could potentially explain differences in the relative timing of RSV between 2021 and previous years.
+The covariates include population density, average household size, and stringency index of non-pharmaceutical interventions against COVID-19.
+\
+\
+$\theta_i$ represents a spatially correlated random effect, which we will learn more in detail later.
+
+## Corresponding model structure without considering spatial autocorrelation
+
+```{r}
+Nospatial_model <- "model{
+for(i in 1:n){
+Z[i] ~ dnorm(mu[i], inv_var)
+ mu[i] <- (beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i])+(beta1+x1[i]*theta1+x2[i]*theta2+theta3*x3[i])*log_W[i]
+}
+
+beta0~dnorm(0,0.0001)
+beta1~dnorm(0,0.0001)
+lambda1~dnorm(0,0.0001)
+lambda2~dnorm(0,0.0001)
+lambda3~dnorm(0,0.0001)
+theta2~dnorm(0,0.0001)
+theta1~dnorm(0,0.0001)
+theta3~dnorm(0,0.0001)
+
+inv_var ~ dgamma(0.01, 0.01)
+}
+"
+```

### 2023-01-17 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
add basic timing model

@@
 
-## Corresponding model structure without considering spatial autocorrelation
+## Jags model code without considering spatial autocorrelation
 
@@
 Nospatial_model <- "model{
+################ model ####################################
 for(i in 1:n){
-Z[i] ~ dnorm(mu[i], inv_var)
- mu[i] <- (beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i])+(beta1+x1[i]*theta1+x2[i]*theta2+theta3*x3[i])*log_W[i]
+Y[i] ~ dnorm(mu[i], inv_var)
+ mu[i] <- alpha_0[i] + beta1*log_Z[i]
+ # log_Z is the log transform average relative timing of RSV
+
+alpha_0[i]= beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i]
+# this is alpha_0i = beta0 + X_i*lambda (coviariates)
 }
 
+################ priors ####################################
 beta0~dnorm(0,0.0001)
@@
 lambda3~dnorm(0,0.0001)
-theta2~dnorm(0,0.0001)
-theta1~dnorm(0,0.0001)
-theta3~dnorm(0,0.0001)
 
-inv_var ~ dgamma(0.01, 0.01)
-}
-"
+inv_var ~ dgamma(0.01, 0.01)}"
 ```
+
+## Load the data and run the jags model to sample posteriors
+
+```{r,eval=FALSE}
+## loading data from the working directory
+data <- readRDS("./Nospatial_model.rds")
+
+### Read in the model file using the jags.model function ##
+non_spatial <- jags.model(textConnection(Nospatial_model),
+                     data=data,n.chains = 4)
+
+############ burn-in period ############
+update(non_spatial,10000)
+
+######## sample for posterior distribution ##########
+non_spatial_sample <- coda.samples(non_spatial,c("beta0","beta1",
+                            "lambda1","lambda2",
+                            "lambda3","inv_var"),
+                             100000,thin=20)
+```
+
+## Summarize the posterior distribution
+
+```{r,eval=FALSE}
+summary(non_spatial_sample)
+```
+
+## Check for model convergence (gelman diagnostic)
+
+```{r,eval=FALSE}
+gelman.diag(non_spatial_sample)
+```
+
+## Trace plots
+
+```{r,eval=FALSE}
+par(mfrow=c(3,2))
+traceplot(non_spatial_sample)
+```
+
+## Density plot of the posterior distributions
+
+```{r,eval=FALSE}
+par(mfrow=c(3,2))
+densplot(non_spatial_sample)
+```

### 2023-01-18 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
add more materials

--- 2023-01-17 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
@@
 \
-In the second section, we will learn how to build up complicated models step by step (censor data and spatial autocorrelation).
+In the second section, we will learn how to build up complicated models step by step (censor data).
 Before reading this section, you will need to familiar yourself with the RSV_timing_explain tutorials and a research article on the [relative RSV timing](https://doi.org/10.1111/irv.12965)
+\
+\
+In the third section, we will introduce the concept of spatial autocorrelation and how to incorporate this in jags model.
 
@@
 After we estimated the [RSV timing](https://doi.org/10.1111/irv.12965) using second derivative method (please check: RSV_timing_explain), we would like to know if some potential drivers contribute to the relative timing difference in RSV epidemic onsets before and during COVID-19 pandemic.
+(In reality, we manually coded the [spatial autocorrelated model](https://github.com/Gigi112/RSVtiming/blob/main/spatialstatis_best.R).
+As this requires taking Josh's classes, I will not go over it in the introduction tutorials.)
 
 The model is given as: 
-$$Y_i=\alpha_{0i}+\beta_1\overline{Z_i}+\epsilon_i$$
+$$Y_i=\alpha_{0i}+\beta_1\overline{W_i}+\epsilon_i$$
 $$\alpha_{0i}=\beta_0+\boldsymbol{x_i^T \lambda}+\theta_i$$
@@
 
-where $Y_i$ is the relative timing of RSV epidemic onset with respect to Florida in 2021 in state $i$ (measured in weeks), and $\overline{Z_i}$ is the average relative timing of RSV seasons in state $i$ compared to Florida from 2016-2019.
+where $Y_i$ is the relative timing of RSV epidemic onset with respect to Florida in 2021 in state $i$ (measured in weeks), and $\overline{w_i}$ is the average relative timing of RSV seasons in state $i$ compared to Florida from 2016-2019.
 The global intercept parameter $\beta_0$ and slope parameter $\beta_1$ describe the overall similarity between a typical year’s relative onset timing across all states and the relative timings observed in 2021.
@@
 Nospatial_model <- "model{
+################ model ####################################
 for(i in 1:n){
-Z[i] ~ dnorm(mu[i], inv_var)
- mu[i] <- (beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i])+(beta1+x1[i]*theta1+x2[i]*theta2+theta3*x3[i])*log_W[i]
+Y[i] ~ dnorm(mu[i], inv_var)
+# for the observed timing (we log transform Y already)
+ mu[i] <- alpha_0[i] + beta1*log_W[i]
+ # log_W is the log transform average relative timing of RSV
+
+alpha_0[i]= beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i]
+# this is alpha_0i = beta0 + X_i*lambda (coviariates)
 }
 
+################ priors ####################################
 beta0~dnorm(0,0.0001)
@@
 lambda3~dnorm(0,0.0001)
-theta2~dnorm(0,0.0001)
-theta1~dnorm(0,0.0001)
-theta3~dnorm(0,0.0001)
 
-inv_var ~ dgamma(0.01, 0.01)
+inv_var ~ dgamma(0.01, 0.01)}"
+```
+
+## Load the data and run the jags model to sample posteriors
+
+```{r, message=FALSE}
+## loading data from the working directory
+data_nospatial <- readRDS("./Nospatial_data.rds")
+
+### Read in the model file using the jags.model function ##
+non_spatial <- jags.model(textConnection(Nospatial_model),
+                     data=data_nospatial,n.chains = 4)
+```
+
+```{r}
+############ burn-in period ############
+update(non_spatial,10000)
+
+######## sample for posterior distribution ##########
+non_spatial_sample <- coda.samples(non_spatial,
+                            c("beta0","beta1",
+                            "lambda1","lambda2",
+                            "lambda3","inv_var"),
+                             100000,thin=20)
+```
+
+## Summarize the posterior distribution
+
+```{r}
+summary(non_spatial_sample)
+```
+
+## Check for model convergence (gelman diagnostic)
+**inv_var did not converge well**.
+This indicates that we should either prolong the iterations or re-parameterize the model.
+Using gelman.plot (next slides), we could tell that prolong iterations may help convergence.
+```{r}
+gelman.diag(non_spatial_sample)
+```
+## Gelman plot
+
+```{r}
+par(mfrow=c(3,2))
+gelman.plot(non_spatial_sample)
+```
+
+## Trace plots
+**note: inv_var is precision = the inverse of variance** 
+```{r}
+par(mfrow=c(3,2))
+traceplot(non_spatial_sample)
+```
+
+## Density plot of the posterior distributions
+
+```{r}
+par(mfrow=c(3,2))
+densplot(non_spatial_sample)
+```
+
+## Section 2.1: right censored survival analysis
+
+In the early phase of an epidemic, we may not yet observe the onset point.
+For example, if you consider an onset threshold and the cases are still below the threshold, the epidemic has not taken off.
+In this case, we need to use right censored survival analysis.
+\
+\
+Click to know more about [right censored survival analysis](https://tinyheero.github.io/2016/05/12/survival-analysis.html)
+
+## Model structure for right censored survival analysis
+
+```{r, eval=FALSE}
+censormodel <- "model{
+for(i in 1:n){
+is.censor[i] ~ dinterval(Z[i],t.cen)
+# whether the data is censor and the censoring time
+
+Z[i] ~ dlnorm(mu[i], inv_var)
+# for the observed timing (we did not log transform Z)
+# if we already log transform Z then we use dnorm
+
+ mu[i] <- alpha_0[i] + beta1*log_W[i]
+# log_W is the log transform average relative timing of RSV  
+alpha_0[i]= beta0+x1[i]*lambda1+x2[i]*lambda2
+# this is alpha_0i = beta0 + X_i*lambda (coviariates)
 }
-"
 ```
+
+## Model structure for right censored survival analysis (continued)
+
+```{r, eval=FALSE}
+################ priors ####################################
+beta0~dnorm(0,0.0001)
+beta1~dnorm(0,0.0001)
+lambda1~dnorm(0,0.0001)
+lambda2~dnorm(0,0.0001)
+inv_var ~ dgamma(0.01, 0.01)}"
+```
+\
+\
+Since the model running process is the same, I will skip the model fitting process in this tutorial (it takes time to knit the PDF).
+To run the model, you need to combine the two chunks of model structures and change the eval from False to True.
+
+## Same process to sample posteriors of right censored model
+
+```{r, message=FALSE, eval=FALSE}
+## loading data from the working directory
+data_censor <- readRDS("./censor_data.rds")
+
+### Read in the model file using the jags.model function ##
+censor_model_jags <- jags.model(textConnection(censormodel),
+                     data=data_censor,n.chains = 4)
+
+############ burn-in period ############
+update(censor_model_jags,10000)
+
+######## sample for posterior distribution ##########
+censor_model_sample <- coda.samples(censor_model_jags,
+                            c("beta0","beta1",
+                            "lambda1","lambda2",
+                            "lambda3","inv_var"),
+                             100000,thin=20)
+```
+
+## Section 3: spatial autocorrelation

### 2023-01-20 JAGS_Bayesian_model/R code from the paper_CARBayes- An R Package for Bayesian Spatial Modeling with Conditional Autoregressive Priors.r
add references

@@ -1,0 +1,135 @@
+################################################################
+#### Load the packages required to do an extended analysis
+################################################################
+library("foreign")
+library("shapefiles")
+library("sp")
+library("boot")
+library("Matrix")
+library("nlme")
+library("maptools")
+library("grid")
+library("deldir")
+library("splines")
+library("spdep")
+library("CARBayes")
+set.seed(1)
+
+
+#############################
+#### Example 1 - house prices
+#############################
+#### Read in the data
+data("housedata", package = "CARBayes")
+data("shp", package = "CARBayes")
+data("dbf", package = "CARBayes")
+
+
+#### Remove the outlying observation
+housedata <- housedata[!rownames(housedata) == "S02000655", ]
+
+#### Combine the data and shapefile
+data.combined <- combine.data.shapefile(housedata, shp, dbf)
+
+#### Plot a map of median property price
+northarrow <- list("SpatialPolygonsRescale", layout.north.arrow(), offset = c(220000, 
+  647000), scale = 4000)
+scalebar <- list("SpatialPolygonsRescale", layout.scale.bar(), offset = c(225000, 
+  647000), scale = 10000, fill = c("transparent", "black"))
+text1 <- list("sp.text", c(225000, 649000), "0")
+text2 <- list("sp.text", c(230000, 649000), "5000 m")
+spplot(data.combined, "price", sp.layout = list(northarrow, scalebar, text1, text2), 
+  scales = list(draw = TRUE), at = seq(min(housedata$price) - 1, max(housedata$price) + 
+    1, length.out = 8), col.regions = c("#FEE5D9", "#FCBBA1", "#FC9272", "#FB6A4A", 
+    "#EF3B2C", "#CB181D", "#99000D"))
+
+
+
+#### Transform the price and crime variables
+housedata$logprice <- log(housedata$price)
+housedata$logcrime <- log(housedata$crime)
+housedata$logdriveshop <- log(housedata$driveshop)
+
+#### Fit a covariate only model
+form <- "logprice ~ logcrime + rooms + sales + factor(type) + logdriveshop"
+model <- lm(formula = form, data = housedata)
+summary(model)
+
+#### Compute a Moran's I test on the residuals
[…]

### 2023-01-20 0_HCUP_data_clean/Data directionary.Rmd
change file names to show order

--- 2023-01-13 HCUP_data_clean/Data directionary.Rmd
@@
   - \textcolor{red}{SASData}. This contains the output csv files from the SAS program.
+    \
+  \
+  - \textcolor{red}{Rcode}. This contains the R codes to convert csv files to Parquet files. It also contains the R codes to run exploratory analyses using Parquet data.
+    \
+  \
 
@@
   \
-(5) Download R file: https://github.com/Gigi112/MedicareCMSExploration/blob/main/2.CreateParquet_states.Rmd and modify it to convert all csv files to Parquet files.
+(5) Within \textcolor{red}{Rcode}, run R code *0_2.CreateParquet_states.Rmd* and modify it to convert all csv files to Parquet files.
   \
   \
-(6) Modify R file: https://github.com/VA-CareDisruptions/MedicareCMSExploration/blob/main/1.%20exploratory%20analyses.Rmd to analyze parquet data in R environment.
+(6) Within \textcolor{red}{Rcode}, modify R code *1_2.exploratory analyses.Rmd* to analyze parquet data in R environment.

### 2023-01-20 3_JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
change file names to show order

--- 2023-01-18 JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
@@
 ```{r, eval=FALSE}
+"model{# delete this one line of code
+# when combine the two chunks. Separating the
+# codes helps to show them in slides
+
 ################ priors ####################################
@@
 ## Section 3: spatial autocorrelation
+
+Very often, the disease incidence is not randomly distributed across space.
+Observations in nearby localities tend to be more similar than observations further away, also known as positive spatial autocorrelation.
+\
+\
+Prof.
+Duncan Lee has developed many statistical tools and tutorials to help evaluate spatial autocorrelation in public health.
+Please refer to the following materials to have an complete overview on Bayesian spatial model for public health and choosing conditional autoregressive priors:
+\
+\
+
+- [Bayesian Disease Mapping for Public Health](./Bayesian Disease Mapping for Public Health)
+
+- [Overview of Spatial Incidence Models](https://atlas.cancer.org.au/developing-a-cancer-atlas/Chapter_5.html)
+
+- [CARBayes: An R Package for Bayesian Spatial Modeling with Conditional Autoregressive Priors](./CARBayes: An R Package for Bayesian Spatial Modeling with Conditional Autoregressive Priors.pdf)
+
+- [Spatio-Temporal Areal Unit Modeling in R with Conditional Autoregressive Priors Using the CARBayesST Package](./Spatio-Temporal Areal Unit Modeling in R with Conditional Autoregressive Priors Using the CARBayesST Package.pdf)
+
+## Spatial model with censored data
+```{r, eval=FALSE}
+censor_spatial_model <- "model{
+for(i in 1:n){
+is.censor[i] ~ dinterval(Z[i],t.cen)
+Z[i] ~ dlnorm(mu[i], inv_var)
+ mu[i] <- beta0[i] + beta1*log_W[i]
+}
+
+beta0[1:n] ~ dmnorm(intercepts, Leroux)
+ for(i in 1:n){
+ intercepts[i] <- b0+x1[i]*b1+x2[i]*b2}
+
+beta1~dnorm(0,0.0001)
+
+b0~dnorm(0,0.0001)
+b1~dnorm(0,0.0001)
+b2~dnorm(0,0.0001)
+```
+
+## Spatial model with censored data (continued)
+```{r, eval=FALSE}
+"model{# delete this one line of code
+# when combine the two chunks. Separating the
+# codes helps to show them in slides
+
+Leroux <- inv.tau*(rho*w.mat + (1 - rho)*Id.mat) 
+# calculate spatial correlated random effects
+  inv.tau ~ dgamma(0.01, 0.01) 
+# note the w.mat here is equal to 
+# diag(rowSums(neighbors_mat)) - neighbors_mat
+
+  rho~ dunif(0,1)
+inv_var ~ dgamma(0.01, 0.01)}"
+```

### 2023-01-20 3_JAGS_Bayesian_model/censormodel.rmd
change file names to show order

---
title: "timing_comp"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(rjags)
```

### base model one
```{r}
Nospatial_model <- "model{
for(i in 1:n){
Z[i] ~ dnorm(mu[i], inv_var)
 mu[i] <- (beta0+x1[i]*lambda1+x2[i]*lambda2+lambda3*x3[i])+(beta1+x1[i]*theta1+x2[i]*theta2+theta3*x3[i])*log_W[i]

}
beta0~dnorm(0,0.0001)

beta1~dnorm(0,0.0001)

lambda1~dnorm(0,0.0001)

lambda2~dnorm(0,0.0001)

lambda3~dnorm(0,0.0001)

theta2~dnorm(0,0.0001)

theta1~dnorm(0,0.0001)

theta3~dnorm(0,0.0001)

inv_var ~ dgamma(0.01, 0.01)

}

"

```

```{r}
popden <- read.csv("D:/re-emergent RSV timing/population-density-2021.csv",header = T)
popden <- popden[-c(1,9,11),]
popden$Population.Density.2020.Census <- as.numeric(scale(as.numeric(popden$Population.Density.2020.Census)))
popden$familysize <- as.numeric(scale(popden$familysize))

data <- list(n=47,Z=y,log_W=log_w,x1=popden$Population.Density.2020.Census,x2=popden$familysize,x3=InD[,'Index'])

censor <- jags.model(textConnection(Nospatial_model),data=data,n.chains = 2)

update(censor,10000)

censorsample <- coda.samples(censor,c("beta0","beta1","lambda1","lambda2","theta1","theta2","lambda3","theta3","inv_var","mu"),100000,thin=20)
```

### right censor survival analysis
```{r}
censormodel <- "model{
for(i in 1:n){
is.censor[i] ~ dinterval(Z[i],t.cen)
Z[i] ~ dlnorm(mu[i], inv_var)
 mu[i] <- (beta0+x1[i]*b01+x2[i]*b02)+(beta1+x1[i]*b11+x2[i]*b12)*log_W[i]

}
beta0~dnorm(0,0.001)

beta1~dnorm(0,0.001)

b01~dnorm(0,0.001)

b11~dnorm(0,0.001)

b02~dnorm(0,0.001)

b12~dnorm(0,0.001)

inv_var ~ dgamma(0.01, 0.01)

}

"

```

## add spatial random effects
```{r}
censor_spatial_model <- "model{
for(i in 1:n){
is.censor[i] ~ dinterval(Z[i],t.cen)
Z[i] ~ dlnorm(mu[i], inv_var)
 mu[i] <- beta0[i] + beta1*log_W[i]
}

beta0[1:n] ~ dmnorm(intercepts, Leroux)

 for(i in 1:n){
 intercepts[i] <- b0+x1[i]*b1+x2[i]*b2}

beta1~dnorm(0,0.0001)

b0~dnorm(0,0.0001)

b1~dnorm(0,0.0001)

b2~dnorm(0,0.0001)

Leroux <- inv.tau*(rho*w.mat + (1 - rho)*Id.mat) # calculate spatial correlated random effects
  inv.tau ~ dgamma(0.01, 0.01) # note the w.mat here is equal to diag(rowSums(neighbors_mat)) - neighbors_mat
  rho~ dunif(0,1)

inv_var ~ dgamma(0.01, 0.01)

}

"
```

```{r}
w.mat=diag(rowSums(neighbors_mat)) - neighbors_mat
data <- list(n=47,Z=exp(z),t.cen=exp(censoring_time),is.censor=censored,log_W=x[,4],x1=x[,2],x2=x[,3],Id.mat=diag(47),w.mat=w.mat)

censor <- jags.model(textConnection(censor_spatial_model),data=data,n.chains = 2)

update(censor,50000)

```

### priors on log_w
```{r}
censormodel_var <- "model{
for(i in 1:n){
is.censor[i] ~ dinterval(Z[i],t.cen[i])
Z[i] ~ dlnorm(mu[i], inv_var)
 mu[i] <- (beta0+x1[i]*b01+x2[i]*b02)+(beta1+x1[i]*b11+x2[i]*b12)*log_W[i]

 log_W[i]~ dnorm(mean_lnW[i],inv_var_w[i])

}
beta0~dnorm(0,0.001)

beta1~dnorm(0,0.001)

b01~dnorm(0,0.001)

b11~dnorm(0,0.001)

b02~dnorm(0,0.001)

b12~dnorm(0,0.001)

inv_var ~ dgamma(0.01, 0.01)

}

"

```

```{r}
popden <- read.csv("D:/re-emergent RSV timing/population-density-2021.csv",header = T)
onset.2020.FL <- onset.2020$grey-onset.2020$grey[onset.2020$state=='FL']
Z <- onset.2020.FL[-c(which(states=="HI"),which(states=="AK"),which(states=="FL"))]
t.cen <- max(onset.2020$onset)-onset.2020$onset[onset.2020$state=='FL']
n <- length(Z)
t.cen <- rep(t.cen,n)

x1 <- as.numeric(popden$Population.Density.2020.Census)
x1[30] <- 1263
x1[39] <- 1061.4
x1 <- scale(x1)
x2 <- as.numeric(popden$familysize)
x2 <- scale(x2)
is.censor <- ifelse(is.na(Z),1,0)

data.FL <- list(n=n,Z=Z+0.01,t.cen=t.cen,is.censor=is.censor,W=onset.whole.FL+0.01,x1=as.numeric(x1),x2=as.numeric(x2),mean_lnW=lnW_FL_mean,inv_var_w=rep(1000,50))

data <- list(n=47,Z=Z,t.cen=censoring_time,is.censor=censored,log_W=lnW_FL_mean,x1=x_1,x2=x_2)
tinits1 <- 1+onset.2020$onset
is.na(tinits1) <- complete.cases(onset.2020$grey)
tinits2 <- tinits1+1
censorinit <- list(list(Z=tinits1),list(Z=tinits2))
```

```{r}
censor <- jags.model(textConnection(censormodel),data=data,n.chains = 2)

update(censor,100000)

censorsample <- coda.samples(censor,c("beta0","beta1","b01","b02","b11","b12"),500000,thin=10)
```

```{r}
#par(ask=T)
traceplot(censorsample)
summary(censorsample)
densplot(censorsample)
```

```{r}
censor <- jags.model(textConnection(censormodel),data=data.FL,n.chains = 2)

update(censor,10000)

censorsample <- coda.samples(censor,c("beta0","beta1","b01","b02","b11","b12","log_W"),50000,thin=10)
```

```{r}
#par(ask=T)
traceplot(censorsample)
summary(censorsample)
densplot(censorsample)
```

```{r}
post <- as.data.frame(as.matrix(censorsample))
b12 <- post[,"b12"]
beta1 <- post[,"beta1"]
plot(b12*x2[1]+beta1,type="l")
plot(b12*x2[2]+beta1,type="l")
```

### right censor sensitivity analysis
```{r}
censormodel_sim <- "model{
for(i in 1:n){
is.censor[i] ~ dinterval(Z[i],t.cen[i])
Z[i] ~ dlnorm(mu[i], inv_var)
 mu[i] <- (beta0)+(beta1)*log(W[i])

}
beta0~dnorm(0,0.001)

beta1~dnorm(0,0.001)

inv_var ~ dgamma(0.01, 0.01)

}

"

```

```{r}
y <- onset.2020$onset-onset.2020$onset[onset.2020$state=='FL']
z <- c()
z[censored == 0]<-y[censored == 0]
z[censored == 1]<-NA
y <- y[-c(which(states=="HI"),which(states=="AK"),which(states=="FL"))]
x_1 <- as.numeric(x1)[-c(which(states=="HI"),which(states=="AK"),which(states=="FL"))]
x_2 <- as.numeric(x2)[-c(which(states=="HI"),which(states=="AK"),which(states=="FL"))]

x <- matrix(c(rep(1,length(y)),x_1,x_2,lnW_FL_mean,lnW_FL_mean*x_1,lnW_FL_mean*x_2),nrow=length(y),ncol = 6)

m <- matrix(0,nrow=length(y),ncol = 2*length(y))
for(i in 1:length(y)){
        m[i,(1+(i-1)*2):(2*i)] <- c(1,log_w[i])
}

saveRDS(y,"D:/re-emergent RSV timing/sample_requiredata/y.rds")
saveRDS(x,"D:/re-emergent RSV timing/sample_requiredata/x.rds")
saveRDS(m,"D:/re-emergent RSV timing/sample_requiredata/m.rds")
saveRDS(neighbors_mat,"D:/re-emergent RSV timing/sample_requiredata/neighbors_mat.rds")
saveRDS(censored,"D:/re-emergent RSV timing/sample_requiredata/censored.rds")
saveRDS(censoring_time,"D:/re-emergent RSV timing/sample_requiredata/censoring_time.rds")
```

```{r}
library(rstan)

fit = stan(model_code=bern.stan, data=list(y=y, N=N), iter=5000)
```

### 2023-01-20 1_MSIS_model_explain/code source/contactmatrix.R

@@ -1,0 +1,45 @@
+## read in matrix of effective contacts relevant to airborne infectious diseases
+USAcontactnew <- read.csv("https://raw.githubusercontent.com/mobs-lab/mixing-patterns/main/data/contact_matrices/United_States_country_level_M_overall_contact_matrix_85.csv",header = F)
+
+# visualize contact matrix
+library(RColorBrewer)
+n.cols=100
+nice.cols <-  colorRampPalette(brewer.pal(9, "YlOrRd"))(n.cols)
+heatmap(as.matrix(USAcontactnew)/sum(diag(as.matrix(USAcontactnew))), Rowv=NA, Colv=NA, scale='none', col=nice.cols)
+
+## create a symmetric contact matrix
+symmetricUSA <- 0.5 * (USAcontactnew + t(USAcontactnew))
+
+## expand the contact under 1 year old
+expendUSA_1 <- matrix(data = symmetricUSA[1,1],nrow = 5,ncol=5)
+expendUSA_2 <- matrix(data = as.numeric(rep(symmetricUSA[1,],5)),byrow = T,nrow = 5,ncol=85)
+expendUSA <- cbind(expendUSA_1,expendUSA_2)
+expendUSA <- rbind(expendUSA,cbind(t(expendUSA_2),as.matrix(symmetricUSA)))
+
+
+colnames(expendUSA)=c("<2m","2-3m","4-5m","6-7m","8-9m","10-11m","1Y",rep("2-4Y",3),rep("5-9Y",5),
+                             rep("10-19Y",10),rep("20-39Y",20),
+                             rep("40-59Y",20),rep("60Y+",25))
+
+# aggregate ages into 13 age groups; average the number of contacts within each age group
+ave_mat_x = aggregate( expendUSA, by = list(colnames(expendUSA)), FUN='sum' )
+ave_mat = aggregate( t(ave_mat_x[,-1]), by = list(colnames(expendUSA)), FUN='mean')
+rownames(ave_mat) <- ave_mat[,1]
+colnames(ave_mat)[2:length(ave_mat)] <- rownames(ave_mat)
+library(dplyr)
+ageorder <- c("Group.1","<2m","2-3m","4-5m","6-7m","8-9m","10-11m","1Y","2-4Y","5-9Y","10-19Y","20-39Y","40-59Y","60Y+")
[…]

### 2023-01-20 MSIS_model_explain/SimpleModel/data_and_parms/simple_model.R
rename

@@ -1,153 +1,0 @@
-simple_model <- function(t,y,parms,time.step='month'){
-  
-  # read in initial states and their names
-  States<-array(y, dim=dim(parms$yinit.matrix))
-  dimnames(States) <- dimnames(parms$yinit.matrix)
-  
-  # unify the time unit of parameter inputs
-  if(parms$time.step=='month'){
-    period=12
-    length.step=30.44 #days
-  }else if(parms$time.step=='week'){
-    period=52.1775
-    length.step=7 #days
-  }
-  
-  # waning rate of maternal immunity (by time step)
-  omega = 1/(parms$DurationMatImmunityDays/length.step)
-  
-  # aging rate (by time step)
-  mu= 1/parms$WidthAgeClassMonth
-  if(parms$time.step=='week'){
-    mu= 1/(WidthAgeClassMonth*4.345)}
-  
-  # rate of recovery of first infection
-  gamma1= 1/(parms$dur.days1/length.step) 
-  # rate of recovery of second infection
-  gamma2= 1/(parms$dur.days2/length.step)
-  # rate of recovery of third infection
-  gamma3= 1/(parms$dur.days3/length.step)  
-  gamma4= gamma3  
-  #gamma3 stands for rate of recovery from subsequent infection
-  
-  # Relative risk of infection (2nd)
-  sigma1=parms$sigma1 
-  # Relative risk of infection (3rd)
-  sigma2=parms$sigma2
-  # Relative risk of infection (4th+)
-  sigma3=parms$sigma3
-  
-  # Relative infectiousness (2nd)
-  rho1=parms$rho1 
-  # Relative infectiousness (3rd+)
-  rho2=parms$rho2
-  
-  #Pull out the states for the model as vectors
-  M <-  States[,'M'] # protected by maternal immunity
-  S0 <-  States[,'S0'] # purely susceptible population
-  I1 <-  States[,'I1'] # first time infection (infectious)
-  
-  S1 <-  States[,'S1'] 
-  # susceptible population with build-up immunity
-  I2 <-  States[,'I2'] # second time infection
-  
-  S2 <-  States[,'S2']
-  # susceptible population with lower risk of re-infection
-  I3 <-  States[,'I3'] # third time infection
-  
[…]

### 2023-01-20 MSIS_model_explain/SimpleModel/data_and_parms/waning_immunity.R

@@ -1,168 +1,0 @@
-waning_immunity_model <- function(t,y,parms, time.step='month'){
-  
-  States<-array(y, dim=dim(parms$yinit.matrix))
-  dimnames(States) <- dimnames(parms$yinit.matrix)
-  
-  if(parms$time.step=='month'){
-    period=12
-    length.step=30.44 #days
-  }else if(parms$time.step=='week'){
-    period=52.1775
-    length.step=7 #days
-  }
-  
-  # if(t>=322 & t<= 335){
-  #   I_ex_t= sum(States)*parms$ex_proportion*parms$ex_increase_new[t-321] # ex_increase_new is a vector indicating how 
-  #   # travel restritions and virus activities in other regions affects virus importation
-  # }else{
-  #   I_ex_t <- sum(States)*parms$ex_proportion 
-  #   ## external infections
-  #   ##  I_ex_t=1/100000*St ; I_ex_t=5/100000*St ; I_ex_t=10/100000*St and etc
-  # }
-  # 
-  
-  I_ex_t=parms$I_ex*parms$airtravel[t]
-  
-  omega = 1/(parms$DurationMatImmunityDays/length.step)
-  
-  mu= 1/parms$WidthAgeClassMonth
-  if(parms$time.step=='week'){
-    mu= 1/(WidthAgeClassMonth*4.345)
-  }
-  
-  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
-  gamma2= 1/(parms$dur.days2/length.step)  
-  gamma3= 1/(parms$dur.days3/length.step)  
-  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
-  
-  if(parms$AllowWaning=='Yes'){
-  theta1= 1/(parms$dur.immunity1/length.step) # waning immunity from S3 to W3;from S2 to W2
-  theta2= 1/(parms$dur.immunity2/length.step) }
-  else{theta1= 0 # waning immunity from S3 to W3;from S2 to W2
-  theta2= 0}# waning immunity from W3 to S2;from W2 to S1
-  
[…]

### 2023-01-20 MSIS_model_explain/code source/Loglikelihood_MSIS.R

@@ -1,83 +1,0 @@
-fitmodel <-  function(parameters,dat) {   # takes the parameter values and dataset as inputs
- 
-  protrans <- parameters[1] # estimate parameter related to R0 (baseline transmission rate)
-  b1 <- parameters[2] # estimate parameter related to seasonal amplitude
-  trans <- parameters[3] # estimate parameter related to seasonal peak timing
-  DMD <- parameters[4] # estimate parameter related to the duration of maternal immunity
-  Amp <- exp(b1) #ensure positive 
-  baseline.txn.rate <- exp(protrans) #ensure positive
-  phi <-  (2*pi*(exp(trans))) / (1+exp(trans)) # transform to its scale in the model
-  durx <- exp(DMD) #ensure positive
-  
-  #we need this for the function and model to recognize these parameters
-  durx <<- durx 
-  baseline.txn.rate <<- baseline.txn.rate
-  Amp <<- Amp
-  phi <<- phi
-  
-  
-  # Run transmission model with initial conditions 
-  # and timesteps defined above, and parameter values from function call
-  results <- ode(y=yinit.vector, t=my_times,  
-                 func=simple_model, 
-                 parms=c(parmset,
-                         baseline.txn.rate=baseline.txn.rate,
-                         Amp=Amp,
-                         phi=phi,
-                         DurationMatImmunityDays=durx))
-  
-  t0 <- parmset$t0 
-  # make sure the evaluation period are the same
-  results<- tail(results,t0) # get rid of burn-in period
-  St <- results[,-1]
-  
-  I1 <- St[,grep('I1', colnames(St))]
-  I2 <- St[,grep('I2', colnames(St))]
-  I3 <- St[,grep('I3', colnames(St))]
-  I4 <- St[,grep('I4', colnames(St))]
-  S0 <- St[,grep('S0', colnames(St))]
-  S1 <- St[,grep('S1', colnames(St))]
-  S2 <- St[,grep('S2', colnames(St))]
-  S3 <- St[,grep('S3', colnames(St))]
-  
-  
-  b=baseline.txn.rate/(parmset$dur.days1/30.44)
-  beta=(b/100)/(sum(yinit.matrix)^(1-parmset$q))*parmset$contact
-  
[…]

### 2023-01-20 MSIS_model_explain/code source/contactmatrix.R

@@ -1,45 +1,0 @@
-## read in matrix of effective contacts relevant to airborne infectious diseases
-USAcontactnew <- read.csv("https://raw.githubusercontent.com/mobs-lab/mixing-patterns/main/data/contact_matrices/United_States_country_level_M_overall_contact_matrix_85.csv",header = F)
-
-# visualize contact matrix
-library(RColorBrewer)
-n.cols=100
-nice.cols <-  colorRampPalette(brewer.pal(9, "YlOrRd"))(n.cols)
-heatmap(as.matrix(USAcontactnew)/sum(diag(as.matrix(USAcontactnew))), Rowv=NA, Colv=NA, scale='none', col=nice.cols)
-
-## create a symmetric contact matrix
-symmetricUSA <- 0.5 * (USAcontactnew + t(USAcontactnew))
-
-## expand the contact under 1 year old
-expendUSA_1 <- matrix(data = symmetricUSA[1,1],nrow = 5,ncol=5)
-expendUSA_2 <- matrix(data = as.numeric(rep(symmetricUSA[1,],5)),byrow = T,nrow = 5,ncol=85)
-expendUSA <- cbind(expendUSA_1,expendUSA_2)
-expendUSA <- rbind(expendUSA,cbind(t(expendUSA_2),as.matrix(symmetricUSA)))
-
-
-colnames(expendUSA)=c("<2m","2-3m","4-5m","6-7m","8-9m","10-11m","1Y",rep("2-4Y",3),rep("5-9Y",5),
-                             rep("10-19Y",10),rep("20-39Y",20),
-                             rep("40-59Y",20),rep("60Y+",25))
-
-# aggregate ages into 13 age groups; average the number of contacts within each age group
-ave_mat_x = aggregate( expendUSA, by = list(colnames(expendUSA)), FUN='sum' )
-ave_mat = aggregate( t(ave_mat_x[,-1]), by = list(colnames(expendUSA)), FUN='mean')
-rownames(ave_mat) <- ave_mat[,1]
-colnames(ave_mat)[2:length(ave_mat)] <- rownames(ave_mat)
-library(dplyr)
-ageorder <- c("Group.1","<2m","2-3m","4-5m","6-7m","8-9m","10-11m","1Y","2-4Y","5-9Y","10-19Y","20-39Y","40-59Y","60Y+")
[…]

### 2023-01-20 3_JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
add new

@@
 
+## A generalized linear mixed effect model (GLMM) framework
+
+To produce accurate estimates of disease incidence, statistical models must take into account the full range of features present in the data, including non-normal distributions of count data, the impact of explanatory variables, and spatial correlation.
+One solution is to utilize generalized linear mixed models (GLMMs), which integrate a generalized linear model (GLM) with normally distributed random effects that account for both spatial correlation and overdispersion.
+\
+\
+An example model structure is given as:
+$$Y_i \sim Poisson (\mu_i)$$
+$$\log \mu_i= \log E_i+x_i^\prime \alpha+\phi_i$$
+where $Y_i$ is the observed number of disease in region $i$, $\mu_i$ is the  conditional mean, $E_i$ is the expected number of disease in region $i$, $x_i$ is the explanatory variables (covariates) of region $i$, $\alpha$ is the coeffiecients of the corresponding explanatory variables, and $\phi_i$ accounts for the random effects, which can be spatially correlated.
+
+## Conditional autoregressive (CAR) priors for spatial modeling
+
+To model spatial autocorrelation, four CAR priors are commonly used, including the intrinsic models, BYM models, Leroux models, and Cressie models.
+In [**Duncan Lee's paper**](https://www.sciencedirect.com/science/article/pii/S1877584511000049?via%3Dihub), he systematically compared these four different priors and concluded that Leroux model is the best from both theoretical and practical standpoints.
+For short, Leroux model is appealing because it uses a single set of random effects to model the varying strengths of spatial autocorrelation $\rho$, where $\rho=0$ means spatially independent while $\rho$ near 1 indicates strong spatial correlation.
+
+## Leroux priors for spatial modeling
+
+The Leroux version of spatial random effects is given as:
+$$\phi_i | \boldsymbol{\phi_{-i}},\rho,\tau^2 \sim MVN (\frac{\rho\sum_{j=1}^nw_{ij}\phi_i}{\rho\sum_{j=1}^nw_{ij}+1-\rho},\frac{\tau^2}{\rho\sum_{j=1}^nw_{ij}+1-\rho})$$
+where $\rho \in [0,1)$ represents the spatial correlation that was estimated from the data; $\rho=0$ corresponds to spatial independence, simplifying the model to an independent random effects model, and $\rho$ near 1 indicates strong spatial correlation.
+The $w_{ij}$ are variables that contain information about spatial proximity between two locations; $w_{ij}$ equals one if regions i and j share a common border and equal zero otherwise ($w_{ii}$=0 for all i).
+The total variance of spatial random effects is described by $\tau^2$.
+
+## Create a spatial adjacency matrix 
+###  Step 1. Set up -- Shapefiles
+
+(I learned this from Kayoko Shioda!)
+
+```{r,message=FALSE,warning=FALSE}
+# Load a shapefile. Make sure each polygon on shapefile  
+# corresponds to only one state. 
+require(rgdal) # for working with shapefiles
+require(ggplot2) # for plot and fortify
+CT1map = readOGR(dsn=path.expand('./WGS84'), layer= 'zipct_37800_0000_2010_s100_census_1_shp_wgs84')
+CTmap.df <- as.data.frame(CT1map)
+CT1map_fort <- fortify(CT1map,region="ZCTA5CE10")
+# ZCTA5CE10 is CT zipcode ID
+# The ID variable is then the zipcode/area unit 
+CT1map_fort$zipcode <- CT1map_fort$id
+
+# Sort both the shapefile and the data file by ID.
+# This is the most important step
+CTmap.sort <- CT1map[order(CT1map$ZCTA5CE10),]
+```
+
+## How to create a spatial adjacency matrix
+
+###  Step 2. Create a neighbor matrix from the sorted shapefile 
+```{r,message=FALSE}
+library(spdep) # for neighbor evaluation
+
+# Neighbors can either be Queen 
+# (any zip that touches another zip - even at corners) 
+# or Rook neighbors (share full sides -- not corners)
+
+# The snap dist is governing the NA in neighbor matrix
+# Can remove SNAP statement; 
+# if Queen = F then single point touch not allowed 
+# and Rook distance used.
+
+# When data is missing -- there's more non-zero links involved 
+# Try to make snap distance as small as possible 
+# while making sure all zipcode have at least one neighbor
+neighb <- poly2nb(CTmap.sort, queen = T, snap = sqrt(0.001))
+
+# Check average number of links -- increase snap distance 
+# if it says there is a zip with zero neighbors
+neighb # Avg Number of links = 8.18705
+
+# Make neighbor matrix 
+# if zero.policy FALSE stop with error for any empty neighbour sets, 
+# if TRUE permit the weights list to be formed with zero-length weights vectors
+# B is the basic binary coding, W is row standardised (both are commonly used)
+neighb.mat <- nb2mat(neighb, zero.policy = T, style = 'B')
+```
+
+## How to create a spatial adjacency matrix
+
+###  Step 3. Create a W matrix for Leroux model 
+```{r,message=FALSE}
+# For Leroux model, neighbor matrix needs to have 0 
+# if not neighboring, -1 if neighbor, 
+# and number of neighbors for each zip on diagonal), so...
+w.mat <- ifelse(neighb.mat==1,-1,neighb.mat)
+num.neigh <- colSums(neighb.mat) 
+# Count the number of neighbors for each state
+for (i in 1:nrow(w.mat)) {
+  # Put the number of neighbors on diagonal
+  for (j in 1:ncol(w.mat)) {
+    if (i==j) {
+      w.mat[i,j] <- num.neigh[i]
+    }
+  }
+}
+```
+
 ## Spatial model with censored data

### 2023-01-23 3_JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
update jags model

@@
 
-```{r,message=FALSE,warning=FALSE}
+```{r message=FALSE,warning=FALSE}
 # Load a shapefile. Make sure each polygon on shapefile  
@@
 require(ggplot2) # for plot and fortify
-CT1map = readOGR(dsn=path.expand('./WGS84'), layer= 'zipct_37800_0000_2010_s100_census_1_shp_wgs84')
+CT1map = readOGR(dsn=path.expand('./WGS84'), 
+layer= 'zipct_37800_0000_2010_s100_census_1_shp_wgs84')
 CTmap.df <- as.data.frame(CT1map)
@@
 CT1map_fort$zipcode <- CT1map_fort$id
-
-# Sort both the shapefile and the data file by ID.
-# This is the most important step
-CTmap.sort <- CT1map[order(CT1map$ZCTA5CE10),]
 ```
@@
 
-###  Step 2. Create a neighbor matrix from the sorted shapefile 
+###  Step 2. Find the number of neighbors from the sorted shapefile  
 ```{r,message=FALSE}
+# Sort both the shapefile and the data file by ID.
+# This is the most important step
+CTmap.sort <- CT1map[order(CT1map$ZCTA5CE10),]
+
 library(spdep) # for neighbor evaluation 
-
 # Neighbors can either be Queen 
@@
 neighb <- poly2nb(CTmap.sort, queen = T, snap = sqrt(0.001))
+```
 
+## How to create a spatial adjacency matrix
+
+###  Step 3. Create a neighbor matrix with 1 suggesting adjacency 
+```{r,message=FALSE}
 # Check average number of links -- increase snap distance 
@@
 # Make neighbor matrix 
-# if zero.policy FALSE stop with error for any empty neighbour sets, 
-# if TRUE permit the weights list to be formed with zero-length weights vectors
-# B is the basic binary coding, W is row standardised (both are commonly used)
+# if zero.policy FALSE,
+# stop with error for any empty neighbor sets, 
+# if TRUE, permit the weights list to be
+# formed with zero-length weights vectors
+# B is the basic binary coding, 
+# W is row standardised (both are commonly used)
 neighb.mat <- nb2mat(neighb, zero.policy = T, style = 'B')
@@
 
-###  Step 3. Create a W matrix for Leroux model 
+###  Step 3. Create the W matrix for Leroux model 
 ```{r,message=FALSE}

### 2023-01-24 1_MSIS_model_explain/MSIS_model_explain.Rmd
update example

--- 2023-01-02 MSIS_model_explain/MSIS_model_explain.Rmd
@@
 
-This age structured transmission dynamic model of RSV was first developed by Ginny Pitzer in her publication: https://doi.org/10.1371/journal.ppat.1004591.
+This age structured transmission dynamic model of RSV was first developed by Ginny Pitzer in her publication: [Environmental Drivers of the Spatiotemporal Dynamics of Respiratory Syncytial Virus in the United States](https://doi.org/10.1371/journal.ppat.1004591).
 With the help of Dan Weinberger, we translated Ginny's Matlab codes to R codes.
@@
 
-- https://doi.org/10.1001/jamanetworkopen.2021.41779
-- https://doi.org/10.1038/s41541-022-00550-5
-- https://doi.org/10.1101/2022.11.10.22282132
-
-The complete code example for this tutorial please refer to
-- https://github.com/weinbergerlab/RSV_metapop/tree/master/SimpleModel
+- [Estimation of the Timing and Intensity of Reemergence of Respiratory Syncytial Virus Following the COVID-19 Pandemic in the U](https://doi.org/10.1001/jamanetworkopen.2021.41779)
+- [Predicted effectiveness of vaccines and extended half-life monoclonal antibodies against RSV hospitalizations in children](https://doi.org/10.1038/s41541-022-00550-5)
+- [Year-round RSV Transmission in the Netherlands Following the COVID-19 Pandemic - A Prospective Nationwide Observational and Modeling Study](https://doi.org/10.1101/2022.11.10.22282132)
 
@@
 
-![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)
+![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/1_MSIS_model_explain/RSV_msis_model_structure.png)
 
@@
 
-![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)
+![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/1_MSIS_model_explain/RSV_msis_model_structure.png)
 \
@@
 
-![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/MSIS_model_explain/RSV_msis_model_structure.png)
+![Ginny's model](/Users/zhezheng/Library/CloudStorage/Box-Box/RSV_tutorials/RSV_tutorials/1_MSIS_model_explain/RSV_msis_model_structure.png)
 ::::
@@
 
-- Each white compartment correspond to a disease status. These statuses are **observable**. These outputs of transmission dynamic models will be used for the model calibration.
+- Each grey compartment correspond to a disease status. These statuses are **observable**. These outputs of transmission dynamic models will be used for the model calibration.
 \
@@
 
-Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/MSIS_model_explain/Introduction of Rstan.pptx) from the separated tutorials in the same file path. 
+Before we start, please learn the [Rstan basic](https://github.com/Gigi112/RSV_tutorials/tree/main/1_MSIS_model_explain/Introduction of Rstan.pptx) from the separated tutorials in the same file path. 
 \

### 2023-01-24 1_MSIS_model_explain/MSIS_model_explain.Rmd
update example

@@
 
-## References
+## Section outlines
+\
+- In the first section, we introduces the background knowledge of transmission dynamic models of RSV.
+\
+\
+- In the second section, we will learn how to run transmission model with all parameters known. 
+\
+\
+- In the third section, we will learn to use maximum a posteriori / maximum likelihood estimation (MLE) to estimate the unknown parameters (one value for each unknown parameter). 
+\
+\
+- In the fourth section, we will learn to use STAN, a probabilistic programming language for statistical inference^[https://en.wikipedia.org/wiki/Stan_(software)], to get the 95% credible intervals from the posterior samples of these unknown parameters.
+
+## Section 1: prerequisite readings (references)
 
@@
 
-## Section outlines
-\
-- In the first section, we will learn how to run transmission model with all parameters known. 
-\
-\
-- In the second section, we will learn to use maximum a posteriori / maximum likelihood estimation (MLE) to estimate the unknown parameters (one value for each unknown parameter). 
-\
-\
-- In the third section, we will learn to use STAN, a probabilistic programming language for statistical inference^[https://en.wikipedia.org/wiki/Stan_(software)], to get the 95% credible intervals from the posterior samples of these unknown parameters.
-
-## Section 1: run transmission model with all parameters known
+## Section 2: run transmission model with all parameters known
 \
@@
 
-## Section 2: use maximum a posteriori (variation of MLE) to estimate the unknown parameters
+## Section 3: use maximum a posteriori (variation of MLE) to estimate the unknown parameters
 
@@
 
-## Section 3: Stan model to estimate posterior samples
+## Section 4: Stan model to estimate posterior samples
 

### 2023-01-26 3_JAGS_Bayesian_model/Jags_Bayesian_model.Rmd
correct minor mistakes

@@
                             "lambda1","lambda2",
-                            "lambda3","inv_var"),
+                            "inv_var"),
                              100000,thin=20)
@@
 ```
+
+## Same process to sample posteriors of spatiotemporal model
+
+```{r, message=FALSE, eval=FALSE}
+## loading data from the working directory
+spatial_model_data <- readRDS("./spatial_model_data.rds")
+
+### Read in the model file using the jags.model function ##
+spatial_model_jags <- jags.model(textConnection(censor_spatial_model),
+                     data=spatial_model_data,n.chains = 4)
+
+############ burn-in period ############
+update(spatial_model_jags,10000)
+
+######## sample for posterior distribution ##########
+spatial_model_sample <- coda.samples(spatial_model_jags,
+                            c("beta0","beta1",
+                            "inv_var",
+                            "b0","b1","b2"),
+                             100000,thin=20)
+```

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
add materials

---
title: "Estimate RSV disease burden using hierarchical baysian regression"
author: "ZHE ZHENG"
date: "2023-01-26"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Outline

### 2023-01-26 5_roadmaps_future_projects/RSV_elderly_vac_cost_eff/Hosp_burden/Hosp_burden_script.Rmd
add materials

---
title: "RSV burden in older adults by age and risk groups"
author: "ZHE ZHENG"
date: "2022-12-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Read in time-series data
```{r}
#####################################################################################
# read in dataset and data cleaning
##################################################################################### 
#getwd()
#rm(list = ls())
library(data.table)
library(sas7bdat)

# the time series data will come from SAS output
# we only select the variable that we are interested because the whole data set is huge
# For the cost-effectiveness study, we should add the cost variables 
Wholeresp <- fread("/Users/zhezheng/Box/data/data/recoded_all_states_resp.csv",select = c("agemonth","zip","AGE","HOSPST","resp","resp_prim","flu","flu_prim","rsv","rsv_prim","resp_other","resp_otherprim","bronchio",'bronchio_prim',"pneumo_other","pneumo_otherprim","pneumopneumo","pneumopneumo_prim","pneumosept","pneumosept_prim","asthma","amonthdate","agecat1","riskcat"))

# Depends on the output of the SAS files. fread is faster
# Wholeresp <- read.sas7bdat("/Users/zhezheng/Box/data/data/recoded_all_states_resp.sas7bdat")
```

```{r}
# If interested in analyzing each state, uncomment these codes

# NYresp <- Wholeresp[HOSPST=="NY",]
# AZresp <- Wholeresp[HOSPST=="AZ",]
# FLresp <- Wholeresp[HOSPST=="FL",]
```

```{r}
# include only the complete cases
Wholeresp <- na.omit(Wholeresp)
# recode the date variable
Wholeresp$amonthdate<- as.Date(Wholeresp$amonthdate, "%d%b%Y")

# make sure the age is a numeric variable not character
Wholeresp$AGE <- as.numeric(Wholeresp$AGE)

# Consider using epidemiological calendar
# Wholeresp.1 <- Wholeresp[Wholeresp$amonthdate>"2016-06-01"&Wholeresp$amonthdate<"2021-07-01",]

# sum over individual records to create monthly time-series of RSV cases, flu cases, respiratory hospitalizations by age and risk category 
resp <- Wholeresp[,list(resp = sum(resp),rsv=sum(rsv),flu=sum(flu)), by = 'amonthdate,AGE,riskcat'] # sum over
```

```{r}
# recode age categories to reflect what we interests: "<1","1-<2","2-<5","5-9","10-19","20-59","60-64","65-69","70-74","75-79","80+"

resp$agecat_new[resp$AGE==0]  <- "<1"
resp$agecat_new[resp$AGE==1]  <- "1-<2"
resp$agecat_new[1<resp$AGE&resp$AGE<1]  <- "2-<5"
resp$agecat_new[5<=resp$AGE&resp$AGE<10]  <- "5-<10"
resp$agecat_new[5<=resp$AGE&resp$AGE<10]  <- "5-<10"
resp$agecat_new[10<=resp$AGE&resp$AGE<20]  <- "10-<20"
resp$agecat_new[20<=resp$AGE&resp$AGE<60]  <- "20-<60"
resp$agecat_new[60<=resp$AGE&resp$AGE<65]  <- "60-<65"
resp$agecat_new[65<=resp$AGE&resp$AGE<70]  <- "65-<70"
resp$agecat_new[70<=resp$AGE&resp$AGE<75]  <- "70-<75"
resp$agecat_new[75<=resp$AGE&resp$AGE<80]  <- "75-<80"
resp$agecat_new[80<=resp$AGE]  <- "80+"
```

```{r}
# reshape to create a respiratory hospitalizations array that has a column for time (48 months), a column for age (11 groups) and a column for risk groups (2 groups, may increase. For hierarchical structure, we need at least 3-4 groups to make)
# adjust the first value to reflect the length of time

# test
df2 <- data.frame(x = rep(1:48,22),y = rep(1:2,528), z=rep(1:11,96))

Array1 <- array(data = unlist(df2),
                  dim = c(48, 11, 2),
                dimnames = list(paste0("month",1:48),                              c("<1","1-<2","2-<5","5-9","10-19","20-59","60-64","65-69","70-74","75-79","80+"),
                                c("high risk","low risk")))
```

```{r}
# create a reported RSV array that has a column for time (48 months), a column for age (11 groups) and a column for risk groups (2 groups)

# adjust the first value (48) to reflect the length of time
y <- array(data=NA,dim = c(48,11,2),
           dimnames = list(paste0("month",1:48),                              c("<1","1-<2","2-<5","5-9","10-19","20-59","60-64","65-69","70-74","75-79","80+"),
                                c("high risk","low risk")))

```

### 2023-01-26 5_roadmaps_future_projects/RSV_elderly_vac_cost_eff/RSV_costeffect_elderly.Rmd
add materials

---
title: "Project outline: RSV vaccination in older adults"
author: "ZHE ZHENG"
date: "2022-11-16"
output:
  html_document: default
  pdf_document: default
---

### Step 1. Estimate the true burden of hospitalizations across risk and age groups in older adults

##### Background

Risk groups will be defined by ACIP.  
references: [ACIP recommendations](https://www.cdc.gov/vaccines/vpd/pneumo/hcp/who-when-to-vaccinate.html#adults-19-64)

* Low risk population
* High risk population: For adults with any of the conditions or risk factors listed below: Adults who have chronic pulmonary (including asthma), cardiovascular (excluding isolated hypertension), renal, hepatic, neurologic, hematologic, or metabolic disorders (including diabetes mellitus).
Persons who are immunocompromised due to any cause (including but not limited to immunosuppression caused by medications or HIV infection).

##### Estimate the "true" burden of RSV hospitalizations and ED visits
Model structure:\
$Y_{ijk}$ ~ Negative Binomial ($p_{ijk}$,$r$) \
$p_{ijk}=r/(r+λ_{ijk})$ \
where $Y_{ijk}$ denotes the number of all-cause respiratory hospitalizations at time (month) $i$, in age group $j$, and risk group $k$. 
$i = 1,2,...,48$ \
We will analyze hospitalizations in 11 age categories (<1, 1, 2-4, 5-9, 10-19, 20-59, 60-64, 65-69, 70-74, 75-79, 80+ years).
$j = 1,2,...,12$ 
\
\
We define the expected value as a function of covariates and random effects such that:

$$\lambda_{ijk}=\beta_{0jk}+\alpha_{1g(i)}+\alpha_{2m(i)}+\beta_{1jk}RSV_{i}+\beta_{2g(i)jk}Flu_{ik}$$
<span style="color:red">  $RSV_{i}$ </span> (means every risk group has the same RSV timing)
\
\
The number of RSV hospitalizations/ED visits in each age group and risk will be <span style="color:blue">$\beta_{1jk}RSV_{i}$</span>

##### Estimate death rate in each risk group

(1) Include DIED element in SAS code
(2) Calculate the death rate in each risk and age group and assume that the death rate is the same between reported cases and the cases that did not reported

\
\

### Step 2.  Fitting transmission model to time-series output

Parameters to estimate (in total 6?):

(1) Transmission parameter (related to transmission probability)
(2) Amplitude of seasonality
(3) Timing of seasonality
(4) Duration of maternal immunity
(5) Proportion of third LRIs that are hospitalized in
older adults (if we assume their contact patterns are similar, we only need to estimate two parameters)
* 60-64 (low $H_{60l}$ and high risk $\frac{\beta_{60h}}{\beta_{60l}} H_{60l}$)
* 65-69 (low $\frac{\beta_{65l}}{\beta_{60l}}H_{60l}$ and high risk $\frac{\beta_{65h}}{\beta_{60l}}H_{60l}$)
* 70-74 (low $\frac{\beta_{70l}}{\beta_{60l}}H_{60l}$ and high risk $\frac{\beta_{70h}}{\beta_{60l}}H_{60l}$)
* 75-79 (low $\frac{\beta_{75l}}{\beta_{60l}}H_{60l}$ and high risk $\frac{\beta_{75h}}{\beta_{60l}}H_{60l}$)
* 80+ years (low $H_{80l}$ and high risk $\frac{\beta_{80h}}{\beta_{80l}}H_{80l}$)
\
\

### Step 3.  Vaccine efficacy

References:

(1) [Efficacy on infectiousness,preF vaccine](https://www.nejm.org/doi/full/10.1056/NEJMoa2116154)
(2) [Efficacy on infectiousness, vector vaccine](https://academic.oup.com/jid/article/226/3/396/6064820)
(3) [Efficacy on symptomatic infection, Pfizer](https://www.pfizer.com/news/press-release/press-release-detail/pfizer-announces-positive-top-line-data-phase-3-trial-older)
(4) [Efficacy on symptomatic infection, GSK](https://www.gsk.com/en-gb/media/press-releases/gsk-s-older-adult-respiratory-syncytial-virus-rsv-vaccine-candidate/)
(5) [Efficacy on symptomatic infection, Janssen](https://www.jnj.com/janssen-announces-phase-2b-data-demonstrating-its-investigational-rsv-adult-vaccine-provided-80-protection-against-lower-respiratory-infections-in-older-adults)

vaccine effects:

(1) Reduce the risk of infection? $(1-VE_{s})\sigma_3\lambda$

\ where $VE_{s}$ is the vaccine efficacy against infection, $\sigma_3$ is the relative susceptibility during subsequent infections, and $\lambda$ is the force of transmission.

(2) Reduce the infectiousness of vaccinees?
$(1-VE_{i})I_v$

\ where $VE_{i}$ is the vaccine efficacy to reduce infectiousness and $I_v$ is the number of vaccinated individuals who are infected by wild-type RSV

(3) Duration of protection? Assume 1 year as a start and sensitivity analysis of 2 years and 5 years

(4) In later stage: we can add age related immunogenicity?
\
\

### Step 4.  Economic Model

###### Health benefits and related costs
(1) Treatment outcomes and costs
(2) QALY estimation
(3) Vaccination costs

* Sample from the uncertainty distributions of the model parameters while running the model 1,000 times to generate 95% prediction interval for the number of infections, GP, ED, hospitalizations, and deaths averted

* Calculate the average net monetary benefit of each vaccination strategy (and no vaccination) for WTP values varying from $10,000 to $50,000 per QALY averted (in increments of $100).

* Plot the Cost-Effectiveness Acceptability Frontier

### 2023-01-26 5_roadmaps_future_projects/RSV_elderly_vac_cost_eff/RSV_costeffect_elderly.tex
add materials

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Scale hospitalizations},
  pdfauthor={ZHE ZHENG},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}

\title{Scale hospitalizations}
\author{ZHE ZHENG}
\date{2022-11-16}

\begin{document}
\maketitle

\hypertarget{estimate-the-true-burden-of-hospitalizations-across-risk-and-age-groups-in-older-adults}{%
\subsection{Estimate the true burden of hospitalizations across risk and
age groups in older
adults}\label{estimate-the-true-burden-of-hospitalizations-across-risk-and-age-groups-in-older-adults}}

\hypertarget{background}{%
\subparagraph{Background}\label{background}}

Risk groups will be defined by The Elixhauser Comorbidity Indices.
{(Question: how to stratify risk groups based on the indices?} For
example, scores are grouped into categories of less than 0, 0, 1--4, and
5 or higher. Chronic pulmonary disease is 3 points and Congestive heart
failure is 7 points) references:
\href{https://www.jstor.org/stable/40221931?seq=7\#metadata_info_tab_contents}{The
Elixhauser Comorbidity Index score and the probablity of inpatient
death}

\begin{itemize}
\tightlist
\item
  Low risk population: Index \textless= 0
\item
  Medium risk population: Index 1-4
\item
  High risk population: 5 or higher
\end{itemize}

\hypertarget{estimate-the-true-burden-of-rsv-hospitalizations-and-ed-visits}{%
\subparagraph{Estimate the ``true'' burden of RSV hospitalizations and
ED
visits}\label{estimate-the-true-burden-of-rsv-hospitalizations-and-ed-visits}}

Model structure:\\
\(Y_{ijk}\) \textasciitilde{} Negative Binomial (\(p_{ijk}\),\(r\))\\
\(p_{ijk}=r/(r+λ_{ijk})\)\\
where \(Y_{ijk}\) denotes the number of all-cause respiratory
hospitalizations at time (month) \(i\), in age group \(j\), and risk
group \(k\). \(i = 1,2,...,48\)\\
We will analyze hospitalizations in 11 age categories (\textless1, 1,
2-4, 5-9, 10-19, 20-59, 60-64, 65-70, 70-74, 75-79, 80+ years).
\(j = 1,2,...,12\)\\
\strut \\
We define the expected value as a function of covariates and random
effects such that:

\[\lambda_{ijk}=\beta_{0jk}+\alpha_{1g(i)}+\alpha_{2m(i)}+\beta_{1jk}RSV_{ik}+\beta_{2g(i)jk}Flu_{ik}\]
{Question 1: Should we use risk group specific RSV indicator
\(RSV_{ik}\)} (means RSV hospitalizations in high-risk or low-risk
children) or{ \(RSV_{i}\) } (means every risk group has the same RSV
timing) ?\\
\strut \\
The number of RSV hospitalizations/ED visits in each age group and risk
will be {\(\beta_{1jk}RSV_{ik}\)}

\hypertarget{estimate-death-rate-in-each-risk-group}{%
\subparagraph{Estimate death rate in each risk
group}\label{estimate-death-rate-in-each-risk-group}}

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  Include DIED element in SAS code
\item
  Calculate the death rate in each risk and age group and assume that
  the death rate is the same between reported cases and the cases that
  did not reported
\end{enumerate}

\hypertarget{fitting-transmission-model-to-time-series-output}{%
\subparagraph{Fitting transmission model to time-series
output}\label{fitting-transmission-model-to-time-series-output}}

Parameters to estimate (in total 6?):

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  Transmission parameter (related to transmission probability)
\item
  Amplitude of seasonality
\item
  Timing of seasonality
\item
  Duration of maternal immunity
\item
  Proportion of third LRIs that are hospitalized in older adults (if we
  assume their contact patterns are similar, we only need to estimate
  two parameters)
\end{enumerate}

\begin{itemize}
\tightlist
\item
  60-64 (low \(H_{60l}\), medium
  \(\frac{\beta_{60m}}{\beta_{60l}} H_{60l}\), and high risk
  \(\frac{\beta_{60h}}{\beta_{60l}} H_{60l}\))
\item
  65-69 (low \(\frac{\beta_{65l}}{\beta_{60l}}H_{60l}\), medium
  \(\frac{\beta_{65m}}{\beta_{60l}}H_{60l}\), and high risk
  \(\frac{\beta_{65h}}{\beta_{60l}}H_{60l}\))
\item
  70-74 (low \(\frac{\beta_{70l}}{\beta_{60l}}H_{60l}\), medium
  \(\frac{\beta_{70m}}{\beta_{60l}}H_{60l}\), and high risk
  \(\frac{\beta_{70h}}{\beta_{60l}}H_{60l}\))
\item
  75-79 (low \(\frac{\beta_{75l}}{\beta_{60l}}H_{60l}\), medium
  \(\frac{\beta_{75m}}{\beta_{60l}}H_{60l}\), and high risk
  \(\frac{\beta_{75h}}{\beta_{60l}}H_{60l}\))
\item
  80+ years (low \(H_{80l}\), medium
  \(\frac{\beta_{80m}}{\beta_{80l}}H_{80l}\), and high risk
  \(\frac{\beta_{80h}}{\beta_{80l}}H_{80l}\))\\
\end{itemize}

\hypertarget{vaccine-efficacy}{%
\subparagraph{Vaccine efficacy}\label{vaccine-efficacy}}

References:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  \href{https://www.nejm.org/doi/full/10.1056/NEJMoa2116154}{Efficacy on
  infectiousness,preF vaccine}
\item
  \href{https://academic.oup.com/jid/article/226/3/396/6064820}{Efficacy
  on infectiousness, vector vaccine}
\item
  \href{https://www.pfizer.com/news/press-release/press-release-detail/pfizer-announces-positive-top-line-data-phase-3-trial-older}{Efficacy
  on symptomatic infection, Pfizer}
\item
  \href{https://www.gsk.com/en-gb/media/press-releases/gsk-s-older-adult-respiratory-syncytial-virus-rsv-vaccine-candidate/}{Efficacy
  on symptomatic infection, GSK}
\item
  \href{https://www.jnj.com/janssen-announces-phase-2b-data-demonstrating-its-investigational-rsv-adult-vaccine-provided-80-protection-against-lower-respiratory-infections-in-older-adults}{Efficacy
  on symptomatic infection, Janssen}
\end{enumerate}

vaccine effects:

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  Reduce the risk of infection? \((1-VE_{s})\sigma_3\lambda\)
\end{enumerate}

~where \(VE_{s}\) is the vaccine efficacy against infection,
\(\sigma_3\) is the relative susceptibility during subsequent
infections, and \(\lambda\) is the force of transmission.

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\setcounter{enumi}{1}
\tightlist
\item
  Reduce the infectiousness of vaccinees? \((1-VE_{i})I_v\)
\end{enumerate}

~where \(VE_{i}\) is the vaccine efficacy to reduce infectiousness and
\(I_v\) is the number of vaccinated individuals who are infected by
wild-type RSV

\end{document}

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/Antibody_Sim.Rmd
add materials

--- 2022-12-31 MSIS_model_explain/SimpleModel/simple-age-structure model.Rmd
@@
 ---
-title: "age structure"
-author: "ZHE ZHENG, Dan Weinberger, modified from Matlab code by Virginia Pitzer"
-date: "12/26/2020"
+title: "Antibody_Sim"
+author: "ZHE ZHENG"
+date: '2022-06-15'
 output: html_document
 ---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+```
+
+```{r}
+time <- 1:(12*9) # time series data point
+period = 12 # time unit of observation (12 corresponds to months; 52 corresponds to weeks)
+AMP = 1.5 # amplitude of antibody seasonality
+phase= 1.6 # shift of peak timing at the beginning
+#Naturally accuring maternal antibdoy levels/ harmonic term to represent seasonal waves
+mab.p <- 13+AMP*cos(2*pi*(time/period-phase))                                 # Maternal antibody level placebo
+
+#Neonatal half-life (days)
+Neo.half.time <- 46.9  # Antibody half life neonates
+
+threshold <-  7      # Protection threshold literature
+```
+
+Transfer function - obtained from fitting a regression on published data (see Scheltema et al 2018)
+```{r}
+transfer.function <- function(transferratio.vector, gestational.age, ga.vector){
+
+  exp.model <- lm(log(transferratio.vector) ~ gestational.age) # regression exponential
+
+  summary.exp.model <- summary(exp.model)
+
+  coef.exp.model <- coef(exp.model) #coefficient
+
+  return(coef.exp.model)
+}
+
+coef.exp.model <- transfer.function(transferratio.vector = c(0.025, 0.1, 0.5, 0.70, 1.5), 
+                                    gestational.age = c(13, 28, 32, 36, 40),
+                                    ga.vector = c(1:42)) 
+```
+
+Ab levels mother at time t
+```{r}
+maternal.ab.level <- mab.p
+
+# change to a function of vaccination
+# if vac_status==0, meaning that the individual did not receive vaccine; all-or-nothing settings, can change based on vaccine efficacy/ vaccine response later
+## exponential function
+# maternal.ab.level <- function(vac_status,t.measure, t.vac, 
+#                               vac.effect, m.ab.base){
+#   if (vac_status==0){m.ab.level <- m.ab.base}
+#   
+#   if (vac_status==1){
+#   t.since.vac <- t.measure*7 - t.vac*7
+#   
+#   # before week 1
+#   if(t.since.vac < 7) {m.ab.level <- m.ab.base}     
+#   
+#   # exp between 1 and 2
+#   if(t.since.vac >= 7 & t.since.vac < 14) {m.ab.level <- m.ab.base * exp((t.since.vac-7)*(log(vac.effect)/7))}  #new!!! 14 days ipv 21
+#   
+#   # after week 2
+#   if(t.since.vac >= 14) {m.ab.level <- (m.ab.base * vac.effect) * (1/2)^((t.since.vac-14)/(mat.half.time))
+#   
+#   }}
+#   
+#   return(m.ab.level)
+# }
+
+```
+
+Ab levels at time of birth
+```{r}
+#### Cord blood level at time of birth ####
+cord.blood.ab.level <- function(gest.age, coef.exp.model, m.ab.base){
+
+  transfer <- exp(coef.exp.model[1] + coef.exp.model[2] * gest.age)
+
+  cb.level <- maternal.ab.level* transfer
+
+  return(cb.level)
+}
+
+# cord.blood.ab.level <- function(vac_status,gest.age, t.vac, coef.exp.model, vac.effect, m.ab.base){
+#   
+#   transfer <- exp(coef.exp.model[1] + coef.exp.model[2] * gest.age)
+#   
+#   if(vac_status==0){
+#   cb.level <- maternal.ab.level(gest.age,vac_status,t.measure, t.vac, 
+#                              vac.effect, m.ab.base) * transfer}
+#   
+#   if(vac_status==1){
+#   cb.level <- maternal.ab.level(gest.age,vac_status,t.measure, t.vac, 
+#                              vac.effect, m.ab.base) * transfer}
+#   
+#   return(cb.level)
+# }
+```
+
+Ab levels after birth
+```{r}
+#### Cord blood level at time of birth ####
+ab.level.afterbirth <- function(t, cb.level){
+
+ab.level <- cb.level * (1/2)^(t/half.time)
+
+  return(ab.level)
+}
+
+```
+
+Length of maternal protection
+
+```{r}
+#### Age at threshold####
+
+age.function <- function(cb.level,threshold, half.time){
+
+   exp.age <- ifelse((half.time * (log((threshold/cb.level),base = exp(1))))/-log(2, exp(1))>0,(half.time * (log((threshold/cb.level),base = exp(1))))/-log(2, exp(1)),0)
+  # expected age at threshold; greater than 0 or at 0
+
+   return(exp.age)
+}
+```
+
+```{r}
+cb.level = cord.blood.ab.level(gest.age = gest.age, coef.exp.model = coef.exp.model, m.ab.base = mab.p)
+
+Timeprotected <- age.function(cb.level =cb.level,threshold = threshold,half.time = Neo.half.time)
+
+plot(maternal.ab.level,type="l",ylim = c(min(cb.level),max(maternal.ab.level)))
+lines(cb.level,col="red")
+
+plot(Timeprotected,type = "l")
+```
+
 ```{r}
@@
 library(reshape2)
+library(dplyr)
+library(stringr)
 ```
@@
 In this example, the 1-4 year olds are divided into 12 month age classes as well.
-
 ```{r}
 #T is N time points; N_ages is number of age groups
-parmset <- readRDS('./data_and_parms/parmset.rds')
-Pop1 <- readRDS('./data_and_parms/pop1.rds') #initial population
-B <-  readRDS('./data_and_parms/Birth_rate.rds')
-c2 <- readRDS( './data_and_parms/c2.rds') # contact matrix
-
+parmset <- readRDS('/Users/zhezheng/Library/CloudStorage/Box-Box/aim3/RSVtransmissionmodel/RSV_metapop/SimpleModel/data_and_parms/parmset.rds')
+Pop1 <- readRDS('/Users/zhezheng/Library/CloudStorage/Box-Box/aim3/RSVtransmissionmodel/RSV_metapop/SimpleModel/data_and_parms/pop1.rds') #intial population
+NL_B <- read.csv("~/Library/CloudStorage/OneDrive-YaleUniversity/SPREAD study/SPREAD/NL_birthrate.csv")
+NL_B <- NL_B[complete.cases(NL_B),]
+NL_B$birthrate <- NL_B$birth/NL_B$Pop
+c2 <- readRDS( '/Users/zhezheng/Library/CloudStorage/Box-Box/aim3/RSVtransmissionmodel/RSV_metapop/SimpleModel/data_and_parms/c2.rds') # contact matrix
 ```
@@
 ```{r}
-###########################################
-# 1/duration of maternal immunity (DAYS)
-DurationMatImmunityDays= 112
-###########################################
-
 ###########################################
@@
 ```{r}
-source('./data_and_parms/simple_model.R') #Read in the model
+source("~/Library/CloudStorage/Box-Box/immuno-epi/Rcode/immuno_epi_model.R") #Read in the model
 ```
@@
  start_time = 1 # start date (years)
- tmax = nrow(B)
+ tmax = nrow((NL_B))
 # end_time = 25 # end date (years)
@@
 #???Should prob be estimated?
-b1=0.1830827 #Seasonal amplitude
-phi=3.327749 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
+b1=0.16 #Seasonal amplitude
+phi=3.25 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
 #########################################
@@
 # 1/duration of maternal immunity (DAYS)
-DurationMatImmunityDays= 16*7
+DurationMatImmunityDays= c(rep(0,4),rep(Timeprotected,nrow(NL_B)/length(Timeprotected)))
 ###########################################
@@
 #Matrix: T rows, N_ages columns; columns 2:N_ages all 0s
-PerCapitaBirthsYear=B 
+PerCapitaBirthsMonth=matrix(c(NL_B$birthrate,rep(0,nrow(NL_B)*(nrow(yinit.matrix)-1))),byrow = F,nrow=nrow(NL_B),ncol=nrow(yinit.matrix))
+
+PerCapitaBirthsMonth_total=rbind(do.call(rbind, replicate(6, PerCapitaBirthsMonth[7:18,], simplify=FALSE)),PerCapitaBirthsMonth[19:24,])
+PerCapitaBirthsMonth_total=rbind(PerCapitaBirthsMonth_total,PerCapitaBirthsMonth)
+PerCapitaBirthsMonth_total=rbind(PerCapitaBirthsMonth_total,PerCapitaBirthsMonth[327:328,])
 #############################################
@@
 
-baseline.txn.rate <- 9.0
+baseline.txn.rate <-9
 q=1
@@
 ```{r saveparms}
-parms<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
+parms<-list(PerCapitaBirthsMonth=PerCapitaBirthsMonth_total,
             DurationMatImmunityDays=DurationMatImmunityDays,
@@
             sigma3=sigma3,
-            time.step='month'
+            time.step='month',
+            cocooning_duration=7
             )
@@
 results <- ode(y=yinit.vector, t=my_times,  
-               func=simple_model, 
+               func=immuno_epi_model, 
             parms=parms )
@@
 ```{r}
-  St <- results[-c(1:300),-1]
+  St <- results[-c(1:220),-1]
    q=1
@@
 ```
-
-## fit the model to simulated data in order to estimate parameter sets
-```{r}
-parmset<-list(PerCapitaBirthsYear=PerCapitaBirthsYear,
-            WidthAgeClassMonth=WidthAgeClassMonth,
-            um=um,
-            rho1=rho1,
-            rho2=rho2,
-            dur.days1=dur.days1,
-            dur.days2=dur.days2,
-            dur.days3=dur.days3,
-            yinit.matrix=yinit.matrix,
-            q=q,
-            c=c2,
-            sigma1=sigma1,
-            sigma2=sigma2,
-            sigma3=sigma3,
-            time.step='month'
-            )
-
-fitmodel <-  function(parameters,dat) {   # takes the parameter values and dataset as inputs
-  protrans <- parameters[1] # estimate parameter related to R0 (baseline transmission rate)
-  amp <- parameters[2] # estimate parameter related to seasonal amplitude
-  trans <- parameters[3] # estimate parameter related to seasonal peak timing
-  DMD <- parameters[4] # estimate parameter related to the duration of maternal immunity
-  b1 <- exp(amp) #ensure positive 
-  baseline.txn.rate <- exp(protrans) #ensure positive
-  phi <-  (2*pi*(exp(trans))) / (1+exp(trans)) # transform to its scale in the model
-  durx <- exp(DMD) #ensure positive
-  durx <<- durx #somehow we need this for the function and model to recognize these parameters
-  baseline.txn.rate <<- baseline.txn.rate
-  b1 <<- b1
-  phi <<- phi
-
-   # Simulate the model with initial conditions and timesteps defined above, and parameter values from function call
-  results <- ode(y=yinit.vector, t=my_times,  
-               func=simple_model, 
-            parms=c(parmset,baseline.txn.rate=baseline.txn.rate,b1,phi,DurationMatImmunityDays=durx))
-  t0 <- 108
-  results<- tail(results,t0)
-  St <- results[,-1]
-
-   b=baseline.txn.rate/(parmset$dur.days1/30.44)
-   beta=(b/100)/(sum(yinit.matrix)^(1-1))*c2
-
-   lambda1=matrix(0,nrow=t0,ncol=al)#Force of infection
-    for (t in 1:t0)
-    {lambda1[t,] <- as.vector((1+b1*cos(2*pi*(t-phi*12)/12))*((St[t,(2*al+1):(3*al)]+rho1*St[t,(4*al+1):(5*al)]+rho2*St[t,(6*al+1):(7*al)]+rho2*St[t,(8*al+1):(9*al)])%*%beta)/sum(St[t,]))}
-
-  H1=matrix(0,nrow=t0,ncol=al)#Number of hospitalizations by age
-  for (i in 1:al){
-    H1[,i]=hosp1[i]*St[,(al+i)]*lambda1[,i]+hosp2[i]*sigma1*St[,(3*al+i)]*lambda1[,i]+hosp3[i]*sigma2*St[,(5*al+i)]*lambda1[,i]+hosp3[i]*sigma3*St[,(7*al+i)]*lambda1[,i]
-    }
-  H <- rowSums(H1)
-
-  agedist <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
-            sum(colSums(H1)[7:9]),
-            sum(colSums(H1)[10:12]),
-            colSums(H1)[13:21])/sum(H1)
-
-   LLall <- sum(dpois(x = dat, lambda =H, log = TRUE)) # fit to timeseries
-   LLmulti <- dmultinom(x= agedist,prob = agedist_Sim,log = T) # fit to age distribution
-
-   #prior
-   durprior <- dgamma(x=durx,22,5,log=T)
-
-   #total LL
-   LL <- LLall+LLmulti+durprior
-
-   return(LL)
-}
-
-fitLL <- optim(par = c(2.198522,  -1.638387 , -1.301047 ,  4.619743 ),           # starting values for beta and gamma - you should get the same result no matter which values you choose here
-      fn = fitmodel,        # the distance function to optimise
-      dat = Hosp_sim,         # the dataset to fit to ("dat" argument is passed to the function specified in fn)
-      control = list(fnscale=-1)) # the log likelihood is negative; here we minimize the negative log likelihood
-```

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/Outline_RSV_VE_maternal.Rmd
add materials

---
title: "Bias and power of demonstrating maternal vaccine efficacy in seasonal RSV epidemic"
author: "ZHE ZHENG"
date: "2022-11-29"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Objectives

* Compare the chance of success of randomized controlled trials between localities (power and efficacy).

* Explain bias introduced to vaccine RCTs by not control for time of recruitment

### Background

(1) RSV epidemics are highly seasonal.

(2) The risk of exposure to RSV varies by season.

(3) Infants are protected by transplacentally acquired antibodies from their mothers.

(4) Cord blood antibodies varies by season. XX weeks time lags from the incidence of RSV infection.

(5) Susceptibility to RSV infection varies by season.

(6) Probability of developing severe RSV disease decreases by age.
\

Generally,
$$VE (\%)=(1-(\frac{ARV}{ARU}))\times100$$
VE stands for vaccine efficacy. ARV stands for attack rate in vaccinated group. ARU stands for attack rate in unvaccinated group.

For maternal vaccines against RSV hospitalizations in infants, the vaccine efficacy was measured as:
$$VE (\%)=(1-(\frac{H_{vac}}{H_{placebo}}))\times100$$

Where $H_{vac}$ stands for the number of infants that are hospitalized due to RSV infection

### VE estimates
Depends on hospitalization incidence in vaccination groups and placebo groups. \

Hypothesis: RSV vaccination delayed the first age of infection. When the epidemic is in increasing phase, it will achieve a higher efficacy against RSV hospitalizations because the probability of developing severe lower respiratory tract illness is age dependent. In decreasing phase, there will be few RSV hospitalizations in both placebo and vaccinated group.

### Bias in VE
Factors that affect hospitalization incidence differently between vaccination groups and placebo groups.\

Hypothesis: risk of exposure during observation is not consistent between vaccinated group and placebo group
because time of recruitment is different between vaccinated and placebo group

### Power estimates
The 4 primary factors that affect the power of vaccine efficacy trials are (1) the probability of the Type I error (hold at 5%), (2) effect size (vaccine efficacy. It needs to be above 50% for RSV), (3) variability, and (4)sample size.

Hypothesis: 
Places have higher/larger:

* birth rate
* population density
* household size/number of children under 5 years old

or have lower/less:

* seasonal variations in RSV transmission
* preterm birth or other physical conditions
* lower socioeconomic status

and recruit 2nd trimester pregnant women whose delivery date is XX weeks prior to the peak of RSV epidemics 
\

are more likely to successfully demonstrate vaccine efficacy.

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/Sim_H_vac.Rmd
add materials

--- 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/Antibody_Sim.Rmd
@@
 ---
-title: "Antibody_Sim"
+title: "simulate Hospitalizations"
 author: "ZHE ZHENG"
-date: '2022-06-15'
+date: '2022-11-04'
 output: html_document
 ---
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-```
-
-```{r}
-time <- 1:(12*9) # time series data point
-period = 12 # time unit of observation (12 corresponds to months; 52 corresponds to weeks)
-AMP = 1.5 # amplitude of antibody seasonality
-phase= 1.6 # shift of peak timing at the beginning
-#Naturally accuring maternal antibdoy levels/ harmonic term to represent seasonal waves
-mab.p <- 13+AMP*cos(2*pi*(time/period-phase))                                 # Maternal antibody level placebo
-
-#Neonatal half-life (days)
-Neo.half.time <- 46.9  # Antibody half life neonates
-
-threshold <-  7      # Protection threshold literature
-```
-
-Transfer function - obtained from fitting a regression on published data (see Scheltema et al 2018)
-```{r}
-transfer.function <- function(transferratio.vector, gestational.age, ga.vector){
-
-  exp.model <- lm(log(transferratio.vector) ~ gestational.age) # regression exponential
-
-  summary.exp.model <- summary(exp.model)
-
-  coef.exp.model <- coef(exp.model) #coefficient
-
-  return(coef.exp.model)
-}
-
-coef.exp.model <- transfer.function(transferratio.vector = c(0.025, 0.1, 0.5, 0.70, 1.5), 
-                                    gestational.age = c(13, 28, 32, 36, 40),
-                                    ga.vector = c(1:42)) 
-```
-
-Ab levels mother at time t
-```{r}
-maternal.ab.level <- mab.p
-
-# change to a function of vaccination
-# if vac_status==0, meaning that the individual did not receive vaccine; all-or-nothing settings, can change based on vaccine efficacy/ vaccine response later
-## exponential function
-# maternal.ab.level <- function(vac_status,t.measure, t.vac, 
-#                               vac.effect, m.ab.base){
-#   if (vac_status==0){m.ab.level <- m.ab.base}
-#   
-#   if (vac_status==1){
-#   t.since.vac <- t.measure*7 - t.vac*7
-#   
-#   # before week 1
-#   if(t.since.vac < 7) {m.ab.level <- m.ab.base}     
-#   
-#   # exp between 1 and 2
-#   if(t.since.vac >= 7 & t.since.vac < 14) {m.ab.level <- m.ab.base * exp((t.since.vac-7)*(log(vac.effect)/7))}  #new!!! 14 days ipv 21
-#   
-#   # after week 2
-#   if(t.since.vac >= 14) {m.ab.level <- (m.ab.base * vac.effect) * (1/2)^((t.since.vac-14)/(mat.half.time))
-#   
-#   }}
-#   
-#   return(m.ab.level)
-# }
-
-```
-
-Ab levels at time of birth
-```{r}
-#### Cord blood level at time of birth ####
-cord.blood.ab.level <- function(gest.age, coef.exp.model, m.ab.base){
-
-  transfer <- exp(coef.exp.model[1] + coef.exp.model[2] * gest.age)
-
-  cb.level <- maternal.ab.level* transfer
-
-  return(cb.level)
-}
-
-# cord.blood.ab.level <- function(vac_status,gest.age, t.vac, coef.exp.model, vac.effect, m.ab.base){
-#   
-#   transfer <- exp(coef.exp.model[1] + coef.exp.model[2] * gest.age)
-#   
-#   if(vac_status==0){
-#   cb.level <- maternal.ab.level(gest.age,vac_status,t.measure, t.vac, 
-#                              vac.effect, m.ab.base) * transfer}
-#   
-#   if(vac_status==1){
-#   cb.level <- maternal.ab.level(gest.age,vac_status,t.measure, t.vac, 
-#                              vac.effect, m.ab.base) * transfer}
-#   
-#   return(cb.level)
-# }
-```
-
-Ab levels after birth
-```{r}
-#### Cord blood level at time of birth ####
-ab.level.afterbirth <- function(t, cb.level){
-
-ab.level <- cb.level * (1/2)^(t/half.time)
-
-  return(ab.level)
-}
-
-```
-
-Length of maternal protection
-
-```{r}
-#### Age at threshold####
-
-age.function <- function(cb.level,threshold, half.time){
-
-   exp.age <- ifelse((half.time * (log((threshold/cb.level),base = exp(1))))/-log(2, exp(1))>0,(half.time * (log((threshold/cb.level),base = exp(1))))/-log(2, exp(1)),0)
-  # expected age at threshold; greater than 0 or at 0
-
-   return(exp.age)
-}
-```
-
-```{r}
-cb.level = cord.blood.ab.level(gest.age = gest.age, coef.exp.model = coef.exp.model, m.ab.base = mab.p)
-
-Timeprotected <- age.function(cb.level =cb.level,threshold = threshold,half.time = Neo.half.time)
-
-plot(maternal.ab.level,type="l",ylim = c(min(cb.level),max(maternal.ab.level)))
-lines(cb.level,col="red")
-
-plot(Timeprotected,type = "l")
-```
 
@@
 ```
+
 Set names of Age groups 
@@
 #???Should prob be estimated?
-b1=0.16 #Seasonal amplitude
-phi=3.25 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
+b1=0.2 #Seasonal amplitude
+phi=3.65 #Seasonal phase shift (weeks;[-26,26] 0=peak @ Jan 1)
 #########################################
@@
 # 1/duration of maternal immunity (DAYS)
-DurationMatImmunityDays= c(rep(0,4),rep(Timeprotected,nrow(NL_B)/length(Timeprotected)))
+DurationMatVac= 60
+DurationNaturalProtection= 30
 ###########################################
@@
 
-##To ESTIMTE from DATA?: Baseline transmission rate
+#############################################
+# frequency dependent or density dependent transmission
+q=1
+
+#############################################
+#contact matrix
+c2=c2
+
+##Baseline transmission rate determine the force of transmission
 #############################################
@@
 baseline.txn.rate <-9
-q=1
-c2=c2
 
@@
 ```{r saveparms}
-parms<-list(PerCapitaBirthsMonth=PerCapitaBirthsMonth_total,
-            DurationMatImmunityDays=DurationMatImmunityDays,
+parms<-list(PerCapitaBirthsYear=PerCapitaBirthsMonth_total*12,
+            DurationMatVac=DurationMatVac,
             WidthAgeClassMonth=WidthAgeClassMonth,
@@
             time.step='month',
-            cocooning_duration=7
+            DurationNaturalProtection=DurationNaturalProtection
             )
@@
             parms=parms )
-
 ```
 
-## Extract the relevant results
-```{r}
-
-#Pitzer used a 40-50 YEAR burn in period
-burnN <- 20*12
-results.burned <- results[-c(1:burnN),]
-
-##Any infected person
-infected.cols <- results.burned[,c(grep('I1', colnames(results.burned)),grep('I2', colnames(results.burned)),grep('I3', colnames(results.burned)),grep('I4', colnames(results.burned)  )) ]
-
-infected.all <- apply(infected.cols,1, sum)
-
-infected.cols.m <- melt(infected.cols)
-infected.cols.m$agegrp <- sub(" .*", "",infected.cols.m$Var2 )
-infected.cols.c <- dcast(infected.cols.m, Var1~agegrp, fun.aggregate = sum)
-
-#Pop
-all.m <- melt(results.burned[,grep('Agegrp',colnames(results.burned))])
-all.m$agegrp <- sub(" .*", "",all.m$Var2 )
-all.c <- dcast(all.m, Var1~agegrp, fun.aggregate = sum)
-pop.all <- apply(all.c[,-1],1,sum)
-
-plot(infected.all, type='l')
-plot(pop.all,type="l")
-
-```
+## Get hospitalizations based on FOI
 ```{r}
@@
 
+Hosp_under5 <- H1[(nrow(H1)-11):nrow(H1),1:16]
+
+# number of hospitalizations depends on the season of observation 
+# (birth to 90 days)
+Hosp_vac_90 <- Hosp_90(t=101,H1)
+# number of hospitalizations depends on the season of observation 
+# (birth to 180 days)
+#Hosp_vac_180 <- Hosp_180(t=90,H1)
+
+# Vaccine efficacy 1-Hosp_mat_vac/Hosp_no_vac
+1-Hosp_vac_90/Hosp_novac_90
 ```
 
+## calculate the age of hospitalizations (age unit: month)
 ```{r}
-plot(1:108,rowSums(H1),type="l",xlab="months",ylab="simulated RSV hospitalizations")
-# need to get rid of the burn-in period
+age_midpoint <- c(seq(1/24,23/24,1/12),1.5,2.5,3.5,4.5)*12
+time.step="month"
 
-barplot(c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
-            sum(colSums(H1)[7:9]),
-            sum(colSums(H1)[10:12]),
-            colSums(H1)[13:21])/sum(H1),xlab="age group",ylab="Number of Hospitalizations")
-
-# simulated age distribution Age groups: The <12 month olds, were divided into per 3-months age classes.
-The 1-4 year olds are  divided into 1yearly age classes, then 5–9 years, 10–19 years, 20–39 years, 40–59-years, and 60+ years old
-
-Hosp_sim <- as.integer(rowSums(H1))
-agedist_Sim <- c(sum(colSums(H1)[1:3]),sum(colSums(H1)[4:6]),
-            sum(colSums(H1)[7:9]),
-            sum(colSums(H1)[10:12]),
-            colSums(H1)[13:21])/sum(H1)
+average_age_H(age_midpoint,Hosp_under5,time.step)
 ```

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/antibody level.Rmd
add materials

---
title: "Antibodylevel"
author: "Joukje Willemsen"
date: '2022-06-07'
output: html_document
---
Parametersettings

```{r}
library(ggplot2)

#Naturally accuring maternal antibdoy levels 
mab.p <- 13                                              # Maternal antibody level placebo
# set neutralizing antibody level at 8; seasonal variations in maternal immunity as seem in "Seasonal Variation of Maternally Derived Respiratory Syncytial Virus Antibodies and Association with Infant Hospitalizations for Respiratory Syncytial Virus"

#Fold increase 14 days after vaccination
factor.v_LME <- 12.17 #fold increase PCA

#Maternal half-life (days)
mat.half.time <- 145.58

#Neonatal half-life (days)
half.time <- 46.9                                  # Antibody half life neonates

threshold <- 40                                          # Protection threshold literature

#Matrix with potential time of vaccination
t.vac <- c(22:42)                                        # Weeks of vaccination vector

```

Transfer function - obtained from fitting a regression on published data (see Scheltema et al 2018)
```{r}
transfer.function <- function(transfer.vector, gestational.age, ga.vector){

  exp.model <- lm(log(transfer.vector) ~ gestational.age) # regression exponential

  summary.exp.model <- summary(exp.model)

  coef.exp.model <- coef(exp.model) #coefficient

  return(coef.exp.model)
}

coef.exp.model <- transfer.function(transfer.vector = c(0.025, 0.1, 0.5, 0.70, 1.5), 
                                    gestational.age = c(13, 28, 32, 36, 40),
                                    ga.vector = c(1:42)) 
```

Ab levels mother at time t
```{r}
## exponential function
maternal.ab.level <- function(t.measure, t.vac, 
                              vac.effect = factor.v_LME, m.ab.base = mab.p){

  t.since.vac <- t.measure*7 - t.vac*7

  # before week 1
  if(t.since.vac < 7) {m.ab.level <- m.ab.base}

  # exp between 1 and 2
  if(t.since.vac >= 7 & t.since.vac < 14) {m.ab.level <- m.ab.base * exp((t.since.vac-7)*(log(vac.effect)/7))}  #new!!! 14 days ipv 21

  # after week 2
  if(t.since.vac >= 14) {m.ab.level <- (m.ab.base * vac.effect) * (1/2)^((t.since.vac-14)/(mat.half.time))

  }

  return(m.ab.level)
}

```

Ab levels at time of birth
```{r}
#### Cord blood level at time of birth ####
cord.blood.ab.level <- function(gest.age, t.vac, coef.exp.model = coef.exp.model, vac.effect = factor.v_LME, m.ab.base = mab.p){

  transfer <- exp(coef.exp.model[1] + coef.exp.model[2] * gest.age)

  cb.level <- maternal.ab.level(gest.age, t.vac) * transfer

  return(cb.level)
}

```

Ab levels after birth
```{r}
#### Cord blood level at time of birth ####
ab.level.afterbirth <- function(t, cb.level){

ab.level <- cb.level * (1/2)^(t/half.time)

  return(ab.level)
}

```

Length protection window

```{r}
#### Age at threshold - by vaccination time ####

age.function <- function(cb.level, half.time = half.time){

#   cb.level <- cord.blood.ab.level(gest.age, t.vac, coef.exp.model) #cordblood level

   exp.age <- (half.time * (log((threshold/cb.level),base = exp(1))))/-log(2, exp(1))
  # expected age at threshold

   if(exp.age <= 0){exp.age <- 0} # if expected age is <= 0

   return(exp.age)
}

#age.function(40, half.time = half.time)
```

```{r}
Antibodysplot <- function(t.vac, gest.age){

Timeprotected <- age.function(cb.level = cord.blood.ab.level(gest.age = gest.age, t.vac = t.vac, coef.exp.model = coef.exp.model, vac.effect = factor.v_LME, m.ab.base = mab.p), half.time = half.time)

Ma <- c(MaternalAblevels=sapply(c(20:gest.age), FUN = maternal.ab.level, t.vac = t.vac), rep(NA, Timeprotected+2))

Faprebirth <- c(sapply(c(20:gest.age), FUN = cord.blood.ab.level, t.vac = t.vac, coef.exp.model = coef.exp.model, vac.effect = factor.v_LME, m.ab.base = mab.p), rep(NA,Timeprotected+2))

FAafterbirth <- c(rep(NA,gest.age-20), sapply(c(0:(Timeprotected+2)), FUN = ab.level.afterbirth, cb.level = cord.blood.ab.level(gest.age = gest.age, t.vac = t.vac, coef.exp.model = coef.exp.model, vac.effect = factor.v_LME, m.ab.base = mab.p)))

df <- data.frame(week=c(20:(gest.age+(Timeprotected+2))), Ma, Faprebirth)

plot <- ggplot() +
  geom_rect(aes(fill = 'Protection Window'), xmin=gest.age,
                  xmax = gest.age+Timeprotected,
                  ymin = -Inf,
                  ymax = 40, alpha = .125/2) +
  geom_line(aes(x = week, y= FAafterbirth, color = "Fetus", linetype = "After birth"), data = df)+
  geom_line(aes(x=week, y=Ma, color = "Maternal", linetype = "Before birth"), data = df)+
  geom_line(aes(x=week, y= Faprebirth, color = "Fetus", linetype = "Before birth"), data = df)+
  geom_vline(xintercept = gest.age, linetype = "dashed")+
  geom_vline(xintercept = t.vac, linetype = "dashed")+
  annotate(geom = "text",
             label = "Tbirth",
             x = gest.age,
             y = 50,
             angle = 360-90, 
             vjust = -1) +
   annotate(geom = "text",
             label = "Tvac",
             x = t.vac,
             y = 50,
             angle = 360-90, 
             vjust = 1) +
  ggtitle("Maternal and fetal antibodylevels over time") +
  xlab("Time since start pregnancy (week)") + ylab("PCA levels") +
  theme_classic() + 
  scale_fill_manual(labels = paste(as.character(round(Timeprotected),0), "days"), breaks = "Protection Window", values = "red",
                    guide = guide_legend(title = "Protection Window",
                                         override.aes = list(alpha = .5/4)))+
 scale_x_continuous(breaks = seq(min(df$week),max(df$week),5))

return(plot)

}

Antibodysplot(24, 38)

```

Step 1: Get a matrix of cordblood Ab levels
```{r}

  gest_age_seq <- seq(24, 44, 1)

  cols <- c("LINE1"="#f04546")

  #Make an empty matrix
  predicted_ab.matrix <- matrix(NA, nrow = length(gest_age_seq), ncol = length(t.vac))

  #Make an empty matrix
  prevented.matrix <- matrix(NA, nrow = length(gest_age_seq), ncol = length(t.vac))

  for(ga in 1:length(t.vac)){
  for(case in 1:length(gest_age_seq)){
  predicted_ab.matrix[case, ga] <-cord.blood.ab.level(gest_age_seq[case], t.vac[ga], coef.exp.model=coef.exp.model, vac.effect=factor.v_LME, m.ab.base = mab.p)}}

rownames(predicted_ab.matrix) <- gest_age_seq
colnames(predicted_ab.matrix) <- t.vac

library(reshape2)
melted_cormat <- melt(predicted_ab.matrix)
head(melted_cormat)

library(ggplot2)
library(tidyverse)
plotProtWindow <- ggplot(data = melted_cormat %>% filter(value >= 1), aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 0))) +
  scale_fill_gradient(low = "white", high = "red", limit=c(40,400))+
  theme_classic()+
  xlab("Gestational age at time of birth") +
  ylab("Gestational age at time of vaccination") +
  scale_x_continuous(breaks = seq(24, 44, 1)) +
  scale_y_continuous(breaks = t.vac)+ 
  guides(fill=guide_legend(title="PCA levels at birth"))

plotProtWindow
```

Step 2: Protetion window per time of vaccination
```{r}
predicted_age.matrix <- apply(predicted_ab.matrix, c(1,2), age.function, half.time = half.time)
```

Get a matrix of protection window per time of vaccination x gestational age
```{r}
library(reshape2)
melted_cormat <- melt(predicted_age.matrix)
head(melted_cormat)

library(ggplot2)
plotProtWindow <- ggplot(data = melted_cormat %>% filter(value >= 1), aes(x=Var1, y=Var2, fill=value)) + 
  geom_tile(aes(fill = value)) + 
  geom_text(aes(label = round(value, 0))) +
  scale_fill_gradient(low = "white", high = "red", limit=c(1,151))+
  theme_classic()+
  xlab("Gestational age at time of birth") +
  ylab("Gestational age at time of vaccination") +
  scale_x_continuous(breaks = seq(24, 44, 1)) +
  scale_y_continuous(breaks = t.vac)+ 
  guides(fill=guide_legend(title="Number of protected days"))

plotProtWindow
```

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/sim_H_Placebo.Rmd
add materials

--- 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/Sim_H_vac.Rmd
@@
 
-**NOTE: time step here is in months--need to adjust seasonality accordingly**
-```{r}
- start_time = 1 # start date (years)
- tmax = nrow((NL_B))
-# end_time = 25 # end date (years)
-my_times <- seq(start_time, tmax, by = 1) # gives a sequence from start to end
-											 # in increments of 1
-```
-
 ##Settings parameters
@@
 # 1/duration of maternal immunity (DAYS)
-DurationMatVac= 60
+DurationMatVac= 0
 DurationNaturalProtection= 30
@@
 PerCapitaBirthsMonth_total=rbind(PerCapitaBirthsMonth_total,PerCapitaBirthsMonth[327:328,])
+
+PerCapitaBirthsYear_total=PerCapitaBirthsMonth_total*12
 #############################################
@@
 
+**NOTE: time step here is in weeks--need to adjust seasonality accordingly**
+```{r}
+ start_time = 1 # start date (years)
+ tmax = nrow((NL_B))
+# end_time = 25 # end date (years)
+my_times <- seq(start_time, tmax, by = 1) # gives a sequence from start to end
+											 # in increments of 1
+```
+
 #Save parameters in a list
 ```{r saveparms}
-parms<-list(PerCapitaBirthsYear=PerCapitaBirthsMonth_total*12,
+parms<-list(PerCapitaBirthsYear=PerCapitaBirthsYear_total,
             DurationMatVac=DurationMatVac,
@@
 # (birth to 90 days)
-Hosp_vac_90 <- Hosp_90(t=101,H1)
+Hosp_novac_90 <- Hosp_90(t=100,H1)
 # number of hospitalizations depends on the season of observation 
 # (birth to 180 days)
-#Hosp_vac_180 <- Hosp_180(t=90,H1)
+#Hosp_novac_180 <- Hosp_180(t=90,H1)
 
 # Vaccine efficacy 1-Hosp_mat_vac/Hosp_no_vac
-1-Hosp_vac_90/Hosp_novac_90
 ```

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/AgeatH.R

@@ -1,0 +1,19 @@
+### function to calculate the average age of hospitalization in children under 5
+
+average_age_H <- function(age_midpoint,Hosp_under5,time.step){
+  
+  if(parms$time.step=='month'){
+    period=12
+  }else if(parms$time.step=='week'){
+    period=52
+  }
+  
+  #multiply age midpoint with hospitalization matrix
+  age_H_matrix <- sweep(Hosp_under5, MARGIN=2, age_midpoint, '*')
+  
+  # average age of hospitalization by year 
+  average_H <- sum(age_H_matrix[1:period,])/sum(Hosp_under5[1:period,])
+
+  
+  return(average_H)
+}

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/NumberH_birthcohort.R

@@ -1,0 +1,12 @@
+# function to calculate hospitalizations in a birth cohort to 90 days
+# number of hospitalizations depends on the season of observation
+Hosp_90 <- function(t,H1){ 
+  Hosp <- round(H1[t,1]+H1[t+1,2]+H1[t+2,3])
+  return(Hosp)
+  }
+
+# function to caculate hospitalizations in a birth cohort to 180 days
+Hosp_180 <- function(t,H1){ 
+  Hosp <- round(H1[t,1]+H1[t+1,2]+H1[t+2,3]+H1[t+3,4]+H1[t+4,5]+H1[t+5,6])
+  return(Hosp)
+}

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/PICU time series.R

@@ -1,0 +1,36 @@
+library(readxl)
+PICU_UK <- read_excel("~/Library/CloudStorage/Box-Box/immuno-epi/PICU data/PICU_UK.xlsx")
+library(lubridate)
+
+PICU_UK$month <- floor_date(PICU_UK$Date_PICU_admission, "month")
+PICU_UK$RSV <- 1
+
+library(dplyr)
+RSV_series <- PICU_UK %>% 
+  group_by(month) %>% 
+  summarise(case=sum(RSV))
+RSV_series <- RSV_series[order(RSV_series$month),]
+
+plot(RSV_series$month,RSV_series$case,type="l")
+
+PICU_NL <- read_excel("~/Library/CloudStorage/Box-Box/immuno-epi/PICU data/PICU2.xlsx")
+PICU_NL$month <- floor_date(PICU_NL$Date_PICU_admission_PICU, "month")
+PICU_NL$RSV <- 1
+
+RSV_series_NL <- PICU_NL %>% 
+  group_by(month) %>% 
+  summarise(case=sum(RSV))
+RSV_series_NL <- RSV_series_NL[order(RSV_series_NL$month),]
+
+plot(RSV_series_NL$month,RSV_series_NL$case,type="l")
+
+## PICU cases are a bit sparse
+
+## PICU age distribution UK
+PICU_UK$age_at_ICU <- floor(PICU_UK$Age_PICUadmission_days/30)
+hist(PICU_UK$age_at_ICU,breaks = 13,probability = T)
+
+## PICU age distribution NL
+PICU_NL$age_at_ICU <- floor(PICU_NL$Age_PICUadmission_days_PICU/30)
+hist(PICU_NL$age_at_ICU,breaks = 13,probability = T)
+

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/immuno_epi_model.R

@@ -1,0 +1,124 @@
+immuno_epi_model <- function(t,y,parms, time.step='month'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/((parms$DurationMatVac+parms$DurationNaturalProtection)/length.step)
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  M <-  States[,'M']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
+  S3 <-  States[,'S3']
+  I4 <-  States[,'I4']
+  
+  N.ages <- length(M)
+  
+  ###Check the standardization of beta and overall structure of lambda here
+  #how does'baseline txn rate' figure in here?
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/burnin.R

@@ -1,0 +1,123 @@
+burnin_quasi_equilibrium <- function(t,y,parms, time.step='week'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  M <-  States[,'M']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
+  S3 <-  States[,'S3']
+  I4 <-  States[,'I4']
+  
+  N.ages <- length(M)
+  
+  ###Check the standardization of beta and overall structure of lambda here
+  #how does'baseline txn rate' figure in here?
+  ##???##################
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/immuno_epi_model_original.R

@@ -1,0 +1,124 @@
+immuno_epi_model <- function(t,y,parms, time.step='week'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/((parms$DurationMatVac+parms$DurationNaturalProtection)/length.step)
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  M <-  States[,'M']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
+  S3 <-  States[,'S3']
+  I4 <-  States[,'I4']
+  
+  N.ages <- length(M)
+  
+  ###Check the standardization of beta and overall structure of lambda here
+  #how does'baseline txn rate' figure in here?
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/immuno_epi_model_revised.R

@@ -1,0 +1,158 @@
+immuno_epi_model <- function(t,y,parms, time.step='week'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+  v=parms$cover # indicator function for when the vaccine will be administered * vaccine coverage
+  RR=parms$RR # relative risk of infection after vaccination; equal to 1 before vaccine introduction
+  waningV=1/(parms$DurationVE/length.step)
+  rrM=1
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  M <-  States[,'M']
+  Mv <-  States[,'Mv']
+  
+  Sv <-  States[,'Sv']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/placebo.R

@@ -1,0 +1,113 @@
+placebo_cohort <- function(t,y,parms, time.step='week'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+ 
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #??Is this right?? Yes, gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  M <-  States[,'M']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
+  S3 <-  States[,'S3']
+  I4 <-  States[,'I4']
+  
+  N.ages <- length(M)
+  
+  ##fixed lambda based on lambda from burn-in period##################
+  lambda <- parms$lambda[t,]
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/server.R

@@ -1,0 +1,404 @@
+library("shiny")
+library("deSolve")
+library("cowplot")
+library("ggplot2")
+library("tidyverse")
+library("ggrepel")
+library("shinydashboard")
+library("tibble")
+# get all prespecified parameters. This is only for 
+# simulation purpose (mostly from US, NL and Kenya info). 
+# Need to take a futher look at it and
+# decide whether all parameters are appropriate.
+parms_sim <- readRDS("./data_and_parms/parms_sim.rds")
+
+source("./burnin.R")  # R script for burn-in period
+source("./placebo.R") # R script for placebo cohort
+source("./vaccine.R") # R script for vaccine cohort
+al = nrow(parms_sim$yinit.matrix)
+states <- ncol(parms_sim$yinit.matrix)
+hosp1=c(.18*rep(.40,3),0.08*rep(.39,3),0.07*rep(.21,3),0.06*rep(.20,3),0.06*0.16,0.05*rep(.14,3),0.02*rep(0.05,al-16))#proportion of first LRIs that are hospitalized
+hosp2=.4*hosp1#proportion of second LRIs that are hospitalized
+hosp3=c(rep(0,al-2),0.00001,0.00004)#proportion of third LRIs that are hospitalized
+
+# Define server 
+#
+server <- function(input, output) {
+  
+  sim_mat_placebo_vaccine <- reactive({ 
+    
+  annual_birth_rate<- input$annual_birth_rate
+  yinit_matrix=as.array(parms_sim$yinit.matrix)
+  yinit_vector <- as.vector(parms_sim$yinit.matrix) #Vectorize the ynit matrix
+  # Create array that has the labels by age, State and use this to name the yinit.vector
+  name.array <- array(NA, dim=dim(yinit_matrix))
+  for(i in 1:dim(name.array)[1]){
+    for(j in 1:dim(name.array)[2]){
+      name.array[i,j] <- paste(dimnames(yinit_matrix)[[1]][i],dimnames(yinit_matrix)[[2]][j]  )
+    }
+  }
+  
+  name.vector <- as.vector(name.array)
+  names(yinit_vector) <- name.vector
+  
+  tmax=input$t_obs+24*52 # burn-in period as 24 years
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/ui.R

@@ -1,0 +1,72 @@
+library("shiny")
+library("deSolve")
+library("cowplot")
+library("ggplot2")
+library("tidyverse")
+library("ggrepel")
+library("shinydashboard")
+library("shinyWidgets")
+library("plotly")
+
+ui <- fluidPage(
+  titlePanel("Modeling maternal vaccination against RSV hospitalizations in infants"),
+  hr(),
+  p(div(HTML("Disclaimer: This simulation is for research and educational purposes only and is not intended to be a tool for epidemic prediction. 
+  There are many uncertainties and debates about the details of RSV infection and transmission. This model has limitations on capturing the constantly changing non-pharmaceutical interventions against COVID-19.
+             This work is licensed under a <a href=https://creativecommons.org/licenses/by-sa/4.0/> Creative Commons Attribution-ShareAlike 4.0 International (CC BY-SA 4.0) License </a>"))),
+  sidebarLayout(
+    sidebarPanel(
+    #  p(div(HTML("<em>Evaluate the impact of various parameters...</em>"))),
+      column(width=6,
+                 sliderInput("baseline.txn.rate",
+                         "Approximate R0:",
+                         min = 9, max = 10.5, value = 9, step=0.1
+             ),
+             sliderInput("annual_birth_rate",
+                         "Annual birth rate:",
+                         min =  0.0080, max = 0.0160,value = 0.0094,  step=0.0002
+             ),
+             sliderInput("phase",
+                         "Peak timing of epidemics:",
+                         0, 2*pi, 3.62, step=0.01
+             ),
+             sliderInput("b1",
+                         "Amplitude of transmission seasonality:",
+                         0.1, 0.3, 0.2, step=0.01
+             ),
+             sliderInput("time_enroll",
+                           "Time of enrollment:",
+                           1, 12,1, step=1
+             )),  
+      column(width=6,
+             sliderInput("DurationMatVac",
+                         "Duration of protection of maternal vaccination:",
+                         90, 180, 90, step=30
+             ), 
+             sliderInput("DurationNaturalProtection",
+                         "Duration of protection of naturally acquired maternal immunity:",
+                          20, 150, 30,step=1
+             ), 
+             sliderInput("t_obs",
+                         "Period of observation (weeks):",
+                         26, 52*2, 27, step=1
+             )),
+    width=4
+  ),
+    
+    mainPanel(navbarPage("Output:",
+                         
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/shiny/vaccine.R

@@ -1,0 +1,143 @@
+maternalvac_cohort <- function(t,y,parms, time.step='week'){
+  
+  yinit.matrix <- parms$yinit.matrix
+  States<-array(y, dim=dim(parms$yinit.matrix))
+  dimnames(States) <- dimnames(parms$yinit.matrix)
+  
+  if(parms$time.step=='month'){
+    period=12
+    length.step=30.44 #days
+  }else if(parms$time.step=='week'){
+    period=52.1775
+    length.step=7 #days
+  }
+  
+  omega = 1/(parms$DurationMatImmunityDays/length.step)
+  RR=parms$RR # relative risk of infection after vaccination in infants
+  waningV=1/(parms$DurationVE/length.step)
+  rrM=1 # relative risk of infection after vaccination in pregnant women
+  
+  mu= 1/parms$WidthAgeClassMonth
+  if(parms$time.step=='week'){
+    mu= 1/(parms$WidthAgeClassMonth*4.345)
+  }
+  
+  gamma1= 1/(parms$dur.days1/length.step)  #converts 1/days to 1/lenth.step
+  gamma2= 1/(parms$dur.days2/length.step)  
+  gamma3= 1/(parms$dur.days3/length.step)  
+  gamma4= gamma3  #gamma4=gamma3 stands for rate of recovery from subsequent infection
+  
+  rho1=parms$rho1#Relative infectiousness for 2nd infections
+  rho2=parms$rho2 #Relative infectiousness for subsequent infections
+  
+  sigma1=parms$sigma1#Relaive risk of infection following 1st infections
+  sigma2=parms$sigma2#Relaive risk of infection following 2nd infections
+  sigma3=parms$sigma3#Relaive risk of infection following subsequent infections
+  
+  #Pull out the states  for the model as vectors
+  Mv <-  States[,'Mv']
+  
+  Sv <-  States[,'Sv']
+  S0 <-  States[,'S0']
+  I1 <-  States[,'I1']
+  
+  S1 <-  States[,'S1']
+  I2 <-  States[,'I2']
+  
+  S2 <-  States[,'S2']
+  I3 <-  States[,'I3']
+  
+  S3 <-  States[,'S3']
+  I4 <-  States[,'I4']
+  
[…]

### 2023-01-26 5_roadmaps_future_projects/RSV_maternal_vac_eff/stochmsis.R

@@ -1,0 +1,151 @@
+### stochastic MSIS model to record flows of individuals
+### start from no age structure
+
+stochmsis = function(fixed,parms){ ## set =1 for fixed infectious period, =0 for exponentially distributed
+  ### input parameter
+  lambda=parms$fixed_lambda 
+  # transmission rate in general population at each time point
+  ## time-varying transmission rate derived from deterministic model in quasi-equilibrium
+  
+  birth=parms$birth
+  # instead of birth[t] = rbinom(1,N[t-1],birthrate[t-1])
+  # we treat birth number as an input over time to simulate a birth cohort
+  # for the clinical trial.
+  # check whether this is correct.
+  
+  #########################################
+  # duration of infectiousness 
+  #Duration in days
+  dur.days1 <- 10 #days; first time infection
+  dur.days2 <- 7 #days; second time infection
+  dur.days3 <- 5 #days; third time and subsequent infection
+  ###########################################  
+  
+  ###########################################
+  # 1/duration of maternal immunity (DAYS)
+  DurationNaturalProtection= 30
+  ###########################################
+  
+  #############################################
+  #Relaive risk of infection following 1st, 2nd, 3rd+ infections
+  sigma1=0.76
+  sigma2=0.6
+  sigma3=0.4
+  #############################################
+  
+  # um= -0.0002227 ## death rate
+  # assume no death for the first try. Do we need death in the model?
+  # for the whole clinical trial period, deaths should be negligible
+  
+  tmax = 365 ### max number of days to simulate
+  dt = 1 ## time step for approximate method 1 day instead of 1/20 day
+  
+  N = M = S0 = I1 = S1 = I2 = S2 = I3 = S3 = I4 = rep(0,tmax)
+  # set up a vector to store the number of individuals in each compartment 
+  
+  Waning = Infect1 = Infect2 = Infect3 = Infect4 = 
+    Recover1 =Recover2 = Recover3 = Recover4 = rep(0,tmax)
[…]

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update

@@
 ---
-title: "Estimate RSV disease burden using hierarchical baysian regression"
+title: "Estimate RSV disease burden using Hierarchical Bayesian regression"
 author: "ZHE ZHENG"
@@
 ## Outline
+
+- Section 1: Background knowledge of estimating disease burden (hospitalizations/mortality etc) with statistical models.
+
+- Section 2: Background knowledge of Hierarchical Bayesian regression.
+
+- Section 3: R code for running the Hierarchical Bayesian regression to estimate in older adults by age and risk groups

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
delte output html

--- 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
@@
 ---
-title: "Estimate RSV disease burden using hierarchical baysian regression"
-author: "ZHE ZHENG"
-date: "2023-01-26"
-output: html_document
+title: "Bayesian model in JAGS"
+subtitle: "Tutorial 3 for transition"
+author: "Zhe Zheng (Gigi)^[zhe.zheng@yale.edu]"
+date: "2023-01-16"
+fontsize: 10pt
+output: 
+  beamer_presentation:
+  theme: CambridgeUS
+  colortheme: "beaver"
+  fonttheme: "structurebold"
 ---
@@
 ## Outline
+
+- Section 1: Background knowledge of estimating disease burden (hospitalizations/mortality etc) with statistical models.
+\
+\
+- Section 2: Background knowledge of Hierarchical Bayesian regression.
+\
+\
+- Section 3: R code for running the Hierarchical Bayesian regression to estimate in older adults by age and risk groups.

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
UPDATE

@@
 \
-- Section 3: R code for running the Hierarchical Bayesian regression to estimate in older adults by age and risk groups.
+- Section 3: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations.
+\
+\
+- Section 4: R code for running the Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups.
+
+## Section 1: Background knowledge of estimating disease burden with statistical models.
+
+There are 5 main statistical methods to estimate the hospitalization/mortality attributable to respiratory virus infection.
+In Dan's class, Public Health Surveillance, he teaches all four methods in details.
+Here, we will only give a brief introduction to each of the method and provide the link to the initial publications.
+
+1. **Serfling regression**
+
+2. **Periseason differences**
+
+3. **Poisson regression with log link**
+
+4. **Negative binomial regression with identity link**
+
+5. **Box-Jenkins transfer function**
+
+## Initial publications of the method
+
+1. *Serfling regression*
+  - [Methods for Current Statistical Analysis of Excess Pneumonia-influenza Deaths](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1915276/pdf/pubhealthreporig00078-0040.pdf)
+  - [The impact of influenza epidemics on mortality: introducing a severity index](https://ajph.aphapublications.org/doi/epdf/10.2105/AJPH.87.12.1944)
+  - [Impact of Influenza Vaccination on Seasonal Mortality in the US Elderly Population](https://jamanetwork.com/journals/jamainternalmedicine/article-abstract/486407)
+
+2. *Periseason differences*
+  - [The effect of influenza on hospitalizations, outpatient visits, and courses of antibiotics in children](https://www.nejm.org/doi/full/10.1056/nejm200001273420401)
+  - [Respiratory illness associated with influenza and respiratory syncytial virus infection](https://adc.bmj.com/content/archdischild/90/7/741.full.pdf)
+  - [Influenza and the Rates of Hospitalization for Respiratory Disease among Infants and Young Children](https://www.nejm.org/doi/full/10.1056/nejm200001273420402)
+
+## Initial publications of the method (continued)
+
+3. *Poisson regression with log link*
+  - ()
+
+4. *Negative binomial regression with identity link*
+  - ()
+
+5. *Box-Jenkins transfer function*
+  - ()
+
+## Other useful references
+
+- []
+
+- [The Need for Validation of Statistical Methods for Estimating Respiratory Virus–Attributable Hospitalization](https://academic.oup.com/aje/article/170/7/925/91526)

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
constant update

@@
 Here, we will only give a brief introduction to each of the method and provide the link to the initial publications.
-
+\
+\
 1. **Serfling regression**
@@
 
+## Brief introduction to 5 statistical methods for estimating respiratory virus–attributable hospitalization/mortality
+
+1. **Serfling regression** \
+  This type of model identifies the epidemic season and establish a epidemic threshold (seasonal baseline) using historical data.
+After predicting the expected diseases baseline, observed diseases above the baseline (epidemic threshold) during the epidemic season will be attributable to the viral infection.
+The basic Serfling regression is given by: $Y_t = \mu+ bt+\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ where $Y_t$ is the expected hospitalizations at time t.
+$\mu$ is the baseline.
+$bt$ captures time trend.
+$\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ are for the seasonal variations.
+Serfling regression is a linear model initially using ordinary least square to fit.
+[\color{blue} Click to see an R code example](https://kmcconeghy.github.io/flumodelr/articles/04-serfling.html)
+  \
+  \
+2. **Periseason differences**
+
 ## Initial publications of the method
 
-1. *Serfling regression*
+1. **Serfling regression**
   - [Methods for Current Statistical Analysis of Excess Pneumonia-influenza Deaths](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC1915276/pdf/pubhealthreporig00078-0040.pdf)
@@
 
-2. *Periseason differences*
+2. **Periseason differences**
   - [The effect of influenza on hospitalizations, outpatient visits, and courses of antibiotics in children](https://www.nejm.org/doi/full/10.1056/nejm200001273420401)
@@
   - [Influenza and the Rates of Hospitalization for Respiratory Disease among Infants and Young Children](https://www.nejm.org/doi/full/10.1056/nejm200001273420402)
+  - [Impact of influenza and respiratory syncytial virus on mortality in England and Wales from January 1975 to December 1990](https://www.cambridge.org/core/services/aop-cambridge-core/content/view/28087EC2F43E75D1907F15CA3411FE6D/S0950268800058957a.pdf/impact-of-influenza-and-respiratory-syncytial-virus-on-mortality-in-england-and-wales-from-january-1975-to-december-1990.pdf)
 
@@
 
-3. *Poisson regression with log link*
-  - ()
+3. **Poisson regression with log link**
+  - [Mortality Associated With Influenza and Respiratory Syncytial Virus in the United States](https://jamanetwork.com/journals/jama/article-abstract/195750)
+  - [The Association of Respiratory Syncytial Virus Infection and Influenza with Emergency Admissions for Respiratory Disease in London: An Analysis of Routine Surveillance Data](https://academic.oup.com/cid/article/42/5/640/316492)
 
-4. *Negative binomial regression with identity link*
-  - ()
+4. **Negative binomial regression with identity link**
+  - [Modelling the unidentified mortality burden from thirteen infectious pathogenic microorganisms in infants](https://www.cambridge.org/core/journals/epidemiology-and-infection/article/modelling-the-unidentified-mortality-burden-from-thirteen-infectious-pathogenic-microorganisms-in-infants/26AE7DA522FD6DCE6CE3E4F2D3CB9A79)
+  - [Hospitalization Attributable to Influenza and Other Viral Respiratory Illnesses in Canadian Children](https://journals.lww.com/pidj/fulltext/2006/09000/Hospitalization_Attributable_to_Influenza_and.8.aspx)
 
-5. *Box-Jenkins transfer function*
-  - ()
+5. **ARIMA model**
+  - [Community influenza outbreaks and emergency department ambulance diversion](https://www.sciencedirect.com/science/article/pii/S0196064403013167)
+  - [Time-Series Analysis of the Relation between Influenza Virus and Hospital Admissions of the Elderly in Ontario, Canada, for Pneumonia, Chronic Lung Disease, and Congestive Heart Failure](https://academic.oup.com/aje/article/149/1/85/205393)
 

### 2023-01-26 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
updates

@@
 \
+
 1. **Serfling regression**
@@
 
-1. **Serfling regression** \
+1.**Serfling regression** \
   This type of model identifies the epidemic season and establish a epidemic threshold (seasonal baseline) using historical data.
@@
 $bt$ captures time trend.
-$\sum \alpha_i cos \theta+ \sum \beta_i sin \theta$ are for the seasonal variations.
+$\sum \alpha_i \cos \theta+ \sum \beta_i \sin \theta$ are for the seasonal variations.
 Serfling regression is a linear model initially using ordinary least square to fit.
@@
   \
-2. **Periseason differences**
+2.**Periseason differences**\
+ This method calculate the hospitalizations attributable to influenza based on the differences in the hospitalization rates when respiratory virus was circulating and the hospitalization rates when there was no respiratory virus circulation in the community.
+   \
+  \
+
+## Brief introduction to 5 statistical methods for estimating respiratory virus–attributable hospitalization/mortality
+
+3.**Poisson regression with log link**
+\
+  Poisson regression is used to model count variables.
+Therefore, it can be used to model the number of hospitalizations over time.
+The logarithm of the expected value of the count variable is modeled as a linear combination of the predictor variables.
+For example, in Thompson's paper^[Thompson WW, Shay DK, Weintraub E, et al.
+Mortality Associated With Influenza and Respiratory Syncytial Virus in the United States.
+JAMA.
+2003;289(2):179–186.
+doi:10.1001/jama.289.2.179], the poisson regression is given by: \small
+  $$Y = \alpha \exp^{(\beta_0 + \beta_1t + \beta_2t^2  +  \beta_3 \sin(2\pi/52) + \beta_4 \cos(2\pi/52) + \beta_5 Flu_A(H1N1) + \beta_6 Flu_A(H3N2) + \beta_7Flu_B + \beta_8RSV)}$$ 
+  In this example, $Y$ is the observed number of deaths/hospitalization in a particular week, $\alpha$ is the population size as an offset term,$\beta_0$ is the baseline hospitalization incidence, $\beta_1t + \beta_2t^2$ represents the long-term trend of hospitalizations, $\sin(2\pi/52) + \beta_4 \cos(2\pi/52)$ captures the annual seasonal variations of hospitalizations, the rest are the pathogens that can contribute to the increase of the observed hospitalizations.
 

### 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
updates

@@
 
-- Section 1: Background knowledge of estimating disease burden (hospitalizations/mortality etc) with statistical models.
+- Section 1: Background knowledge of estimating respiratory virus attributable disease burden (hospitalizations/mortality etc) with statistical models.
 \
@@
 
-## Section 1: Background knowledge of estimating disease burden with statistical models.
+**Note: In this tutorial, we will focus on using time-series data.
+Therefore, we will not introduce other methods to estimate disease burden such as the disease pyramid.** For other methods, please check out Ginny's class *quantitative method in infectious disease epidemiology* and Dan's class *Public Health Surveillance*.
+
+## Section 1: Background knowledge of estimating respiratory virus attributable disease burden with statistical models.
 
@@
 
-5. **Box-Jenkins transfer function**
+5. **Box-Jenkins transfer function (ARIMA model)**
 
-## Brief introduction to 5 statistical methods for estimating respiratory virus–attributable hospitalization/mortality
+## Brief introduction to serfling regression and periseason differences
 
@@
 
-## Brief introduction to 5 statistical methods for estimating respiratory virus–attributable hospitalization/mortality
+## Brief introduction to poisson regression with log link
 
@@
 2003;289(2):179–186.
-doi:10.1001/jama.289.2.179], the poisson regression is given by: \small
+doi:10.1001/jama.289.2.179], the poisson regression with log link is given by: \small
   $$Y = \alpha \exp^{(\beta_0 + \beta_1t + \beta_2t^2  +  \beta_3 \sin(2\pi/52) + \beta_4 \cos(2\pi/52) + \beta_5 Flu_A(H1N1) + \beta_6 Flu_A(H3N2) + \beta_7Flu_B + \beta_8RSV)}$$ 
   In this example, $Y$ is the observed number of deaths/hospitalization in a particular week, $\alpha$ is the population size as an offset term,$\beta_0$ is the baseline hospitalization incidence, $\beta_1t + \beta_2t^2$ represents the long-term trend of hospitalizations, $\sin(2\pi/52) + \beta_4 \cos(2\pi/52)$ captures the annual seasonal variations of hospitalizations, the rest are the pathogens that can contribute to the increase of the observed hospitalizations.
+
+## Brief introduction to negative binomial regression with identity link
+
+4.**Negative binomial regression with identity link**
+\
+Negative binomial regression is also used to model count variables.
+It is a generalization of the poisson regression, in which the variance is assumed to be equal to the mean.
+The new approach use an identity link instead of an log link because log link corresponds to a multiplicative relationship among covariates (including baseline hospitalizations, seasonality, and several respiratory viruses) while identity link assumes additive effects.
+Apparently, additive effects are more realistic than multiplicative effects.
+The negative binomial regression with identity link is given by:
+\small$$Y \sim NB(p,r), \space p = \frac{r}{r+\lambda}$$
+\small$$\lambda=\beta_0 + \beta_1t + \beta_2t^2  +  \beta_3 \sin(2\pi/52) + \beta_4 \cos(2\pi/52) + ... + \beta_8RSV$$
+
+where $Y$ is the number of observed failures (diseases), $p$ is the probability of success (survival/not infected), $r$ is the number of successes (survival/not infected) and $\lambda$ is the expected value of failures (diseases).
+All other parameters remain the same as what we introduced in Poisson regression.
+
+## Brief introduction to ARIMA model
+
+5.**ARIMA model**
+\
 

### 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
updates

@@
 
-5.**ARIMA model**
+5.**Autoregressive integrated moving average (ARIMA) model**
 \
+Autoregressive integrated moving average (ARIMA) model can account for the autorcorrelation of the time-series and also the time-delayed associations.
+For infectious diseases, the number of the infected individuals in the next time point is correlated with the number of the infected individuals currently.
+This phenomenon is described as autocorrelated.
+The influenza/RSV infection reported this week may contribute to hospitalizations/deaths next week, which means the time-delayed associations.
+For example, in Gilca's paper^[Rodica Gilca, Gaston De Serres, Danuta Skowronski, Guy Boivin, David L.
+Buckeridge, The Need for Validation of Statistical Methods for Estimating Respiratory Virus–Attributable Hospitalization, American Journal of Epidemiology, Volume 170, Issue 7, 1 October 2009, Pages 925–936, https://doi.org/10.1093/aje/kwp195], ARIMA model is given by:
+\small
+$$Y_t=\alpha_0 + \alpha_1Temp_t + \alpha_2Holiday_t + \beta_1Flu_t+ ... + \sum_{i=0}^2\omega_iRSV_{t-i}$$
+$Y$ is the observed number of deaths/hospitalization in a particular week $t$, $\alpha_0$ is the baseline hospitalization incidence,$\alpha_1Temp_t + \alpha_2Holiday_t$ are the hospitalizations attributable to seasonal changes and holiday gathering, $\beta_1Flu_t$ is the flu attributable hospitalizations, and $\sum_{i=0}^2\omega_iRSV_{t-i}$ is the RSV attributable hospitalizations.
+This example suggested that not only RSV infections this week but also RSV infections reported 1 week and 2 weeks ago associates with hospitalizations reported this week.
 
@@
 
-5. **ARIMA model**
+5. **Autoregressive integrated moving average (ARIMA) model**
   - [Community influenza outbreaks and emergency department ambulance diversion](https://www.sciencedirect.com/science/article/pii/S0196064403013167)
@@
 
-- []
-
 - [The Need for Validation of Statistical Methods for Estimating Respiratory Virus–Attributable Hospitalization](https://academic.oup.com/aje/article/170/7/925/91526)

### 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update

@@
 \
-Autoregressive integrated moving average (ARIMA) model can account for the autorcorrelation of the time-series and also the time-delayed associations.
+ARIMA model can account for the autorcorrelation of the time-series and also the time-delayed associations.
 For infectious diseases, the number of the infected individuals in the next time point is correlated with the number of the infected individuals currently.
 This phenomenon is described as autocorrelated.
-The influenza/RSV infection reported this week may contribute to hospitalizations/deaths next week, which means the time-delayed associations.
-For example, in Gilca's paper^[Rodica Gilca, Gaston De Serres, Danuta Skowronski, Guy Boivin, David L.
+The influenza infection reported this week may contribute to hospitalizations next week, which means the time-delayed associations.
+For example, in Gilca's paper^[\tiny Rodica Gilca, Gaston De Serres, Danuta Skowronski, Guy Boivin, David L.
 Buckeridge, The Need for Validation of Statistical Methods for Estimating Respiratory Virus–Attributable Hospitalization, American Journal of Epidemiology, Volume 170, Issue 7, 1 October 2009, Pages 925–936, https://doi.org/10.1093/aje/kwp195], ARIMA model is given by:
 \small
-$$Y_t=\alpha_0 + \alpha_1Temp_t + \alpha_2Holiday_t + \beta_1Flu_t+ ... + \sum_{i=0}^2\omega_iRSV_{t-i}$$
-$Y$ is the observed number of deaths/hospitalization in a particular week $t$, $\alpha_0$ is the baseline hospitalization incidence,$\alpha_1Temp_t + \alpha_2Holiday_t$ are the hospitalizations attributable to seasonal changes and holiday gathering, $\beta_1Flu_t$ is the flu attributable hospitalizations, and $\sum_{i=0}^2\omega_iRSV_{t-i}$ is the RSV attributable hospitalizations.
-This example suggested that not only RSV infections this week but also RSV infections reported 1 week and 2 weeks ago associates with hospitalizations reported this week.
+$$Y_t=\alpha_0 + \alpha_1Temp_t + \alpha_2Holiday_t + \beta_1Flu_t + \sum_{i=0}^2\omega_iRSV_{t-i}$$
+$Y_t$ is the observed number of hospitalizations in a particular week $t$, $\alpha_0$ is the baseline hospitalization incidence,$\alpha_1Temp_t + \alpha_2Holiday_t$ are the hospitalizations attributable to seasonal changes and holiday gathering, $\beta_1Flu_t$ is the flu attributable hospitalizations, and $\sum_{i=0}^2\omega_iRSV_{t-i}$ is the RSV attributable hospitalizations.
+**This example suggested that not only RSV infections this week but also RSV infections reported 1 week and 2 weeks ago associates with hospitalizations reported this week.**
 
@@
 - [The Need for Validation of Statistical Methods for Estimating Respiratory Virus–Attributable Hospitalization](https://academic.oup.com/aje/article/170/7/925/91526)
+
+## Section 2: Background knowledge of Hierarchical Bayesian regression

### 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update

@@
 \
-- Section 2: Background knowledge of Hierarchical Bayesian regression.
+- Section 2: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations.
 \
 \
-- Section 3: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations.
-\
-\
-- Section 4: R code for running the Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups.
+- Section 3: R code for running the Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups.
 
@@
 
-## Section 2: Background knowledge of Hierarchical Bayesian regression
+## Section 2: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations
+
+### Background knowledge of Hierarchical Bayesian regression
+
+Hierarchical Bayesian regression can make population inference based on group and individual observations.
+The benefits of using a Hierarchical Bayesian structure is that it can shrink the uncertainty of inference while allow between-group variability.
+
+## Background knowledge of Hierarchical Bayesian regression: example
+
+For example, if we would like to estimate the baseline hospitalizations $\beta_{0jk}$ in age $j$ and socioeconomic group $k$, we can treat this parameter as (1) independent, meaning that the baseline hospitalizations are different in each group and we will need to estimate 27 parameters $\beta_{0j=1,k=1}$, $\beta_{0j=1,k=2}$...$\beta_{0j=9,k=3}$, (2) the same, meaning that the baseline hospitalizations are the same in all age and socioeconomic group and we will only need to estimate one parameter $\beta_0$, and (3) random draws from the a normal distribution, meaning that baseline hospitalizations are normally distributed in the whole population and baseline hospitalization of each subpopulation $jk$ is an realization of the distribution (a point in the distribution).
+In this case, we will estimate the mean and variations of the normal distribution and then get the subpopulation estimates.
+
+## Further readings on Hierarchical Bayesian regression
+
+- [Bayesian hierarchical modeling by wikipedia](https://en.wikipedia.org/wiki/Bayesian_hierarchical_modeling)
+
+- [Chapter 10 Bayesian Hierarchical Modeling by Drs. Jim Albert and Jingchen Hu](https://bayesball.github.io/BOOK/bayesian-hierarchical-modeling.html)
+
+- [Hierarchical Bayesian Modeling by Dr. Angie Wolfgang](https://astrostatistics.psu.edu/RLectures/hierarchical.pdf)

### 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update

@@
 
-For example, if we would like to estimate the baseline hospitalizations $\beta_{0jk}$ in age $j$ and socioeconomic group $k$, we can treat this parameter as (1) independent, meaning that the baseline hospitalizations are different in each group and we will need to estimate 27 parameters $\beta_{0j=1,k=1}$, $\beta_{0j=1,k=2}$...$\beta_{0j=9,k=3}$, (2) the same, meaning that the baseline hospitalizations are the same in all age and socioeconomic group and we will only need to estimate one parameter $\beta_0$, and (3) random draws from the a normal distribution, meaning that baseline hospitalizations are normally distributed in the whole population and baseline hospitalization of each subpopulation $jk$ is an realization of the distribution (a point in the distribution).
+For example, if we would like to estimate the baseline hospitalizations $\beta_{0jk}$ in age $j$ and socioeconomic group $k$, we can treat this parameter as (1) independent, meaning that the baseline hospitalizations are different in each group and we will need to estimate 27 parameters $\beta_{0j=1,k=1}$, $\beta_{0j=1,k=2}$...$\beta_{0j=9,k=3}$, (2) the same, meaning that the baseline hospitalizations are the same in all age and socioeconomic group and we will  estimate one parameter $\beta_0$, and (3) random draws from the a normal distribution, meaning that baseline hospitalizations are normally distributed in the whole population and baseline hospitalization of each subpopulation $jk$ is an realization of the distribution (a point in the distribution).
 In this case, we will estimate the mean and variations of the normal distribution and then get the subpopulation estimates.
+\
+\
+If we model the parameters in Bayesian framework, the third scenario corresponds to a Hierarchical Bayesian model.
+
+## Background knowledge of Hierarchical Bayesian regression: Benefits
+
+From the aforementioned example, e
 

### 2023-01-30 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update

--- 2023-01-27 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
@@
   fonttheme: "structurebold"
+header-includes:
+  - \usepackage{pdfpages}
+  - \setbeamercolor{background canvas}{bg=}
+  - \makeatletter\beamer@ignorenonframefalse\makeatother
 ---
@@
 \
-- Section 2: Background knowledge of Hierarchical Bayesian regression.
+- Section 2: Background knowledge of Hierarchical Bayesian regression
 \
 \
-- Section 3: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations.
-\
-\
-- Section 4: R code for running the Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups.
+- Section 3: Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups and its R code.
 
@@
 
-## Section 2: Background knowledge of Hierarchical Bayesian regression
+## Section 2: Background knowledge of using Hierarchical Bayesian regression to estimate RSV attributable hospitalizations
+The benefits of using a Hierarchical Bayesian structure is that it can shrink the uncertainty of inference while allow between-group variability. 
+\
+\
+For example, if we would like to estimate the baseline hospitalizations $\beta_{0jk}$ in age $j$ and socioeconomic group $k$, we can treat this parameter as (1) independent, meaning that the baseline hospitalizations are different in each group and we will need to estimate 27 parameters $\beta_{0j=1,k=1}$, $\beta_{0j=1,k=2}$...$\beta_{0j=9,k=3}$, (2) the same, meaning that the baseline hospitalizations are the same in all age and socioeconomic group and we will  estimate one parameter $\beta_0$, and (3) random draws from the a normal distribution, meaning that baseline hospitalizations are normally distributed in the whole population and baseline hospitalization of each subpopulation $jk$ is an realization of the distribution (a point in the distribution).
+In this case, we will estimate the mean and variations of the normal distribution and then get the subpopulation estimates.
+
+## Background knowledge of Hierarchical Bayesian regression: Benefits
+
+If we model the parameters in Bayesian framework, the third scenario corresponds to a Hierarchical Bayesian model.
+\
+\
+Compared with treating the parameters as group-specific, Bayesian Hierarchical structure pools information across groups to help shrink the uncertainty of the single group estimates. 
+\
+\
+Compared with treating the group specific parameters as homogeneous, Bayesian Hierarchical structure gives the flexibility and possibility of estimating group specific parameters.
+It can help evaluate the hypothesis of between-group variation.
+
+## Visualization of a Hierarchical Bayesian structure:
+
+![Hierarchical Bayesian structure](./Hier_Bay_figure.png)
+
+```{r,eval=FALSE,echo=FALSE}
+curve(dnorm(x, 0, 100), from=-200, to=200,main="prior distribution of mean of beta0",xlab="mu",ylab="dnorm(0,100)")
+curve(dinvgamma(x, shape=0.01, rate = 0.01, log = FALSE), from=0, to=200,ylab="dinvgamma(0.01,0,01)",main="prior distribution of\nvariance sigma square",xlab="sigma square")
+```
+
+## Further readings on Hierarchical Bayesian regression
+
+- [Bayesian hierarchical modeling by wikipedia](https://en.wikipedia.org/wiki/Bayesian_hierarchical_modeling)
+
+- [Chapter 10 Bayesian Hierarchical Modeling by Drs. Jim Albert and Jingchen Hu](https://bayesball.github.io/BOOK/bayesian-hierarchical-modeling.html)
+
+- [Hierarchical Bayesian Modeling by Dr. Angie Wolfgang](https://astrostatistics.psu.edu/RLectures/hierarchical.pdf)
+
+## Section 3: Hierarchical Bayesian regression to estimate RSV attributable hospitalizations in older adults by age and risk groups and its R code.
+
+\includepdf[pages=2]{Hier_Bay_figure.pdf}
+
+## Hierarchical Bayesian regression structure
+
+\includepdf[pages=3]{Hier_Bay_figure.pdf}
+
+``` {=latex}
+\end{frame}
+\includepdf[pages=1-7]{model_structure.pdf}
+\begin{frame}
+```
+
+## R code (JAGS code) for Hierarchical Bayesian regression structure

### 2023-01-31 4_RSV_disease_burden/4_RSV_disease_burden.Rmd
update tutorials

@@
 ## R code (JAGS code) for Hierarchical Bayesian regression structure
+
+Please open Jags_model.R for the whole R script.
+\
+\
+$k=1,2,3$ represent each socioeconomic groups, $j=1,...,9$ represent each age groups, and $i=1,...,108$ represent each month.
+\
+\
+Hierarchical Bayesian model are specified in jags as follows:
+```{r, eval=FALSE}
+model {
+for(k in 1:n.group){  
+  for (j in 1:n.age){
+      for (i in 1:n.date) {
+
+ y[i,j,k] ~ dnegbin(prob[i,j,k],r) 
+ # negative binomial
+  prob[i,j,k]<- r/(r+lambda[i,j,k])  ## likelihood 
+```
+
+## R code (JAGS code): expected respiratory hospitalizations
+
+```{r, eval=FALSE}
+# expected respiratory hospitalizations lambda
+    lambda[i,j,k] <- rd0[j,k]+ # baseline hosp
+exp(epi[epi.year[i]]) + # year to year variation
+# epi.year[i] is an input indicator for the calendar year
+# epi.year[i] = 1,2,...,9
+exp(delta[month[i]]) + # monthly variations
+# month[i] is an input indicator for the epi month
+# month[i] = 1,2,...,12 (1 for July and 12 for June)
+rsv[i,j,k]*rd2[j,k] + # RSV attributable hosp
+flu[i,j,k]*rd1[epi.year[i],j,k] # flu attributable hosp
+      }
+
+# baseline hospitalizations 
+#(must be greater than or equal to 0)
+  rd0[j,k] <- exp(beta0[j,k])
+  beta0[j,k]~ dnorm(mu0,tau0) 
+# hierarchical structure: shared mean and variance
+```
+
+## The coefficient of flu in each age and SES group
+
+```{r, eval=FALSE}
+ for (p in 1:n.year) { 
+# epi-year effects for antigen shift
+# coefficient of influenza-associated respiratory 
+# hospitalization varies annually 
+    rd1[p,j,k] <- exp(beta1[p,j,k]) 
+    # ensure positive coefs
+    beta1[p,j,k] ~ dnorm(beta1_mean[p,j,k],tau.flu)
+    # shared variance
+    # the mean of the coefficient is 
+# a multiplicative combination of 
+    beta1_mean[p,j,k] <- gamma_flu[k]+ omega_flu[j]+ xi_flu[p]
+    # SES effects gamma_flu[k]
+    # age effects omega_flu[j]
+    # epi-year (antigen) xi_flu[p]
+    }
+```
+
+## The coefficient of RSV in each age and SES group
+
+```{r, eval=FALSE}
+ # this coefficient depends on SES and age 
+    rd2[j,k] <- exp(beta2[j,k]) # ensure positive
+    beta2[j,k] ~ dnorm(beta2_mean[j,k],tau.rsv)
+    beta2_mean[j,k]<- gamma[k]+omega[j]
+   # SES effects gamma[k]
+  # age effects omega[j]
+```
+
+## hyperparameters
+
+```{r, eval=FALSE}
+# n.year = 1,2,...,9
+for (p in 1:n.year) { 
+epi[p] ~ dnorm(0, tau.epi) 
+# prior for yearly variation of baseline hosp
+xi_flu[p] ~ dnorm(0,tau7)}
+# prior for flu annual variation (antigen shift)
+
+  for (m in 1:12){
+    delta[m] ~ dnorm(0,disp.m)
+  }
+# prior for monthly variation of baseline hosp
+```
+
+## hyperparameters (continued)
+
+```{r, eval=FALSE}
+for(k in 1:n.group){
+gamma_flu[k] ~ dnorm(0,tau3)
+gamma[k] ~ dnorm(0, tau4)
+}
+# prior for SES effects of flu and RSV
+
+for(j in 1:n.age){
+omega[j] ~ dnorm(0, tau5)
+omega_flu[j] ~ dnorm(0,tau6)
+}
+# prior for age effects of flu and RSV
+```
+
+## Priors
+```{r, eval=FALSE}
+ r ~ dunif(0,250) 
+#r > 0 serves as an overdispersion parameter
+  mu0 ~ dnorm(0,0.0001) # 0.0001 = 1/variance
+# In jags, dnorm specify precision
+
+# the conjugate priors for variance of normal distribution
+# is an inverse Gamma distribution
+# In jags, we put priors on precision
+  tau0 ~ dgamma(0.01, 0.01)
+  tau1 ~ dgamma(0.01, 0.01)
+  tau2 ~ dgamma(0.01, 0.01)
+  tau3 ~ dgamma(0.01, 0.01)
+  tau4 ~ dgamma(0.01, 0.01)
+  tau5 ~ dgamma(0.01, 0.01)
+  tau6 ~ dgamma(0.01, 0.01)
+  tau7 ~ dgamma(0.01, 0.01)
+  tau8 ~ dgamma(0.01, 0.01)
+  tau.epi ~ dgamma(0.01, 0.01)
+  tau.flu ~ dgamma(0.01, 0.01)
+  tau.rsv ~ dgamma(0.01, 0.01)
+  disp.m ~ dgamma(0.01, 0.01)
+```
+
+## Clarification for input data
+
+**Note: we coded as rsv[i,j,k] and flu[i,j,k] for the purpose of a clear dimension for multiplication.**  This is a 1 to 1 multiplication, not a matrix multiplication.
+rsv[i,j,k] are the same for j=1,...,9, representing ICD-9 recorded RSV hospitalizations in children under 2 in each SES group in each month.
+flu[i,j,k] are the same for j=1,...,9, representing the total ICD-9 recorded flu hospitalizations in the entire age spectrum in each SES group in each month.
+
+\
+\
+
+In the equation, we simplify them to $RSV_{ik}$ and $Flu_{ik}$.
+
+## Specify input data
+
+```{r, eval=FALSE}
+epi.year <- as.factor(rep(1:9, each=12))
+# indicator for calendar year
+month <- rep(1:12,9)
+# indicator for epi month (1 for July and 12 for June)
+
+dataset <- list('y' = y_income_whole, 
+# respiratory hospitalizations in each age and  
+# SES group in each month
+"rsv"=rsv_income_whole,
+# ICD-9 recorded RSV hospitalizations   
+# in children under 2 in each SES group in each month
+"flu"=flu_income_whole, 
+# the total ICD-9 recorded flu hospitalizations in 
+# the entire age spectrum in each SES group in each month 
+'epi.year'=epi.year,"month"=month,
+n.age=9,n.date=108,n.year=9,"n.group"=3) 
+# number of age/SES groups; number of months and years
+```
+
+## Posterior sampling
+
+```{r, eval=FALSE}
+jags_post <- jags.model(
+  textConnection(model_string), 
+  data = dataset,n.chains = 3)
+# specify the number of chains
+
+update(jags_post, n.iter=82500) # burn-in period
+
+rsv_resp <- coda.samples(jags_post,
+                         variable.names=
+  c("rd2","lambda","rd1", "rd0","epi","delta"),
+                      thin = 10,n.iter = 12500)
+# posterior samples
+```
+
+## Estimate the RSV-attributable hospitalizations in each age and SES group
+
+```{r, eval=FALSE}
+## Remember to check convergence first
+## use: gelman.diag() see 3_JAGS_Bayesian_model
+## Calculate RSV attributable respiratory 
+## hospitalization incidence and percent
+  post1 <- as.data.frame(as.matrix(rsv_resp[[1]]))
+  post2 <- as.data.frame(as.matrix(rsv_resp[[2]]))
+  post3 <- as.data.frame(as.matrix(rsv_resp[[3]]))
+  post <- bind_rows(post1,post2,post3)
+## A total of 3750 posterior samples
+  lambda <- post[, grep("lambda", colnames(post), fixed=T)]
+  rd2.resp <- post[, grep("rd2[", colnames(post), fixed=T)]
+
+rsv_count <- array(data = NA,dim = c(3750,108,9,3))
+for (i in 1:108) {
+  for (j in 1:9) {
+    for (k in 1:3) {
+      rsv_count[,i,j,k] <- rsv_income[i,j,k]*rd2.resp[,j+9*(k-1)]}}}
+## posterior samples of RSV-attributable hospitalizations
+## in each SES and age group in each month
+```

### 2023-01-31 4_RSV_disease_burden/Jags_model.R

@@ -1,0 +1,71 @@
+model_string<-"
+model {
+for(k in 1:n.group){  
+  for (j in 1:n.age){
+      for (i in 1:n.date) { 
+    
+   lambda[i,j,k] <-   rd0[j,k]+
+               exp(epi[epi.year[i]]) +
+               exp(delta[month[i]]) +
+               rsv[i,j,k]*rd2[j,k] + 
+               flu[i,j,k]*rd1[epi.year[i],j,k]
+  y[i,j,k] ~ dnegbin(prob[i,j,k],r)
+  prob[i,j,k]<- r/(r+lambda[i,j,k])  ## likelihood 
+}
+  # baseline hospitalizations (must be greater than or equal to 0)
+  rd0[j,k] <- exp(beta0[j,k])
+  beta0[j,k]~ dnorm(mu0,tau0)
+  
+  for (p in 1:n.year) { 
+    # epi-year effects
+    # random slope
+    # coefficient of influenza-associated respiratory hospitalization
+    # this coefficient varies annually
+    rd1[p,j,k] <- exp(beta1[p,j,k]) # ensure positive coefs
+    beta1[p,j,k] ~ dnorm(beta1_mean[p,j,k],tau.flu)
+    beta1_mean[p,j,k] <- gamma_flu[k]+ omega_flu[j]+ xi_flu[p]
+    }
+  
+    # coefficient of RSV-associated respiratory hospitalization
+    # this coefficient depends on SES and age 
+    rd2[j,k] <- exp(beta2[j,k]) # ensure positive
+    beta2[j,k] ~ dnorm(beta2_mean[j,k],tau.rsv)
+    beta2_mean[j,k]<- gamma[k]+omega[j]
+    
+  
+  # Month dummy variable, account for other pathogens and general seasonality 
+}}
+## hyperpriors
+for (p in 1:n.year) { 
+epi[p] ~ dnorm(0, tau.epi) }
+  for (m in 1:12){
+    delta[m] ~ dnorm(0,disp.m)
+}
+for(k in 1:n.group){
+gamma_flu[k] ~ dnorm(0,tau3)
+gamma[k] ~ dnorm(0, tau4)
+}
+for(j in 1:n.age){
[…]
